### 2025-02-21 paper/markdown/manuscript_figures_tables.Rmd
Figure legends
Legends for the main figures

---
title: "Prognostic and biological relevance of collagen-related genes in prostate cancer"
subtitle: "Methods, figures and tables for the transcriptome part"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: coll_biblio.bib

csl: frontiers-in-immunology.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

library(bookdown)
library(flextable)

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "png", 
                      dpi = 600, 
                      tab.cap.pre = 'Table ', 
                      tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

# Figures

```{r fig-uni-mrna, fig.width = figur::convert(fig$uni_mrna, to = 'in')$w, fig.height = figur::convert(fig$uni_mrna, to = 'in')$h, fig.cap = 'Univariable analysis of BCR-free survival with collagen-related transcripts as candidate standalone prognostic factors.'}

fig$uni_mrna$plot

```

__Figure \@ref(fig:fig-uni-mrna). Univariable analysis of BCR-free survival with collagen-related transcripts as candidate standalone prognostic factors.__

_Prostate cancer (PCA) patients in the pooled GEO (total: n = `r uni_globals$n_numbers$n_total["geo_pool"]`, biochemical relapse [BCR]: n = `r uni_globals$n_numbers$n_events["geo_pool"]`), TCGA (total: n = `r uni_globals$n_numbers$n_total["tcga"]`, BCR: n = `r uni_globals$n_numbers$n_events["tcga"]`), and DKFZ cohort (total: n = `r uni_globals$n_numbers$n_total["dkfz"]`, BCR: n = `r uni_globals$n_numbers$n_events["dkfz"]`) were classified as high and low expressors for each of the `r `length(globals$genes)` collagen-related transcripts with the expression cutoffs corresponding to the largest differences in survival between the high and low expressors._
_BCR risk was compared between the high and low expressors for each gene by univariable Cox proportional hazard regression. P values were corrected for multiple testing with the false discovery rate (FDR) method._
_Standalone BCR risk markers were defined as transcripts associated with unfavorable prognosis (high versus low expressors, hazard ratio [HR] > 1, pFDR < 0.05) or favorable prognosis (HR < 1, pFDR < 0.05) in all three cohorts._

_HR with 95% confidence intervals for BCR risk in the high versus low expression strata are presented in a Forest plot (left). The model performance statistics, Harrell's concordance index (C-index, measure of model accuracy), and Integrated Brier Scores (IBS, measure of overall model calibration, low values indicative of good calibration) are presented in a dot plot (right)._

\newpage

```{r fig-ml-mrna, fig.width = figur::convert(fig$ml_mrna, to = 'in')$w, fig.height = figur::convert(fig$ml_mrna, to = 'in')$h, fig.cap = 'Machine learning modeling of BCR-free survival with expression levels of the collagen-related transcripts as explanatory factors.'}

fig$ml_mrna$plot

```

__Figure \@ref(fig:fig-ml-mrna). Machine learning modeling of BCR-free survival with expression levels of the collagen-related transcripts as explanatory factors.__

_Risk of biochemical relapse (BCR) in prostate cancer (PCA) was modeled by multi-parameter machine learning models with ComBat-processed $log_2$-transformed expression levels of `r length(globals$genes)` collagen-related transcripts as explanatory factors._ 
_The models were trained in the pooled GEO cohort with six machine learning algorithms (RIDGE Cox regression, Elastic Net Cox regression, LASSO Cox regression, survival Support Vector Machines [SVM], survival Random Forest, and survival Gradient Boosted Machines [GBM])._
_Performance of the models was evaluated in the pooled GEO training cohort (total: n = `r uni_globals$n_numbers$n_total["geo_pool"]`, BCR: n = `r uni_globals$n_numbers$n_events["geo_pool"]`), and the TCGA (total: n = `r uni_globals$n_numbers$n_total["tcga"]`, BCR: n = `r uni_globals$n_numbers$n_events["tcga"]`) and DKFZ (total: n = `r uni_globals$n_numbers$n_total["dkfz"]`, BCR: n = `r uni_globals$n_numbers$n_events["dkfz"]`) validation collectives._

_(A) Metrics of performance of the survival models, Harrell's concordance index (C-index, measure of model accuracy) and Integrated Brier Score (IBS, measure of overall model calibration, low values are characteristic for good calibration), are presented in dot plots. Dashed lines represent C-index and IBS values expected for random BCR risk prediction. Note the superior accuracy and calibration of the GBM model in the validation collectives._

_(B) BCR-free survival in PCA patients stratified by tertiles of the predictor scores of the best performing GBM model. Statistical significance of differences between the predictor score tertiles was determined by Peto-Peto test corrected for multiple testing wit the false discovery rate method. Fractions of BCR-free patients and minimum/maximum scaled time after diagnosis are visualized in Kaplan-Meier plots. Numbers of patients and BCR in the predictor score tertiles are indicated in the legends. P values are displayed in the plots._

_(C) Importance of the explanatory variables for BCR-risk prediction by the GBM model was gauged by difference in sum of squared error ($\Delta SSE$) attributed to inclusion of a particular collagen-related gene in the GBM learner ensemble. High $\Delta SSE$ values are characteristic for highly influential variables. $\Delta SSE$ values for the variables that contributed substantially to the BCR risk prediction ($\Delta SSE$ > 0) are presented in a bar plot._

\newpage

# References

### 2025-02-21 paper/markdown/manuscript_figures_tables_legacy.Rmd
Figure legends
Legends for the main figures

---
title: "Prognostic and biological relevance of collagen-related genes in prostate cancer"
subtitle: "Methods, figures and tables for the transcriptome part"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: coll_biblio.bib

csl: frontiers-in-immunology.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

library(bookdown)
library(flextable)

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "png", 
                      dpi = 600, 
                      tab.cap.pre = 'Table ', 
                      tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

# Methods

## Magnetic resonance imaging of the prostate

_Dear co-authors, it's your part as experts:)_

ADC was compared between patient-matched measurements of benign prostate and malignancy by paired T test with Cohen's d effect size statistic.

## Transcriptome analyses

The transcriptome analysis was performed with R version 4.2.3. 
Its details are presented in __Supplementary Methods__.

## Transcriptome data sets

Six publicly available transcriptome data sets of primary prostate cancer samples, 
the TCGA prostate cancer cohort [@Liu2018; @Abeshouse2015] 
(n = `r tcga$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), 
GSE54460 [@Long2014] (n = `r gse54460$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), 
GSE70768  [@Ross-Adams2015] (n = `r gse70768$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray), 
GSE70769  [@Ross-Adams2015] (n = `r gse70769$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray), 
GSE220095 [@Schimmelpfennig2023] (n = `r gse220095$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), 
DKFZ [@Gerhauser2018] (n = `r dkfz$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing) 
were re-analyzed (__Supplementary Table S1__). 
Normalized gene expression data and clinical information were fetched from the cBioportal repository (DKFZ, TCGA) or Gene Expression Omnibus (GEO). 
Selection criteria of the data sets were: availability of Gleason scoring, information of biochemical relapse and relapse-free survival, and availability of expression data for `r nrow(globals$genes_interest)` collagen-related genes covered by proteomic analysis as well (__Supplementary Table S2__).
Donor-matched benign and malignant tissue samples were available for the 
GSE70768 (n = `r norm_tumor$n_numbers$gse70768` tissue pairs) and TCGA cohort (n = `r norm_tumor$n_numbers$tcga` tissue pairs). 
Expression values were $log_2$-transformed prior to analyses. 
Non-immune cell content in cancer samples was estimated with the _MCP Counter_ and _xCell_ algorithms [@Aran2017; @Becht2016].
single sample gene set enrichment scores (ssGSEA) for the Reactome pathway gene signatures retrieved from the MSig database version 7.5.1 were computed with the _GSVA_ algorithm [@Hanzelmann2013].

The collagen-related genes were retrieved from the published Collagen Signature [@Kocher2021] and the Reactome collagen formation pathway (R-HSA-1474290) and restricted to genes covered by the proteomic analyses (__Supplementary Table S2__).

## Differential expression in benign and malignant prostate tissue and cancers stratified by Gleason score

Expression of the collagen-related genes of interest between patient-matched benign and malignant specimens was compared by the false discovery rate method (FDR) corrected paired two-tailed T test with Cohen's d effect size statistic. 
Differentially expressed genes were defined by pFDR < 0.05 and d $\geq$ 0.2 criteria indicative of significant, at least weak expression regulation (__Supplementary Table S3__).

Differential gene expression in cancer samples stratified by the ISUP risk strata (ISUP1: Gleason score 5 - 6, ISUP2: Gleason score 7 and ISUP3: Gleason score $\geq$ 8) was assessed by FDR-corrected one-way ANOVA with $\eta^2$ effect size statistic. 
Genes with pFDR < 0.05 and $\eta^2 \geq$ 0.02 were deemed differentially regulated (__Supplementary Table S4__).

## Prediction of biochemical relapse-free survival by gradient boosted machine modeling

For survival analyses, $log_2$-transformed expression values of the collagen-related genes (__Supplementary Table S2__) were subjected to batch-adjustment with the ComBat algorithm [@Leek2012]. 
Subsequently, the ComBat-adjusted expression data sets of the GSE54460, GSE70768, GSE70769, GSE220095 cohorts were merged (further referred to as 'pooled GEO' cohort).

Biochemical relapse-free survival in the pooled GEO training cohort was modeled with normalized (Z-scores) expression values of the collagen-related genes as explanatory variables by the Gradient Boosted Machine (GBM) algorithm [@Friedman2001; @Greenwell2022; @Natekin2013; @Friedman2002]. 
Selection of the optimal set of modeling parameters was accomplished by tuning with 10-fold cross-validation and out-of-fold model deviance as performance statistic (__Supplementary Tables S15__). 
The Transcriptomic Collagen Score was computed as the linear predictor score of the GBM model. 
Importance of the explanatory variables for the predictive performance of the GBM model were measured with the relative influence method [@Friedman2001] and expressed as difference in the sum of squared errors ($\Delta SSE$, __Figure \@ref(fig:fig-coll-score)A__).

Performance of the Transcriptomic Collagen Score at prediction of biochemical relapse-free survival in the training pooled GEO collective and validation data sets (TCGA and DKFZ) was assessed by univariable Cox regression with Harrell's concordance index [@Harrell1996] and integrated Brier score [@Graf1999] as metrics of goodness of fit and calibration (__Figure \@ref(fig:fig-coll-score)B__, __Supplementary Table 16__). 
Of note, predictive performance of the GBM algorithm was excellent and reproducible in the investigated cohorts as compared with several other machine learning algorithms (e.g. Ridge Cox regression, Elastic Net or Support Vector Machines, __Supplementary Figure S9__). 
Differences in biochemical relapse-free survival in cancer patients stratified by tertiles of the Transcriptomic Collagen Scores were assessed by false discovery rate (FDR) corrected Peto-Peto test (__Figure \@ref(fig:fig-coll-score)C__).

## Data and code availability

Publicly available data sets were analyzed. 
Formatted data sets used for analyses will be made available upon request to the corresponding author. 
The transcriptome R analysis pipeline is available from GitHub (https://github.com/PiotrTymoszuk/collagen_pca).

\newpage

# Figures

```{r fig-norm-tumor, fig.width = figur::convert(paper_fig$norm_tumor, to = 'in')$w, fig.height = figur::convert(paper_fig$norm_tumor, to = 'in')$h, fig.cap = 'Differences in tissue diffusion capacity and expression of collagen-related genes between the prostate cancer and benign tissue. Expression of collagen-related genes in cancers stratified by Gleason scores.'}

paper_fig$norm_tumor$plot

```

__Figure \@ref(fig:fig-norm-tumor). Differences in tissue diffusion capacity and expression of collagen-related genes between the prostate cancer and benign tissue. Expression of collagen-related genes in cancers stratified by Gleason scores.__

_(A) Representative MRI axial sequences of the prostate of an 83-year-old patient with prostate cancer (left panel). The T2-weighted images (a) show a 23mm suspicious hypo-intense lesion in the left peripheral zone (red star) of the prostate, which has an early enhancement in the contrast-enhanced T1-weighted image (b). The DWI scan (c) and ADC map (d) show a strong diffusion restriction (black and white character) of the tumor area suggesting a PI-RADS 5 lesion. Statistical significance for differences in diffusion capacity in patient-matched tumor and benign tissue measured by ADC was determined by paired T test with Cohen's d effect size statistic (right panel). Single ADC values are visualized as points, grey lines connect measurements of the same donors. The number of analyzed benign - tumor pairs, effect size and p value are displayed in the plot caption._

_(B) Differences in $log_2$-transformed expression of `r nrow(globals$genes_interest)` collagen-related genes between donor-matched tumor and benign prostate tissue samples were investigated by paired T test with Cohen's d effect size statistic in the GSE70768 (n = `r norm_tumor$n_numbers$gse70768`) and TCGA (n = `r norm_tumor$n_numbers$tcga` tissue pairs) cohorts. P values were corrected for multiple testing with the false discovery rate method. Mean normalized $log_2$-transformed expression levels in benign and cancer tissue for `r length(unique(unlist(norm_tumor$common)))` genes significantly regulated in both cohorts with at least weak effect size (pFDR < 0.05, d $\geq$ 0.2) are presented in a heat map._

_(C) Differences in $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` collagen-related genes between cancer samples stratified according to the ISUP risk system were assessed by one-way ANOVA with $\eta^2$ effect size statistic in the GSE54460 (`r format_strata_n(gs_uni$n_numbers$gse54460)`), GSE70768 (`r format_strata_n(gs_uni$n_numbers$gse70768)`), GSE70769 (`r format_strata_n(gs_uni$n_numbers$gse70769)`), GSE220095 (`r format_strata_n(gs_uni$n_numbers$gse220095)`), DKFZ (`r format_strata_n(gs_uni$n_numbers$dkfz)`), and TCGA cohort (`r format_strata_n(gs_uni$n_numbers$tcga)`). Mean normalized $log_2$-transformed expression levels of genes significantly regulated between the Gleason score strata with at least weak effect size (pFDR < 0.05, $\eta^2 \geq$ 0.02) are presented in a heat map._

\newpage

```{r fig-tme, fig.width = figur::convert(paper_fig$tme, to = 'in')$w, fig.height = figur::convert(paper_fig$tme, to = 'in')$h, fig.cap = 'Expression of collagen-related genes in components of the tumor microenvironment.'}

paper_fig$tme$plot

```

__Figure \@ref(fig:fig-tme). Expression of collagen-related genes in components of the tumor microenvironment.__

_(A) Cancer samples in six published cohorts were classified as collagen low and collagen high tumors in respect to `r nrow(globals$genes_interest)` collagen-related transcripts by semi-supervised PAM clustering (GSE54460: `r format_strata_n(ana_globals$n_numbers$gse54460)`, GSE70768: `r format_strata_n(ana_globals$n_numbers$gse70768)`, GSE70769: `r format_strata_n(ana_globals$n_numbers$gse70769)`, GSE220095: `r format_strata_n(ana_globals$n_numbers$gse220095)`, DKFZ: `r format_strata_n(ana_globals$n_numbers$dkfz)`, TCGA: `r format_strata_n(ana_globals$n_numbers$tcga)`). Counts of cancer-associated fibroblasts and endothelial cells were predicted by the MCP Counter algorithm and compared between the collagen clusters by Mann-Whitney test with r effect size statistic. Median normalized cell counts with interquartile ranges are presented as boxes, whiskers span over 150% of the interquartile ranges. Values for single cancer samples are depicted as points. All presented effects were significant (pFDR < 0.05) with at least weak (endothelial cells, r > 0.21) or moderate effect size (fibroblasts, r > 0.55)._

_(B) Placeholder for the single cell RNA seq data._

\newpage

```{r fig-coll-score, fig.width = figur::convert(paper_fig$coll_score, to = 'in')$w, fig.height = figur::convert(paper_fig$coll_score, to = 'in')$h, fig.cap = 'Transcriptional Collagen Model of biochemical relapse-free survival.'}

paper_fig$coll_score$plot

```

__Figure \@ref(fig:fig-coll-score). Transcriptional Collagen Model of biochemical relapse-free survival.__

_A multi-parameter Gradient Boosted Machine (GBM) model of biochemical relapse-free survival was trained in the pooled GEO cohort with normalized $log_2$-transformed and ComBat-adjusted expression levels of `r nrow(globals$genes_interest)` collagen-related genes._

_(A) Importance of explanatory variables measured by the gradient of sum of squared errors ($\Delta SSE$) during the training process attributed to particular collagen-related transcript levels._

_(B) Performance of the GBM model at prediction of biochemical relapse-free survival in the training pooled GEO cohort and test collectives assessed by concordance index (C-index, high values indicate high concordance between the predicted and observed survival) and integrated Brier score (IBS, low values indicate good model fit and proper calibration). Values of C-index and IBS expected for random survival prediction are depicted as dashed lines._

_(C) Fractions of biochemical relapse-free patients in low, intermediate and high tertiles of Transcriptional Collagen Score in the training pooled GEO cohort, and the TCGA and DKFZ validation collectives presented in Kaplan-Meier plots. Differences in survival between the tertiles were investigated by Peto-Peto test. Total numbers of patients and numbers of biochemical relapses are displayed in the plot captions. P values are shown in the plots._

\newpage

# References

### 2025-02-21 paper/markdown/manuscript_supplement_legacy.Rmd
Figure legends
Legends for the main figures

---
title: "Prognostic and biological relevance of collagen-related genes in prostate cancer"
subtitle: "Supplementary material, transcriptome part"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: coll_biblio.bib

csl: frontiers-in-immunology.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

library(bookdown)
library(flextable)

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "png", 
                      dpi = 600, 
                      tab.cap.pre = 'Table ', 
                      tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

# Supplementary Methods

## Software

The analysis was done with R version 4.2.3 (R Foundation).

Tabular data were handled with the packages _tidyverse_ [@Wickham2019], _rlang_ [@Henry2022] and [_trafo_](https://github.com/PiotrTymoszuk/trafo). 
Text data were handled with _stringi_ [@Gagolewski2021].

Import of the TCGA and DKFZ data sets from the cBioportal repository was accomplished with in-house-developed R scripts. 
Transcriptome datasets from the Gene Expression Omnibus were fetched with the _GEOquery_ package [@Sean2007]. 
Gene and probe annotation was accomplished with the _AnnotationDbi_ [@Pages2022] and _org.Hs.eg.db_ packages [@Carlson2022]. 
For prediction of non-malignant cell counts and fractions in cancer samples, the R implementation of the _MCP Counter_ and _xCell_ algorithms provided by the _immunedeconv_ package was used [@Sturm2020; @Aran2017; @Becht2016]. 
Single sample gene set enrichment analysis scores (ssGSEA) were computed with the _GSVA_ algorithm [@Hanzelmann2013] implemented by the package [_gseaTools_](https://github.com/PiotrTymoszuk/gseaTools).

Semi-supervised clustering was done with the package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) employing algorithms from the _philentropy_, _cluster_, _factoextra_ and _umap_ packages [@Kassambara2020; @Schubert2019; @Drost2018; @Konopka2022].

For statistical hypothesis testing, effect size calculation, gene set variation analysis (GSVA) and differential gene expression analysis, the packages _rstatix_ [@Kassambara2021], [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) and [_microViz_](https://github.com/PiotrTymoszuk/microViz) were employed. 
Biological process gene ontology (GO) enrichment analysis was performed with the _goana_ algorithm implemented by the _limma_ [@Young2010] and [_microViz_](https://github.com/PiotrTymoszuk/microViz) packages. 
Clustering of GO terms by their semantic similarity was accomplished with _GOSemSim_ [@Yu2010], [_microViz_](https://github.com/PiotrTymoszuk/microViz) and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools). 
Differential modulation of transcriptional regulons and signaling pathways was analyzed with the _collecTRI_ [@Muller-Dott2023] and _PROGENy_ [@Schubert2018] databases by linear modeling tools of the _decoupleR_ package [@Badia-I-Mompel2022]. 
Metabolic reaction activity modeling and metabolic subsystem enrichment analysis were performed with the packages _BiGGR_ [@Gavai2015; @King2016] and [_biggrExtra_](https://github.com/PiotrTymoszuk/biggrExtra).

Multi-parameter modeling of biochemical relapse-free survival was done with the packages 
_glmnet_ [@Friedman2010] (Ridge, Elastic Net and LASSO cox regression), 
_survivalsvm_ [@Fouodo2018] (Support Vector Machines [SVM]), 
_rfsrc_ [@Ishwaran2008; @Ishwaran2022a] (Random Forest), 
_gbm_ [@Greenwell2022] (Gradient Boosted Machines [GBM]), 
_survival_ [@Therneau2000], 
_surviminer_ [@Kassambara2016] and 
[_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) were used. 
Univariable analysis of relapse-free survival was accomplished with [_kmOptimizer_](https://github.com/PiotrTymoszuk/kmOptimizer').

For visualization of the results, the packages 
_ggplot2_ [@Wickham2019] (scatter, stack, bubble and bar plots, heat maps), 
[_ExDA_](https://github.com/PiotrTymoszuk/ExDA) (violin, stack and ribbon plots), _survminer_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) (Kaplan-Meier plots), and 
[_microViz_](https://github.com/PiotrTymoszuk/microViz) (heat maps) were used. 
Figures and tables were created with the packages _cowplot_ [@Wilke2019] and _flextable_ [@Gohel2022].

## Data sources and data import

Transcriptome cohorts for analysis in the current report were selected from studies deposited at cBioportal and Gene Expression Omnibus (GEO) with the following criteria: availability of information on Gleason scoring or ISUP risk strata (International Society of Urological Pathology), biochemical relapse and biochemical relapse-free survival, and expression data for `r nrow(globals$genes_interest)` collagen-related genes investigated also in the proteomic part of the project (__Supplementary Table S\@ref(tab:tab-cohorts)__ and __S\@ref(tab:tab-genes)__).

The TCGA prostate cancer data set [@Liu2018; @Abeshouse2015] and DKFZ data set [@Gerhauser2018] consisted of normalized RNA sequencing data for `r tcga$expression %>% filter(tissue_type == 'tumor') %>% nrow` and `r dkfz$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples, respectively, with accompanying clinical information and were obtained from the cBioportal repository with in-house-developed R scripts. 
The TCGA cohort included also 
`r norm_tumor$n_numbers$tcga` donor-matched samples of the benign prostate and prostate cancer tissue.  
The GSE54460 [@Long2014] (RNA sequencing, n = `r gse54460$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples), 
GSE70768 [@Ross-Adams2015] (Illumina HumanHT-12 V4.0 expression beadchip microarray, n = `r gse70768$expression %>% filter(tissue_type == 'tumor') %>% nrow`), 
GSE70769 [@Ross-Adams2015] (Illumina HumanHT-12 V4.0 expression beadchip microarray, n = `r gse70769$expression %>% filter(tissue_type == 'tumor') %>% nrow`), and 
GSE220095 [@Schimmelpfennig2023] (RNA sequencing, n = `r gse220095$expression %>% filter(tissue_type == 'tumor') %>% nrow`) data sets included normalized whole-transcriptome and basic clinical information and were fetched from GEO with the `getGEO()` function from the _GEOquery_ package. 
The GSE70768 data set [@Ross-Adams2015] included also `r norm_tumor$n_numbers$gse70768` donor-matched normal prostate and cancer tissue specimens. 
Demographic and clinical characteristic of the investigated transcriptome data sets is provided in  __Supplementary Table S\@ref(tab:tab-cohorts)__.

Gene expression levels were transformed with the $log_2(expression)$ and $log_2(count + 1)$ function for the microarray and RNA sequencing data sets prior to further analyses. 
The _MCP Counter_ and _xCell_ estimates of non-malignant cell content in cancer samples were computed with the `deconvolute()` function from the _immunedeconv_ package. 
Gene signatures of the [Reactome pathways](https://reactome.org/) were obtained from the [MSig database](https://www.gsea-msigdb.org/gsea/msigdb/index.jsp), version 7.5.1. 
ssGSEA scores were computed with the `calculate()` function from the [_gseaTools_](https://github.com/PiotrTymoszuk/gseaTools) package. 
ISUP strata were defined based on Gleason Score information as follows: ISUP1: Gleason score 5 - 6, ISUP2: Gleason score 7, ISUP3+: Gleason score $\geq$ 8.

The investigated collagen-related genes included the published Collagen Signature [@Kocher2021] and the Reactome collagen formation pathway genes (R-HSA-1474290) and were constrained to features covered by proteomic analyses of the current report (__Supplementary Table S\@ref(tab:tab-genes)__).

## Statistical hypothesis testing, effect size and statistical significance

Differences in numeric variables between two groups were investigated by paired and unpaired two-tailed T tests with Cohen's d effect size statistic or Mann-Whitney tests with r effect size statistic. 
Differences in numeric variables between three or more groups were assessed by one-way ANOVA with $\eta^2$ effect size statistic. 
Differences in distribution of categorical variables were assessed by $\chi^2$ test with Cramer's V effect size statistic. 
Odds ratio served as an effect size statistic in enrichment analyses. 
Effect sizes were interpreted as follows [@Cohen2013]:

* Cohen's d, weak: 0.2 - 0.5, moderate: 0.5 - 0.8, large: $\geq$ 0.8

* r, weak: 0.1 - 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5

* $\eta^2$: weak: 0.02 - 0.13, moderate: 0.13 - 0.26, large: $\geq$ 0.26

* Cramer's V, weak: 0.1 - 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5

* OR, weak: 1.44 - 2.48, moderate: 2.48 - 4.27, large: $\geq$ 4.27

P values were corrected for multiple testing with the false discovery rate method (FDR) [@Benjamini1995] within each analysis step and cohort. 
Effects were considered statistically significant for FDR-corrected p values < 0.05.

## Comparison of gene expression between the benign and cancer tissue, and cancer samples stratified by the ISUP risk scheme

Differences in $log_2$-transformed expression levels of the collagen pathway genes between donor-matched cancer and benign prostate specimens were investigated with paired two-tailed T test with Cohen's effect size statistic in the GSE70768 and TCGA cohorts (__Figure 1B__, __Supplementary Figure S\@ref(fig:fig-norm-tumor)__ and __Supplementary Table S\@ref(tab:tab-norm-tumor)__). 
Differences in gene expression between cancer samples stratified according to the ISUP risk scheme (ISUP1: Gleason Score 5 - 6, ISUP2: Gleason score 7, ISUP3+: Gleason Score $\geq$ 8) were explored by one-way ANOVA with $\eta^2$ effect size statistic (__Figure 1C__, __Supplementary Table S\@ref(tab:tab-gleason)__).
Differentially regulated genes were defined by pFDR < 0.05 and at least weak effect size of the difference (d $\geq$ 0.2 or $\eta^2 \geq$ 0.02). 
The analyses were done with the function `compare_variables()` from the [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) package.

## Collagen clusters of prostate cancer samples

Prostate cancer samples in the TCGA cohort were clustered by normalized $log_2$-transformed expression levels of the collagen-related genes of interest with the PAM (partition around medoids) algorithm with cosine distance between the samples (function `kcluster()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)). 
The choice of the clustering algorithm was motivated by the following criteria (__Supplementary Figure S\@ref(fig:fig-clust-dev)A__):

* cluster separation measured by mean silhouette statistic [@Rousseeuw1987]

* fraction of potentially misclassified observations (observations with negative silhouette widths) [@Rousseeuw1987]

* explanatory performance gauged by fraction of explained clustering variance (ratio of the total between-cluster sum of squares to the total sum of squares)

* preservation of the nearest neighborhood (mean fraction of the five nearest neighbors placed in the same cluster) [@Venna2001]

* cluster re-assignment accuracy in 5-fold cross-validation [@Lange2004; @Leng2014]. 
Computation of those numeric statistics of clustering quality in a cross-validation setting was done with the methods `summary()` and `cv()` provided by the [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package.
The cluster number choice was based on the bend of the peak of the average silhouette statistic [@Rousseeuw1987] (method `plot()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools), __Supplementary Figure S\@ref(fig:fig-clust-dev)B__). 
By this means, two clusters of cancer samples were identified: 'collagen high' and 'collagen low' cancers.

Assignment of cancers samples of the GSE54460, GSE70768, GSE70769, GSE220095, and DKFZ cohorts to the collagen clusters based on normalized $log_2$-transformed expression levels of the collagen-related genes was done with an inverse distance weighted k-nearest neighbor classifier (method `predict()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) [@Leng2014]. 
The quality of the cluster assignment was assessed by comparing the collagen cluster distribution and numeric statistics of clustering quality described above the training TCGA data set and the test cohorts (__Supplementary Figure S\@ref(fig:fig-clust-dev)CD__). 
In addition, separation of the clusters was assessed by a visual inspection of the UMAP layout plots (method `plot()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) and heat maps of the mean levels of the clustering factors in the collagen clusters (__Supplementary Figure S\@ref(fig:fig-clusters)__).

Differences in the $log_2$-transformed collagen pathway gene expression between the collagen clusters were investigated by two-tailed T test with Cohen's d effect size statistic (__Supplementary Table S\@ref(tab:tab-cluster)__).

## Clinical characteristic of the collagen clusters

Differences in numeric clinical variables (age and PSA) between the collagen clusters were assessed by Mann-Whitney test with r effect size statistic (`test_two_groups()`, [_microViz_](https://github.com/PiotrTymoszuk/microViz)).  
Differences in qualitative clinical variables between the collagen clusters were investigated by $\chi^2$ test with Cramer's V effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)). 
The demographic and clinical characteristic of the collagen clusters is presented in __Supplementary Figure S\@ref(fig:fig-clust-tumor)__ and __Supplementary Table S\@ref(tab:tab-clinic)__.

## Non-malignant cell infiltration in the collagen clusters

Non-malignant cell counts and non-malignant cell fractions in the cancer samples were predicted by the _MCP Counter_ [@Becht2016] and _xCell_ [@Aran2017] algorithms, respectively.
Differences in the predicted infiltration levels between the collagen clusters were investigated by Mann-Whitney test with r effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)). 
Populations differing between the clusters with pFDR < 0.05 and at least weak effect size (r $\geq$ 0.1) were considered biologically relevant. 
The analysis results are shown in __Supplementary Figure S\@ref(fig:fig-intiltration)__ and __Supplementary Table S\@ref(tab:tab-infiltration)__.

## Gene set variance analysis of Reactome pathway gene signatures

ssGSEA scores [@Hanzelmann2013] of the Reactome pathway gene signatures were compared between the collagen clusters by two-tailed T test with Cohen's d effect size statistic. 
Signatures found to be significantly regulated (pFDR < 0.05) with at least weak effect size (d $\geq$ 0.2) in at least five cohorts were further investigated. 
To get additional functional insight and for purposes of visualization, these common regulated signatures were classified by their co-expression patterns in the TCGA data set by unsupervised KMEANS clustering (function `kcluster()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)). 
The analysis results are presented in __Supplementary Figure S\@ref(fig:fig-biology)__ and __Supplementary Table S\@ref(tab:tab-gsva)__.

## Differential gene expression and GO enrichment analysis

Differences in whole-transcriptome gene expression between the collagen clusters were investigated by FDR-corrected two-tailed T test with Cohen's d effect size statistic (function `test_two_groups()`, package [_microViz_](https://github.com/PiotrTymoszuk/microViz)). 
Genes significantly differentially expressed between the clusters were identified by the significance cutoff (pFDR < 0.05) and at least weak effect size of the expression differences (d $\geq$ 0.2). 
Genes found to be differentially regulated between the clusters are listed in __Supplementary Table S\@ref(tab:tab-dge)__.

Biological process GO enrichment within genes upregulated in each of the collagen high and collagen low clusters was analyzed with the _limma_ package _goana_ algorithm [@Young2010], re-implemented by the development package [_microViz_](https://github.com/PiotrTymoszuk/microViz) (function `GOana()`). 
Enrichment p values were adjusted for multiple testing with the FDR method. 
Odds ratio (OR) calculated for a GO term with the following formula served as a measure of effect size:

$$OR_{GO} = \frac{N_{GO|DGE} \times N_{total}}{N_{GO|total} \times N_{DGE}}$$
where $N_{GO|DGE}$ stands for the number of differentially regulated genes assigned to the GO term, 
$N_{total}$ is the total number of investigated genes, 
$N_{GO|total}$ is the number of investigated genes assigned to the GO term, and 
$N_{DGE}$ stands for the number of differentially regulated genes. 
GO terms found to be significantly enriched with at least weak effect size (pFDR < 0.05, OR $\geq$ 1.44) in at least five cohorts were further investigated. 
For additional functional insight, such common enriched GO terms were subjected to multi-dimensional scaling and unsupervised KMEANS clustering with pairwise Wang distance as a measure of semantic similarity (functions `go_sem()` and `kcluster()`, packages _GOSemSim_, [_microViz_](https://github.com/PiotrTymoszuk/microViz) and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) [@Yu2010]. 
Results of the GO enrichment analysis are shown in __Supplementary Figure S\@ref(fig:fig-biology)__ and __Supplementary Table S\@ref(tab:tab-go-enrichment)__.

## Transcriptional regulons and signaling pathway activity in the collagen clusters

Modulation of transcriptional regulons, i.e. sets of genes whose expression is controlled by a common transcriptional factor, and activity of selected signaling pathways in collagen high cancers as compared with collagen low tumors were explored by linear modeling with the _collecTRI_ [@Muller-Dott2023] and _PROGENy_ [@Schubert2018] databases, respectively. 
Liner models were constructed for whole-transcriptome effect sizes of gene regulation (Cohen's d, see: __Differential gene expression and GO enrichment analysis__) with the `run_ulm()` and `run_mlm()` tools from the _decoupleR_ package [@Badia-I-Mompel2022]. 
In each case, linear model score (LM score) served as a metric of activity magnitude. 
P values (LM score $\neq$ 0) were corrected for multiple testing with the FDR method. 
Regulons and signaling pathways found to be significantly activated or inhibited in at least five cohorts were further discussed. 
The analysis results are presented in __Supplementary Figure S\@ref(fig:fig-signaling)__,and __Supplementary Tables S\@ref(tab:tab-regulons)__ and __S\@ref(tab:tab-signaling)__.

## Activity of metabolic reactions in the collagen clusters

Rules of assignment of genes to biochemical reactions were retrieved from the Recon2 human metabolism model available via the BiGG database [@King2016] and the R package _BiGGR_ [@Gavai2015]. 
Estimates and standard errors of $log_2$ fold-regulation of expression for all available genes between the collagen clusters were calculated by two-tailed T test as described in __Differential gene expression and GO enrichment analysis__. 
Estimates of biochemical pathway fold-regulation were computed by evaluation of the gene assignment rules in the Recon2 model. 
The 'gene~A~ OR gene~B~' operator was interpreted as arithmetic mean of expression regulation estimates for the genes A and B. 
The 'gene~A~ AND gene~B~' operator was interpreted as minimum of expression regulation estimates for the gene A and gene B [@Gavai2015]. 
Standard deviation, 95% confidence intervals and p values for the predicted reaction regulation estimates were obtained by a Monte Carlo simulation with n = 3000 draws from normal distribution of gene expression regulation estimates (mean: gene expression regulation estimate, standard deviation: standard error of the gene expression regulation estimate) [@Gavai2015; @Pichler2023]. 
P values were corrected for multiple testing with the FDR method. 
The analysis was done with the packages _BiGGR_ [@Gavai2015] and [_biggrExtra_](https://github.com/PiotrTymoszuk/biggrExtra) [@Pichler2023] (function `build_geneSBML()`). 
Biochemical reactions predicted to be significantly activated or inhibited in the collagen high cluster as compared with the collagen low cluster are listed in __Supplementary Table S\@ref(tab:tab-reactions)__.

Enrichment of significantly activated or inhibited biochemical reactions within the Recon model metabolic subsystems [@King2016] was investigated by comparing the frequency of significantly activated or inhibited reactions in the given subsystem with 10000 random samples from the total reaction pool (function `suba()`, package [_biggrExtra_](https://github.com/PiotrTymoszuk/biggrExtra)). 
The magnitude of enrichment was assessed with OR defined for the $i$-th metabolic subsystem as follows:

$$OR_i = \frac{n_{regulated, i} \times n_{total}}{n_i \times n_{regulated}}$$

where $n_{regulated, i}$ denotes the number of significantly activated or inhibited biochemical reactions within the $i$-th metabolic subsystem, 
$n_{total}$ denotes the total number of investigated reactions, 
$n_i$ denotes the number of metabolic reactions within the $i$-th metabolic subsystem, and 
$n_{regulated}$ is the total number of significantly activated or inhibited biochemical reactions. 
Common significantly activated or inhibited metabolic subsystems were identified as metabolic subsystems significantly activated or inhibited in at least four cohorts, with at least weak effect size defined by OR $\geq$ 1.44 (__Supplementary Figure S\@ref(fig:fig-metabolism)__, __Supplementary Table S\@ref(tab:tab-subsystems)__).

## Multi-parameter survival modeling with transcriptome data

For survival analyses, $log_2$-transformed expression values of the collagen-related genes (__Supplementary Table S2__) were subjected to batch-adjustment with the ComBat algorithm [@Leek2012]. 
Subsequently, the ComBat-adjusted expression data sets of the GSE54460, GSE70768, GSE70769, GSE220095 cohorts were merged (further referred to as 'pooled GEO' cohort).

Biochemical relapse-free survival in the pooled GEO training cohort was modeled with normalized (Z-scores) expression values of the collagen-related genes as explanatory variables by the GBM algorithm [@Friedman2001; @Greenwell2022; @Natekin2013; @Friedman2002]. 
Selection of the optimal set of modeling parameters (number of trees, shrinkage factor, interaction depth, and minimal number of observations in the tree node) was accomplished by tuning with 10-fold cross-validation and out-of-fold model deviance as performance statistic (__Supplementary Table S\@ref(tab:tab-tuning)__). 
The tuning was performed with an in-house developed script. 
The GBM model was trained in the pooled GEO cohort with the optimal tuning parameter set with the `gbm()` function (package _gbm_). 
The Transcriptomic Collagen Score was defined as the linear predictor score of the GBM model and was computed for the best model iteration (function `gbm.perf(method = 'cv')`) with the `predict()` method and expression data from the training pooled GEO cohort and the TCGA and DKFZ test collectives. 
Importance of the explanatory variables for the predictive performance of the GBM model in the training pooled GEO cohort were measured with the relative influence method [@Friedman2001] and expressed as gradient of the sum of squared errors attributed to particular variables ($\Delta SSE$, __Figure 3A__).

Performance of the Transcriptomic Collagen Score at prediction of biochemical relapse-free survival was assessed by uni-variable Cox regression (functions `coxph()` and `as_coxex()`, packages _survival_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions)) [@Therneau2000]. 
The proportional hazard assumption of the Cox models was checked with the `summary(type = 'assumptions')` method, which implements the genuine `cox.zph()` algorithm [@Grambsch1994]. 
Harrell's concordance index [@Harrell1996] and integrated Brier score [@Graf1999] were employed to investigate the overall concordance between the predicted and observed survival and to assess overall calibration (method `summary(type = 'fit')`, package [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions), __Figure 3B__ and __Supplementary Table S\@ref(tab:tab-model-survival)__). 
Of note, as compared with several other machine learning algorithms (Ridge, Elastic Net, and LASSO Cox regression, SVM and Random Forest) [@Friedman2010; @Fouodo2018; @Ishwaran2008; @Ishwaran2022a], the GBM model demonstrated an excellent performance and the highest reproducibility at prediction of biochemical relapse-free survival in the investigated cohorts (__Supplementary Figure S\@ref(fig:fig-surv-algos)__, __Supplementary Table S\@ref(tab:tab-model-survival)__). 
Statistical significance of differences in biochemical relapse-free survival between patients stratified by tertiles of the Transcriptional Collagen Score was determined by Peto-Peto test (functions `surv_fit()` and `surv_pvalue()`, package _survminer_, __Figure 3C__).

In order to assess the confounding effect of single cohorts used to generate the pooled GEO data set on results of modeling of relapse-free survival with the best performing GBM algorithm, three nested Cox proportional hazard model were constructed with the pooled GEO data: (1) a model with cohort as the sole explanatory variable, (2) a model with normalized (Z-score) GBM linear predictor score as the sole explanatory variable, and (3) an additive full mode with cohort and normalized GBM linear predictor score as the explanatory variables. 
Performance of the models was evaluate with C-index and integrated Brier score as described above. 
Hazard ratio estimates for the cohort and GBM predictor terms with 95% confidence intervals were extracted from the models with the `summary()` method from the [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) package. 
Contribution of the cohort and GBM predictor terms to the additive model were explored by likelihood ratio test (LRT, method `anova()`, package _stats_). 
As presented in __Supplementary Figure S\@ref(fig:fig-surv-cohort)A__, predictive performance of the GBM predictor and the additive Cox model adjusted for the cohort effect was nearly identical. 
Furthermore, the cohort adjustment did not change the hazard ratio estimate of the GBM term in comparison of inference statistics of the GBM predictor-only and the additive Cox model. 
Finally, the highly significant results of the LRT for the GBM predictor term and non-significant LRT results for the cohort term suggest strongly that the GBM predictor but not the cohort effect contribute substantially to the additive model (__Supplementary Figure S\@ref(fig:fig-surv-cohort)B__). 
Collectively, these results let us conclude that the highly predictive results of GBM survival modeling in the GOE pooled cohort are independent from differences in relapse-free survival in single GEO cohorts.

## Survival analysis for single collagen-related transcripts

The univariable analysis of biochemical relapse-free survival for single collagen-related genes was done in the pooled GEO, TCGA and DKFZ data sets described above for the multi-parameter survival modeling. 
For each of the collagen-related genes, cancer patients were classified as high and low expressors by an expression cutoff corresponding to the largest difference in biochemical relapse-free survival measured by FDR-corrected Mentel-Henszel test (function `find_cutoff()`, package [_kmOptimizer_](https://github.com/PiotrTymoszuk/kmOptimizer), minimal expression strata size set to 25% of the observations). 
Hazard ratio (HR) with 95% confidence intervals for relapse risk in high expressors as compared with low expressors was calculated with the canonical _survival_ package functions _coxph()_, _coef_ and _confint()_. 
Genes associated with unfavorable prognosis were defined by pFDR < 0.05 and HR > 1 for the high vs low expressor comparison. 
Genes associated with favorable prognosis were defined by pFDR < 0.05 and HR < 1 (__Supplementary Figure S\@ref(fig:fig-surv-uni-volcano)__). 
Common survival markers were defined as genes significantly associated with survival in all cohorts (__Supplementary Figure S\@ref(fig:fig-surv-uni-forest)__ - __S\@ref(fig:fig-surv-uni-detal3)__, __Supplementary Table S\@ref(tab:tab-rfs-univariable)__).

## Data and code availability

Publicly available data sets were analyzed. 
Formatted data sets used for analyses will be made available upon request to the corresponding author. 
The transcriptome R analysis pipeline is available from GitHub (https://github.com/PiotrTymoszuk/collagen_pca).

\newpage

# Supplementary Tables

```{r tab-cohorts, tab.cap = 'Characteristic of the analyzed cohorts. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Qualitative variables are presented as percentages of categories within the complete observation set.'}

flextable::flextable(suppl_paper_tbl$cohorts) %>% 
  set_widths(c(3, rep(2.2, 6))) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('PSA: prostate-specific antigen.'), 
           part = 'header', 
           ref_symbols = 'a') %>%
  theme_vanilla

```

\newpage

```{r tab-genes, tab.cap = 'Collagen-related genes and their classification.'}

flextable::flextable(suppl_paper_tbl$genes) %>% 
  set_widths(widths = c(4.5, 3.5, 3)) %>% 
  italic(j = 2, part = 'body') %>% 
  merge_v(1) %>% 
  theme_vanilla

```

\newpage

```{r tab-norm-tumor, tab.cap = "Expression of the collagen pathway genes in the malignant and benign tissue compared by paired T test with Cohen's d effect size statistic. P values were corrected for multiple testing with the false discovery rate method. log2-transformed expression values are presented as medians with interquartile ranges (IQR) and ranges. The table is available as a supplementary Excel file."}

suppl_paper_tbl$norm_tumor[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r tab-gleason, tab.cap = 'Expression of the collagen pathway genes in cancer samples stratified by the ISUP risk groups compared by one-way ANOVA with eta-square effect size statistic. P values were corrected for multiple testing with the false discovery rate method. log2-transformed expression values are presented as medians with interquartile ranges (IQR) and ranges. The table is available as a supplementary Excel file.'}

suppl_paper_tbl$gleason[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r tab-cluster, tab.cap = "Expression of the cluster-defining collagen pathway genes in the collagen clusters of prostate cancer. Statistical significance was assessed by two-tailed T test with Cohen's d effect size statistic. P values were corrected for multiple testing with the false discovery rate method. log2-transformed expression values are presented as medians with interquartile ranges (IQR) and ranges. The table is available as a supplementary Excel file."}

suppl_paper_tbl$cluster[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r tab-clinic, tab.cap = 'Clinical characteristic of the collagen clusters. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Nominal variables are presented as percentages and counts of categories within the cluster.'}

flextable::flextable(suppl_paper_tbl$clinic) %>% 
  set_widths(c(2.8, 3.5, 4, 4, 2.8, 2.8)) %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('PSA: prostate-specific antigen; ISUP: International Society of Urological Pathology risk stratification system.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 5:6, 
           value = as_paragraph('Qualitative variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Mann-Whitney test with r effect size statistic. P values corrected for multiple testing with the false discovery rate.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

\newpage

```{r tab-infiltration, tab.cap = 'Non-malignant cell numbers predicted for the collagen clusters by the MCP Counter and xCell algorithms. Statistical significance was assessed by Mann-Whitney test with r effect size statistic. P values were corrected for multiple testing with the false discovery method. The table is available as a supplementary Excel file.'}

suppl_paper_tbl$infiltration[1, 1] %>% 
  mutate(Algorithm = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r tab-gsva, tab.cap = "Gene set variation analysis with the Reactome pathway gene signatures. Differences in ssGSEA scores between collagen high and collagen low cancers were investigated by two-tailed T test with Cohen's d effect size statistic. Results for signatures significantly regulated with at least weak effect size (d at least 0.2) in at least five cohorts are presented. P values were corrected for multiple testing with the false discovery rate method (FDR). The table is available as a supplementary Excel file."}

suppl_paper_tbl$reactome[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r tab-dge, tab.cap = "Genes differentially expressed in the collagen high cluster as compared with collagen low cancers were identified by two-tailed T test with the 1.25-fold regulation cutoff and the Cohen's d effect size statistic of 0.2 P values were corrected for multiple testing with the false discovery rate method (FDR). The table is available as a supplementary Excel file."}

suppl_paper_tbl$dge[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r tab-go-enrichment, tab.cap = 'Biological process gene ontology (GO) term enrichment within genes differentially regulated in the collagen clusters. The enrichment analysis was performed with goana tool, enrichment p values were corrected for multiple testing with the false discovery rate (FDR) method. Significant enrichment was defined by pFDR < 0.05 and odds ratio (OR) for enrichment within differentially regulated genes of at least 1.44. OR for enrichment within genes upregulated in the collagen high and collagen low clusters are presented for significant GO terms shared by at least five cohorts. The table is available as a supplementary Excel file.'}

suppl_paper_tbl$go_enrichment[1, 1] %>% 
  mutate(`Collagen cluster` = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r tab-regulons, tab.cap = 'Activity of transcriptional regulons in the collagen high cluster as compared with the collagen low cluster predicted by the collecTRI model. Regulon activity was estimated with uni-parameter linear modeling with whole-transcriptome effect sizes of differential gene expression, p values were corrected for multiple testing with the false discovery rate (FDR) method. Linear model scores are presented for regulons significantly activated or inhibited in at least five cohorts. The table is available as a supplementary Excel file.'}

suppl_paper_tbl$regulons[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r tab-signaling, tab.cap = 'Activity of signaling pathways in the collagen high cluster as compared with the collagen low cluster predicted by the PROGENy model. Pathway activity was estimated with multi-parameter linear modeling with whole-transcriptome effect sizes of differential gene expression. P values were corrected for multiple testing with the false discovery rate (FDR) method,  linear model scores serve as measures of pathway activity. The table is available as a supplementary Excel file.'}

suppl_paper_tbl$signaling[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r tab-reactions, tab.cap = 'Biochemical reactions predicted to be significantly activated in collagen high as compared with collagen low cancers. Statistical significance was determined by a Monte Carlo simulation. P values were corrected for multiple testing with the false discovery rate (FDR) method. The table is available as a supplementary Excel file.'}

suppl_paper_tbl$metabolism[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r tab-subsystems, tab.cap = 'Results of enrichment analysis for significantly activated and inhibited biochemical reactions within the Recon metabolism subsystem. Statistical significance was determined by random sampling from the entire reaction pool p values were corrected for multiple testing with the false discovery rate (FDR) method. Effect size of enrichment of the subsystem in significantly activated or inhibited reactions was measured by odds ratio (OR) statistic. The table is available as a supplementary Excel file.'}

suppl_paper_tbl$subsystems[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r tab-tuning, tab.cap = 'Selection of the optimal parameters of machine learining survival models of biochemical relapse-free survival with expression of the collagen-related genes. The selection process was accomplished by cross-validation tuning in the pooled GEO cohort.'}

flextable::flextable(suppl_paper_tbl$tuning) %>% 
  set_widths(c(3.5, 4, 3, 3)) %>% 
  merge_v(1:2, combine = TRUE) %>% 
  footnote(1, 1, 
           value = as_paragraph('SVM: Support Vector Machines; GBM: Gradient Boosted Machines.'),
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('CV: cross-validation.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

\newpage

```{r tab-model-survival, tab.cap = 'Performance of machine learning models at prediction of biochemical relapse-free survival with expression of the collagen-related genes.'}

flextable::flextable(suppl_paper_tbl$model_survival) %>% 
  set_widths(c(2.8, 2.7, 2.7, 4.6, 2.5)) %>% 
  merge_v(1:2) %>% 
  footnote(1, 1, 
           value = as_paragraph('SVM: Support Vector Machines; GBM: Gradient Boosted Machines.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 3, 
           value = as_paragraph('pooled GEO: GSE54460, GSE70768, GSE70769, and GSE220095'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

\newpage

```{r tab-rfs-univariable, tab.cap = 'Results of univariable analysis of biochemical relapse-free survival with expression of the collagen-related genes. Prostate cancer patients were stratified by expression cutoffs corresponding to the largest difference in survival assessed by Mentel-Henszel test. Genes found to be significantly associated with the survival in all analyzed cohorts (pooled GEO, TCGA and DKFZ) are presented. The table is available in a supplementary Excel file.'}

suppl_paper_tbl$rfs_univariable[1, 1] %>% 
  mutate( `Association with survival` = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(18)

```

\newpage

# Supplementary Figures

```{r fig-norm-tumor, fig.width = figur::convert(suppl_paper_fig$norm_tumor, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$norm_tumor, to = 'in')$h, fig.cap = 'Expression of collagen-related genes in the normal prostate and prostate cancer tissue.'}

suppl_paper_fig$norm_tumor$plot

```

__Supplementary Figure S\@ref(fig:fig-norm-tumor). Expression of collagen-related genes in the normal prostate and prostate cancer tissue.__

_Differences in $log_2$-transformed expression of `r nrow(globals$genes_interest)` collagen-related genes between donor-matched pairs of the prostate cancer and benign tissue were assessed by paired T test with Cohen's effect size statistic in the GSE70768 and TCGA cohorts. P values were corrected for multiple testing with the false discovery rate (FDR) method. Full analysis results are listed in Supplementary Table S\@ref(tab:tab-norm-tumor)._

_(A) Numbers of significantly up- and downregulated genes in the tumor tissue as compared with the benign tissue in the investigated cohorts presented in Venn plots._

_(B) Mean normalized $log_2$ expression of the collagen-related genes presented as lines. Tinted regions represent the 2 $\times$ SEM (standard error of the mean) intervals. Significant effects are highlighted with bold font in the Y axis. Numbers of tissue pairs are displayed in the plot captions._

\newpage

```{r fig-clust-dev, fig.width = figur::convert(suppl_paper_fig$clust_dev, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$clust_dev, to = 'in')$h, fig.cap = 'Semi-supervised clustering of prostate cancer samples in respect to expression of collagen-related genes.'}

suppl_paper_fig$clust_dev$plot

```

__Supplementary Figure S\@ref(fig:fig-clust-dev). Semi-supervised clustering of prostate cancer samples in respect to expression of collagen-related genes.__

_Cancer samples in the TCGA training cohort (n = `r nrow(clust_globals$data$tcga)`) were clustered in respect to normalized, $log_2$-transformed expression levels of the collagen-related genes of interest with the PAM (partition around medoids) algorithm with cosine distance metric. Two clusters were defined: collagen low and collagen high. Assignment of the tumor samples in the training GSE54460 (n = `r nrow(clust_globals$data$gse54460)`), GSE70768 (n = `r nrow(clust_globals$data$gse70768)`), GSE70769 (n = `r nrow(clust_globals$data$gse70769)`), GSE220095 (n = `r nrow(clust_globals$data$gse220095)`), and DKFZ collective (n = `r nrow(clust_globals$data$dkfz)`) to the collagen clusters was accomplished by an inverse distance-weighted k-nearest neighbor classifier._

_(A) Comparison of performance of several clustering algorithms (PAM: partition around medoids, HCL: hierarchical clustering/Ward D2 and KMEANS) and distance metrics (Euclidean, Manhattan, squared Euclidean and cosine) in 5-fold cross-validation of the TCGA data set gauged by mean silhouette width as a measure of cluster separation, fraction of observations with negative silhouette widths indicative of possible misclassification, explained clustering variance and fraction of misclassified 5-nearest neighbors, cluster assignment accuracy in the cross-validation folds. Note the superior performance of the PAM/cosine algorithm used for definition of the collagen clusters._

_(B) Choice of cluster number (k = 2) for PAM/cosine clustering of cancer samples in the TCGA cohort motivated by the peak mean silhouette width statistic._

_(C) Distribution of sizes of the collagen clusters in the TCGA training cohort and the validation collectives. Percentages of samples assigned to the collagen clusters are presented in a stack plot._

_(D) Quality of collagen clusters in the training cohort (TCGA) and validation collectives. Cluster separation was assessed by mean silhouette width, misclassification rate is expressed as fraction of observations with negative silhouette widths. Explanatory performance was gauged by fraction of explained clustering variance. Neighborhood preservation was assessed by mean fraction of 5-nearest neighbors placed in the same cluster._

\newpage

```{r fig-clusters, fig.width = figur::convert(suppl_paper_fig$clusters, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$clusters, to = 'in')$h, fig.cap = 'Expression of the collagen-related genes in the collagen clusters of prostate cancers.'}

suppl_paper_fig$clusters$plot

```

__Supplementary Figure S\@ref(fig:fig-clusters). Expression of the collagen-related genes in the collagen clusters of prostate cancers.__

_The collagen clusters of prostate cancer samples were developed by semi-supervised PAM clustering as presented in Supplementary Figure S\@ref(fig:fig-clust-dev). Results of comparison of levels of the cluster-defining factors between the collagen clusters are listed in Supplementary Table S\@ref(tab:tab-cluster)._

_(A) Uniform manifold approximation and projection (UMAP) layout of normalized $log_2$-transformed expression values of the collagen-related genes utilized for definition of the collagen clusters. Each point represents a single cancer sample, cluster assignment is color-coded. Numbers of cancer samples in the collagen clusters are indicated in the figure legends._

_(B) Mean normalized $log_2$-transformed expression values of the cluster-defining collagen-related genes visualized in a heat map. The cluster-defining genes were classified by their biological function (Pro: proline metabolism, ECM: extracellular matrix, Adh.: adhesion)._

\newpage

```{r fig-clust-tumor, fig.width = figur::convert(suppl_paper_fig$clust_tumor, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$clust_tumor, to = 'in')$h, fig.cap = 'ISUP and pathological tumor stage in the collagen clusters.'}

suppl_paper_fig$clust_tumor$plot

```

__Supplementary Figure S\@ref(fig:fig-clust-tumor). ISUP and pathological tumor stage in the collagen clusters.__

_Differences in distribution of ISUP risk strata (International Society of Urological Pathology: ISUP1, ISUP2 and ISUP3+) and pathological tumor stages (B) between the collagen clusters were investigated by $\chi^2$ test with Cramer V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Percentages of samples assigned to the ISUP strata or tumor stages within each cluster are shown in stack plots. Effect sizes and p values are displayed in the plot facets. Numbers of complete observations in the clusters are indicated in the Y axes. Full clinical and pathological characteristic of the collagen clusters is shown in Supplementary Table S\@ref(tab:tab-clinic)._

\newpage

```{r fig-intiltration, fig.width = figur::convert(suppl_paper_fig$infiltration, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$infiltration, to = 'in')$h, fig.cap = 'Cancer-associated fibroblasts and endothelial cell infiltration in the collagen clusters.'}

suppl_paper_fig$infiltration$plot

```

__Supplementary Figure S\@ref(fig:fig-intiltration). Cancer-associated fibroblasts and endothelial cell infiltration in the collagen clusters.__

_Levels of cancer-associated fibroblasts (A) and endothelial cells (B) in the collagen clusters were predicted by the MCP Counter and xCell algorithms. Statistical significance of differences between the clusters was assessed by Mann-Whitney test with r effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Median infiltration levels with interquartile ranges are visualized as boxes, whiskers span over 150% of the interquartile ranges. Points represent single cancer samples. Effect sizes and p values are displayed in the Y axes, significant effects are highlighted in bold. GSE54460: `r format_strata_n(ana_globals$n_numbers$gse54460)`, GSE70768: `r format_strata_n(ana_globals$n_numbers$gse70768)`, GSE70769: `r format_strata_n(ana_globals$n_numbers$gse70769)`, GSE220095: `r format_strata_n(ana_globals$n_numbers$gse220095)`, DKFZ: `r format_strata_n(ana_globals$n_numbers$dkfz)`, TCGA: `r format_strata_n(ana_globals$n_numbers$tcga)`. Results of the analysis for other cell populations are listed in Supplementary Table S\@ref(tab:tab-infiltration)._

\newpage

```{r fig-biology, fig.width = figur::convert(suppl_paper_fig$biology, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$biology, to = 'in')$h, fig.cap = 'Gene set variation analysis of Reactome pathway gene signatures and gene ontology term enrichment analysis for the collagen clusters.'}

suppl_paper_fig$biology$plot

```

__Supplementary Figure S\@ref(fig:fig-biology). Gene set variation analysis of Reactome pathway gene signatures and gene ontology term enrichment analysis for the collagen clusters.__

_(A) Gene set variation analysis (GSVA) with Reactome pathway gene signatures. Statistical significance for differences in signature single sample gene set enrichment analysis scores (ssGSEA) between the collagen clusters was determined by two-tailed T test with Cohen's d effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Mean ssGSEA scores in the clusters and cohorts for signatures significantly regulated in at least five cohorts with at least weak effect size (d $\geq$ 0.2) are presented in a heat map. Full analysis results are presented in Supplementary Table S\@ref(tab:tab-gsva). The common regulated gene signatures were classified in respect their co-expression patterns by KMEANS clustering; signature classification is indicated in the Y axis facets of the heat map (ECM: extracellular matrix, GF: growth factor, GPCR: G protein-coupled receptor)._

_(B) Genes differentially expressed between the collagen clusters were identified by false discovery rate (FDR) corrected two-tailed T test with Cohen's d effect size statistic. Differentially expressed genes were defined by the pFDR < 0.05 and d $\geq$ 0.2 cutoffs (Supplementary Table \@ref(tab:tab-dge)). Biological process gene ontology (GO) term enrichment within genes significantly unregulated in the collagen high and the collagen low clusters was investigated with the goana algorithm. Enrichment p values were adjusted for multiple testing with the FDR method, odds ratio of enrichment over the whole-transcriptome GO frequency served as an effect size statistic. GO terms significantly enriched with at least weak effect size (OR $\geq$ 1.44) in at least five cohorts were subjected to multi-dimensional scaling and unsupervised KMEANS clustering in respect to their semantic similarity. Multi-dimensional scaling layouts of Wang distances between the common enriched GO terms are presented. Each point represents a single GO term, GO cluster assignment is color coded. Complete results of the GO enrichment analysis are listed in Supplementary Table S\@ref(tab:tab-go-enrichment)._

\newpage

```{r fig-signaling, fig.width = figur::convert(suppl_paper_fig$signaling, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$signaling, to = 'in')$h, fig.cap = 'Transcriptional regulons and signaling pathway activity in the collagen clusters.'}

suppl_paper_fig$signaling$plot

```

__Supplementary Figure S\@ref(fig:fig-signaling). Transcriptional regulons and signaling pathway activity in the collagen clusters.__

_Differences in activity of transcriptional regulons (gene sets controlled by a common transcription factor, A) and signaling pathways (B) in the collagen high cluster as compared with the collagen low cluster were assessed by linear modeling with the collecTRI and PROGENy databases implemented by the decoupler algorithm. Linear model score (LM score) served as a measure of regulon or pathway activity, p values for significant modulation of activity (LM score $\neq$ 0) were corrected for multiple testing with the false discovery rate method. Full analysis results are listed in Supplementary Tables S\@ref(tab:tab-regulons) and S\@ref(tab:tab-signaling)._

_(A) LM scores of top modulated transcriptional regulons found to be significantly activated or inhibited in at least five cohorts visualized as bubble plots. Point color codes for the regulation sign, point size codes for absolute value of the LM score. LM scores for regulon activity in particular cohorts are presented next to the data points._

_(B) LM scores of signaling pathways found to be significantly activated or inhibited in at least five cohorts visualized as bubble plots. Point color codes for the regulation sign, point size codes for absolute value of the LM score. LM scores for regulon activity in particular cohorts are presented next to the data points._

\newpage

```{r fig-metabolism, fig.width = figur::convert(suppl_paper_fig$metabolism, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$metabolism, to = 'in')$h, fig.cap = 'Metabolism in the collagen clusters.'}

suppl_paper_fig$metabolism$plot

```

__Supplementary Figure S\@ref(fig:fig-metabolism). Metabolism in the collagen clusters.__

_Modulation of Recon2 model metabolic reactions in collagen high tumors as compared with collagen low cancers was predicted by Monte Carlo simulation implemented by the BiGGR and biggrExtra algorithms based on whole-genome differential gene expression estimates (Supplementary Table \@ref(tab:tab-reactions)). Enrichment of significantly activated and significantly inhibited reactions in the Recon metabolic subsystems was assessed by comparing the observed frequency of the subsystem's significantly regulated reactions with 10000 random draws from the entire reaction pool. Enrichment p values were corrected for multiple testing with the false discovery rate method. Enrichment odds ratios (OR) for significantly enriched metabolic subsystems with at least weak effect size (OR $\geq$ 1.44)  shared by at least five cohorts are presented in a bubble plot. Point color codes for reaction modulation status, point sizes represent absolute values of OR. Points are labeled with their OR values. Full analysis results are presented in Supplementary Table S\@ref(tab:tab-subsystems)._

\newpage

```{r fig-surv-algos, fig.width = figur::convert(suppl_paper_fig$surv_algos, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_algos, to = 'in')$h, fig.cap = 'Comparison of performance of various machine learning algorithms at prediction of biochemical relapse-free survival with expression of the collagen-related genes.'}

suppl_paper_fig$surv_algos$plot

```

__Supplementary Figure S\@ref(fig:fig-surv-algos). Comparison of performance of various machine learning algorithms at prediction of biochemical relapse-free survival with expression of the collagen-related genes.__

_Biochemical relapse-free survival was modeled with normalized, $log_2$-transformed and ComBat-adjusted expression levels of the collagen-related genes by various machine learning algorithms (Ridge Cox regression, Elastic Net Cox regression, LASSO Cox regression, Support Vector Machines [SVM], Random Forest, and Gradient Boosted Machines [GBM]). Their performance at prediction of biochemical relapse-free survival in the training pooled GEO cohort, and the test TCGA and DKFZ collectives was assessed by Harrell's concordance index (C-index) and integrated Brier scores (IBS). High C-index values indicate good concordance between the observed and predicted survival. Low values of IBS indicate good model calibration._
_C-index and IBS values are presented in scatter plots. Each point represents a single algorithm. Values of C-index and IBS expected for a model predicting survival at random are visualized as dashed lines._

\newpage

```{r fig-surv-cohort, fig.width = figur::convert(suppl_paper_fig$surv_cohort, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_cohort, to = 'in')$h, fig.cap = 'Assessment of the confounding effect of the cohort on results of survival GBM modeling in the pooled GEO data set.'}

suppl_paper_fig$surv_cohort$plot

```

__Supplementary Figure S\@ref(fig:fig-surv-cohort). Assessment of the confounding effect of the cohort on results of survival GBM modeling in the pooled GEO data set.__ 
_To investigate the potentially confounding effect of single cohorts in the pooled GEO data set on results of GBM survival modeling, three Cox models were constructed: a cohort-only model, a model with the GBM linear predictor score as the sole explanatory factor, and an additive model with the cohort and GBM predictor score as explanatory factors._
_(A) Evaluation of predictive performance with the Cox models in the pooled GEO cohort. Concordance between the predicted and observed survival was assessed by concordance index (C-index). Model calibration and forecast quality was measured with integrated Brier score (IBS). Single Cox models are visualized as points. Dashed lines represent values of C-index and IBS expected for a meaningless model._
_(B) Inference of the cohort and GBM predictor terms in the Cox models. Hazard ratios (HR) with are presented with their 95% confidence intervals (CI) in a Forest plot. Results of likelihood ratio test for contribution of the cohort and GBM predictor terms to the additive model are displayed in the plot facets._

\newpage

```{r fig-surv-uni-volcano, fig.width = figur::convert(suppl_paper_fig$surv_uni_volcano, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_uni_volcano, to = 'in')$h, fig.cap = 'Differences in biochemical relapse-free survival between high and low expressors of the collagen-related genes.'}

suppl_paper_fig$surv_uni_volcano$plot

```

__Supplementary Figure S\@ref(fig:fig-surv-uni-volcano). Differences in biochemical relapse-free survival between high and low expressors of the collagen-related genes.__

_Cancer patients were stratified by cutoffs of normalized expression for each collagen-related genes which corresponded to the largest differences in biochemical relapse-free survival assessed by false discovery rate (FDR) corrected Mentel-Henszel test into high and low expressors for particular genes. Relative risk of biochemical relapse in high expressors as compared with low expressors was modeled by Cox proportional hazard regression and expressed as hazard ratio (HR). Unfavorable survival markers were defined by pFDR < 0.05 and HR > 1. Favorable survival markers were defined by pFDR < 0.05 and HR < 1. HR and pFDR values are presented in Volcano plots. Each point represents a single collagen-related gene. Significance and sign of association with the relapse risk are color coded. Significant genes are labeled with their symbols. Dashed lines indicate the significance and regulation cutoffs._

\newpage

```{r fig-surv-uni-forest, fig.width = figur::convert(suppl_paper_fig$surv_uni_forest, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_uni_forest, to = 'in')$h, fig.cap = 'Collagen-related transcriptional markers of biochemical relapse risk shared by all investigated cohorts.'}

suppl_paper_fig$surv_uni_forest$plot

```

__Supplementary Figure S\@ref(fig:fig-surv-uni-forest). Collagen-related transcriptional markers of biochemical relapse risk shared by all investigated cohorts.__

_Differences in biochemical relapse-free survival in high expressors versus low expressors of each collagen-related gene were compared by Cox modeling and false discovery rate (FDR) corrected Mentel-Henszel test as presented in Supplementary Figure S\@ref(fig:fig-surv-uni-volcano). Unfavorable survival markers were defined by pFDR < 0.05 and hazard ratio (HR) > 1. Favorable survival markers were defined by pFDR < 0.05 and HR < 1. HR with 95% confidence intervals (CI) for unfavorable and favorable markers shared by all investigated cohorts are presented in a Forest plot._

\newpage

```{r fig-surv-uni-detail1, fig.width = figur::convert(suppl_paper_fig$surv_uni_detail1, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_uni_detail1, to = 'in')$h, fig.cap = 'Differences in biochemical relapse-free survival between high and low gene expressors: _COL1A1_, _COL3A1_, and _COL5A2_.'}

suppl_paper_fig$surv_uni_detail1$plot

```

__Supplementary Figure S\@ref(fig:fig-surv-uni-detail1). Differences in biochemical relapse-free survival between high and low gene expressors: _COL1A1_, _COL3A1_, and _COL5A2_.__

_Differences in biochemical relapse-free survival in high expressors versus low expressors of each collagen-related gene were compared by Cox modeling and false discovery rate (FDR) corrected Mentel-Henszel test as presented in Supplementary Figure S\@ref(fig:fig-surv-uni-volcano). Unfavorable survival markers were defined by pFDR < 0.05 and hazard ratio (HR) > 1. Differences in biochemical relapse-free survival for the COL1A1, COL3A1, and COL5A2 genes identified as unfavorable markers in all investigated cohorts are shown in Kaplan-Meier plots. Numbers of complete cases and relapses, expression cutoff values, and numbers of high and low expressors are displayed in the plot captions. HR with 95% confidence intervals and pFDR obtained in Mentel-Henszel test are indicated in the plots._

\newpage

```{r fig-surv-uni-detail2, fig.width = figur::convert(suppl_paper_fig$surv_uni_detail2, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_uni_detail2, to = 'in')$h, fig.cap = 'Differences in biochemical relapse-free survival between high and low gene expressors: _COL11A1_, _COL11A2_, _LOXL1_, and _SERPINH1_.'}

suppl_paper_fig$surv_uni_detail2$plot

```

__Supplementary Figure S\@ref(fig:fig-surv-uni-detail2). Differences in biochemical relapse-free survival between high and low gene expressors: _COL11A1_, _COL11A2_, _LOXL1_, and _SERPINH1_.__

_Differences in biochemical relapse-free survival in high expressors versus low expressors of each collagen-related gene were compared by Cox modeling and false discovery rate (FDR) corrected Mentel-Henszel test as presented in Supplementary Figure S\@ref(fig:fig-surv-uni-volcano). Unfavorable survival markers were defined by pFDR < 0.05 and hazard ratio (HR) > 1. Differences in biochemical relapse-free survival for the COL11A1, COL11A2, LOXL1, and SERPINH1 genes identified as unfavorable markers in all investigated cohorts are shown in Kaplan-Meier plots. Numbers of complete cases and relapses, expression cutoff values, and numbers of high and low expressors are displayed in the plot captions. HR with 95% confidence intervals and pFDR obtained in Mentel-Henszel test are indicated in the plots._

\newpage

```{r fig-surv-uni-detal3, fig.width = figur::convert(suppl_paper_fig$surv_uni_detal3, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_uni_detal3, to = 'in')$h, fig.cap = 'Differences in biochemical relapse-free survival between high and low gene expressors: _COL4A6_, _DST_, _LAMB3_, and _PCOLCE2_.'}

suppl_paper_fig$surv_uni_detal3$plot

```

__Supplementary Figure S\@ref(fig:fig-surv-uni-detal3). Differences in biochemical relapse-free survival between high and low gene expressors: _COL4A6_, _DST_, _LAMB3_, and _PCOLCE2_.__

_Differences in biochemical relapse-free survival in high expressors versus low expressors of each collagen-related gene were compared by Cox modeling and false discovery rate (FDR) corrected Mentel-Henszel test as presented in Supplementary Figure S\@ref(fig:fig-surv-uni-volcano). Unfavorable survival markers were defined by pFDR < 0.05 and hazard ratio (HR) > 1. Differences in biochemical relapse-free survival for the COL4A6, DST, LAMB3, and PCOLCE2 genes identified as favorable markers in all investigated cohorts are shown in Kaplan-Meier plots. Numbers of complete cases and relapses, expression cutoff values, and numbers of high and low expressors are displayed in the plot captions. HR with 95% confidence intervals and pFDR obtained in Mentel-Henszel test are indicated in the plots._

\newpage

# References

### 2025-02-21 paper/markdown/manuscript_figures_tables.Rmd
Supplementary Figures
Supplementary figure legends

@@
 _Prostate cancer (PCA) patients in the pooled GEO (total: n = `r uni_globals$n_numbers$n_total["geo_pool"]`, biochemical relapse [BCR]: n = `r uni_globals$n_numbers$n_events["geo_pool"]`), TCGA (total: n = `r uni_globals$n_numbers$n_total["tcga"]`, BCR: n = `r uni_globals$n_numbers$n_events["tcga"]`), and DKFZ cohort (total: n = `r uni_globals$n_numbers$n_total["dkfz"]`, BCR: n = `r uni_globals$n_numbers$n_events["dkfz"]`) were classified as high and low expressors for each of the `r `length(globals$genes)` collagen-related transcripts with the expression cutoffs corresponding to the largest differences in survival between the high and low expressors._
-_BCR risk was compared between the high and low expressors for each gene by univariable Cox proportional hazard regression.
+_BCR risk was compared between the high and low expressors for each transcript by univariable Cox proportional hazard regression.
 P values were corrected for multiple testing with the false discovery rate (FDR) method._
@@
 
-_(B) BCR-free survival in PCA patients stratified by tertiles of the predictor scores of the best performing GBM model.
+_(B) BCR-free survival of PCA patients stratified by tertiles of the predictor scores of the best performing GBM model.
 Statistical significance of differences between the predictor score tertiles was determined by Peto-Peto test corrected for multiple testing wit the false discovery rate method.
@@
 
-_(C) Importance of the explanatory variables for BCR-risk prediction by the GBM model was gauged by difference in sum of squared error ($\Delta SSE$) attributed to inclusion of a particular collagen-related gene in the GBM learner ensemble.
+_(C) Importance of the explanatory variables for BCR-risk prediction by the GBM model was gauged by difference in sum of squared error ($\Delta SSE$) attributed to inclusion of a particular collagen-related transcript in the GBM learner ensemble.
 High $\Delta SSE$ values are characteristic for highly influential variables.

### 2025-02-21 paper/markdown/manuscript_supplement.Rmd
Supplementary Figures
Supplementary figure legends

---
title: "Prognostic and biological relevance of collagen-related genes in prostate cancer"
subtitle: "Supplementary material, transcriptome part"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: coll_biblio.bib

csl: frontiers-in-immunology.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

library(bookdown)
library(flextable)

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "png", 
                      dpi = 600, 
                      tab.cap.pre = 'Table ', 
                      tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

# Supplementary Figures

```{r fig-nets, fig.width = figur::convert(suppl_fig$nets, to = 'in')$w, fig.height = figur::convert(suppl_fig$nets, to = 'in')$h, fig.cap = 'Co-expression networks of the collagen-related transcripts in PCA tissue.'}

suppl_fig$nets$plot

```

__Supplementary Figure S\@ref(fig:fig-nets). Co-expression networks of the collagen-related transcripts in PCA tissue.__

_Co-expression networks of the collagen-related transcripts in PCA were constructed in the pooled GEO (cancer samples: n = `r nrow(net$data$geo_pool)`), TCGA (n = `r row(net$data$tcga)`), and DKFZ cohort (n = `r row(net$data$dkfz)`). The network edges were defined by pairwise correlations between ComBat-adjusted transcript levels in the cancer tissue with Spearman's $\rho \geq$ 0.5. Isolated vertices, i.e. transcripts without correlation partners, were removed._ 
_The co-expression networks were visualized with Fruchterman-Reingold algorithm. Each point represents a single transcript in the network; point color codes for functional classification of the transcript. Edges, i.e. correlations of expression levels with $\rho \geq$ 0.5, are depicted as gray lines. Numbers of transcripts (vertices) and correlations with $\rho \geq$ 0.5 (edges) are displayed in the plot captions._

\newpage

```{r fig-hubs, fig.width = figur::convert(suppl_fig$hubs, to = 'in')$w, fig.height = figur::convert(suppl_fig$hubs, to = 'in')$h, fig.cap = 'Vertex importance statistics for co-expression networks of the collagen-related transcripts in PCA.'}

suppl_fig$hubs$plot

```

__Supplementary Figure S\@ref(fig:fig-hubs). Vertex importance statistics for co-expression networks of the collagen-related transcripts in PCA.__

_Co-expression networks of the collagen-related transcripts were constructed in the pooled the pooled GEO (cancer samples: n = `r nrow(net$data$geo_pool)`), TCGA (n = `r row(net$data$tcga)`), and DKFZ cohort (n = `r row(net$data$dkfz)`) as presented in Supplementary Figure S\@ref(fig:fig-nets)._ 
_The following numeric statistics of vertex importance were computed for each collagen-related transcript: degree as a measure of connectivity (number of correlation partners with Spearman's $\rho \geq$ 0.5), hub score as a measure of overall correlation strength (eigenvector of the affinity matrix), and betweenness as a measure of connectivity and centrality (minimum/maximum-scaled number of the shortest paths between all pairs of vertices that pass through the vertex of interest)._ 
_The vertex importance metrics for the collagen-related transcripts in the co-expression networks are shown in dot plots. Each point represents a single collage-related transcript, point color codes for functional classification of the transcript._

\newpage

```{r fig-km-unfavorable, fig.width = figur::convert(suppl_fig$km_unfavorable, to = 'in')$w, fig.height = figur::convert(suppl_fig$km_unfavorable, to = 'in')$h, fig.cap = 'Top three unfavorable standalone collagen-related transcriptional markers of BCR-free survival in PCA.'}

suppl_fig$km_unfavorable$plot

```

__Supplementary Figure S\@ref(fig:fig-km-unfavorable). Top three unfavorable standalone collagen-related transcriptional markers of BCR-free survival in PCA.__

_Prostate cancer (PCA) patients in the pooled GEO (total: n = `r uni_globals$n_numbers$n_total["geo_pool"]`, biochemical relapse [BCR]: n = `r uni_globals$n_numbers$n_events["geo_pool"]`), TCGA (total: n = `r uni_globals$n_numbers$n_total["tcga"]`, BCR: n = `r uni_globals$n_numbers$n_events["tcga"]`), and DKFZ cohort (total: n = `r uni_globals$n_numbers$n_total["dkfz"]`, BCR: n = `r uni_globals$n_numbers$n_events["dkfz"]`) were classified as high and low expressors for each of the `r `length(globals$genes)` collagen-related transcripts with the expression cutoffs corresponding to the largest differences in survival between the high and low expressors._ 
_Collagen-related transcripts associated with unfavorable and favorable BCR prognosis were identified by univariable Cox proportional hazard regression as presented in Figure 1._ 
_COL1A1, COL11A1, and COL11A2 were identified as the strongest unfavorable BCR risk markers, as measured by mean Harrell's concordance index (measure of model accuracy) in the investigated cohorts. Fractions of BCR-free patients in the high and low expression strata are visualized in Kaplan-Meier plots. The log2 expression cutoffs used for the high/low expressor classification, hazard-ratios (HR) of BCR risk in the high as compared with the low expression strata with 95% confidence intervals (95% CI) are displayed in the plot captions. P values (p(HR = 0)) corrected for multiple testing with the false discovery rate method are shown in the plots. Numbers of patients and BCR cases in the expression strata are indicated in the plot legends._

\newpage

```{r fig-km-favorable, fig.width = figur::convert(suppl_fig$km_favorable, to = 'in')$w, fig.height = figur::convert(suppl_fig$km_favorable, to = 'in')$h, fig.cap = 'Top three favorable standalone collagen-related transcriptional markers of BCR-free survival in PCA.'}

suppl_fig$km_favorable$plot

```

__Supplementary Figure S\@ref(fig:fig-km-favorable). Top three favorable standalone collagen-related transcriptional markers of BCR-free survival in PCA.__

_Prostate cancer (PCA) patients in the pooled GEO (total: n = `r uni_globals$n_numbers$n_total["geo_pool"]`, biochemical relapse [BCR]: n = `r uni_globals$n_numbers$n_events["geo_pool"]`), TCGA (total: n = `r uni_globals$n_numbers$n_total["tcga"]`, BCR: n = `r uni_globals$n_numbers$n_events["tcga"]`), and DKFZ cohort (total: n = `r uni_globals$n_numbers$n_total["dkfz"]`, BCR: n = `r uni_globals$n_numbers$n_events["dkfz"]`) were classified as high and low expressors for each of the `r `length(globals$genes)` collagen-related transcripts with the expression cutoffs corresponding to the largest differences in survival between the high and low expressors._ 
_Collagen-related transcripts associated with unfavorable and favorable BCR prognosis were identified by univariable Cox proportional hazard regression as presented in Figure 1._ 
_COL4A6, LAMB3, and DST were identified as the strongest favorable BCR risk markers, as measured by mean Harrell's concordance index (measure of model accuracy) in the investigated cohorts. Fractions of BCR-free patients in the high and low expression strata are visualized in Kaplan-Meier plots. The log2 expression cutoffs used for the high/low expressor classification, hazard-ratios (HR) of BCR risk in the high as compared with the low expression strata with 95% confidence intervals (95% CI) are displayed in the plot captions. P values (p(HR = 0)) corrected for multiple testing with the false discovery rate method are shown in the plots. Numbers of patients and BCR cases in the expression strata are indicated in the plot legends._

\newpage

```{r fig-gbm-cohort, fig.width = figur::convert(suppl_fig$gbm_cohort, to = 'in')$w, fig.height = figur::convert(suppl_fig$gbm_cohort, to = 'in')$h, fig.cap = 'Investigation of confounding effect of the cohort on prediction of BCR-free survival by the GBM model in the pooled GEO cohort.'}

suppl_fig$gbm_cohort$plot

```

__Figure \@ref(fig:fig-gbm-cohort). Investigation of confounding effect of the cohort on prediction of BCR-free survival by the GBM model in the pooled GEO cohort.__

_Biochemical relapse (BCR) free survival was modeled by the GBM algorithm with ComBat-adjusted $log_2$-transformed expression levels of the collagen-related genes as explanatory factors. The Gradient Boosted Machines (GBM) model was trained in the pooled GEO cohort (total: n = `r surv_globals$n_numbers$n_total[1]`, BCR: n = `r surv_globals$n_numbers$n_events[1]`)._ 
_To investigate if and to which extend the BCR risk prediction depends on the confounding effect of single data sets constituting the pooled GEO cohort (`r globals$study_labels[geo_pool$studies] %>% collapse_and`), we constructed a canonical Cox proportional hazard regression model of BCR-free survival with the GBM predictor score and assignment to the data sets as independent variables. Performance and inference of this data set-adjusted GBM model was compared with performance and inference of the initial GBM model._

_Coefficient estimates expressed as hazard ratios (HR) with 95% confidence intervals of the data set-adjusted, the initial and the data set-only survival models are presented in a Forest plot. Significance of contribution of the cohort and GBM predictor score to the BCR-free survival prediction were assessed by likelihood ratio test (LRT), whose results are presented in the plot facets. Note the significant contribution of the GBM predictor score to the prediction of BCR-free survival and the non-significant effect of the data set._

\newpage

```{r fig-clinic, fig.width = figur::convert(suppl_fig$clinic, to = 'in')$w, fig.height = figur::convert(suppl_fig$clinic, to = 'in')$h, fig.cap = 'Modeling of BCR-free survival by GBM algorithm with clinical predictors and mRNA expression of the collagen-related genes.'}

suppl_fig$clinic$plot

```

__Figure \@ref(fig:fig-clinic). Modeling of BCR-free survival by GBM algorithm with clinical predictors and mRNA expression of the collagen-related genes.__

_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and blood PSA concentration at diagnosis), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors. Performance of the GBM models was evaluated in the pooled GEO training collective (total: n = `r nrow(surv_clin$score_tbl$geo_pool)`, BCR: n = `r sum(surv_clin$score_tbl$geo_pool$relapse)`), and TCGA (total: n = `r nrow(surv_clin$score_tbl$tcga)`, BCR: n = `r sum(surv_clin$score_tbl$tcga$relapse)`) and DKFZ (total: n = `r nrow(surv_clin$score_tbl$dkfz)`, BCR: n = `r sum(surv_clin$score_tbl$dkfz$relapse)`) validation cohorts._

_(A) Numeric metrics of survival model performance, Harrell's concordance index (C-index, measured of model accuracy) and Integrated Brier Score (IBS, measure of overall calibration, low values are characteristic for well calibrated models), are presented in a dot plot. Each point represents a single GBM model, point color codes for the model type. Models fit in the same cohort are connected with lines. Note the substantial improvement of the GBM model accuracy and calibration upon inclusion of the collagen-related transcript expression levels in the explanatory variable set in the pooled GEO and DKFZ cohorts._

_(B) Brier Scores for unique survival time points for the GBM models in the pooled GEO, TCGA, and DKFZ cohort. Low Brier Scores at a particular time point indicate accurate and confident prediction of the BCR risk. Note: gray lines represent random survival predictions by a null model._

_(C) Importance of the explanatory variables for BCR-risk prediction by the GBM model with the clinical and transcriptional predictors. The variable importance was computed as difference in sum of squared error ($\Delta SSE$) attributed to inclusion of a particular explanatory factor in the GBM learner ensemble. High $\Delta SSE$ values are characteristic for highly influential variables. $\Delta SSE$ values for the variables that contributed substantially to the BCR risk prediction ($\Delta SSE$ > 0) are presented in a bar plot._

\newpage

```{r fig-cl-plot, fig.width = figur::convert(suppl_fig$cl_plot, to = 'in')$w, fig.height = figur::convert(suppl_fig$cl_plot, to = 'in')$h, fig.cap = 'GBM modeling of BCR-free survival in PCA with clinical prediction and expression of collagen-related transcripts: survival in tertiles of predictor scores.'}

suppl_fig$cl_plot$plot

```

__Figure \@ref(fig:fig-cl-plot). GBM modeling of BCR-free survival in PCA with clinical prediction and expression of collagen-related transcripts: survival in tertiles of predictor scores.__

_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and blood PSA concentration at diagnosis), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors, as presented in Supplementary Figure S\@ref(fig:fig-cl-plot)._ 
_BCR-free survival in tertiles of the predictor scores of the clinical-only and the combined clinical - expression GBM models was visualized with Kaplan-Meier plots. Numbers of patients and BCR cases in the predictor score tertiles are indicated in the figure legend. P values for differences in survival between the tertiles obtained by Peto-Peto test are displayed in the plots. Note improved resolution of the survival curves for the combined clinical/expression model as compared with the clinical-only model in the pooled GEO and DKFZ cohort._

\newpage

```{r fig-os, fig.width = figur::convert(suppl_fig$os, to = 'in')$w, fig.height = figur::convert(suppl_fig$os, to = 'in')$h, fig.cap = 'Modeling of overall survival in PCA with expression levels of the collagen-related transcripts as explanatory factors.'}

suppl_fig$os$plot

```

__Figure \@ref(fig:fig-os). Modeling of overall survival in PCA with expression levels of the collagen-related transcripts as explanatory factors.__

_Overall survival in prostate cancer (PCA) was modeled by multi-parameter machine learning models with ComBat-processed $log_2$-transformed expression levels of `r length(globals$genes)` collagen-related transcripts as explanatory factors._ 
_The models were trained in the training subset of the GSE16560 cohort with six machine learning algorithms (RIDGE Cox regression, Elastic Net Cox regression, LASSO Cox regression, survival Support Vector Machines [SVM], survival Random Forest, and survival Gradient Boosted Machines [GBM])._
_Performance of the models was evaluated in the training (total patients: n = `r nrow(ridge_os$score_tbl$training)`, deaths: n = `r sum(ridge_os$score_tbl$training$death)`) and test (total patients: n = `r nrow(ridge_os$score_tbl$test)`, deaths: n = `r sum(ridge_os$score_tbl$test$death)`) subsets of the GSE16560 cohort._

_(A) Metrics of performance of the survival models, Harrell's concordance index (C-index, measure of model accuracy) and Integrated Brier Score (IBS, measure of overall model calibration, low values are characteristic for good calibration), are presented in dot plots. Dashed lines represent C-index and IBS values expected for random death risk prediction. Note the superior accuracy and calibration of the GBM model in the validation collectives._

_(B) Brier Scores for unique survival time points for the best performing RIDGE Cox model. Low Brier Scores at a particular time point indicate accurate and confident prediction of the BCR risk. Note: gray lines represent random survival predictions by a null model._

_(C) Overall survival of PCA patients stratified by tertiles of the predictor scores of the best performing RIDGE Cox model. Statistical significance of differences between the predictor score tertiles was determined by Peto-Peto test corrected for multiple testing wit the false discovery rate method. Fractions of alive patients and time after diagnosis are visualized in Kaplan-Meier plots. Numbers of patients and deaths in the predictor score tertiles are indicated in the legends. P values are displayed in the plots._

\newpage

# References

### 2025-02-21 BCR scripts/cohort.R

@@ -156,7 +156,7 @@
                     color_variable = 'model_type', 
                     palette = unname(globals$cohort_model_colors), 
                     labels = globals$cohort_model_labels, 
-                    plot_title = 'Cohort confounder, pooled GEO cohort', 
+                    plot_title = 'Data set confounder, pooled GEO cohort', 
                     plot_subtitle = 'GBM survival modeling')
   
 # Forest plot of the HRs -------
@@ -208,7 +208,7 @@
     theme(axis.title.y = element_blank(), 
           strip.text.y = element_text(angle = 0, 
                                       hjust = 0)) + 
-    labs(title = 'Cohort confounder, pooled GEO cohort', 
+    labs(title = 'Data set confounder, pooled GEO cohort', 
          subtitle = 'Inference and likelihood ratio test', 
          x = 'HR, 95% CI')
   

### 2025-02-21 paper scripts/figures.R

@@ -67,7 +67,7 @@
           theme(plot.subtitle = element_blank(), 
                 legend.position = 'bottom', 
                 legend.title = element_blank(), 
-                legend.text = element_text(size = 6))) %>% 
+                legend.text = element_text(size = 7))) %>% 
     plot_grid(plotlist = ., 
               ncol = 3, 
               align = 'hv', 
@@ -75,7 +75,7 @@
   
   fig$ml_mrna$bottom <- surv_plots$importance_plots$gbm + 
     guides(x = guide_axis(angle = 45)) + 
-    theme(axis.text.x = element_markdown(size = 7), 
+    theme(axis.text.x = element_markdown(size = 6.5), 
           strip.background.y = element_blank(), 
           strip.text.y = element_blank())
   
@@ -91,7 +91,7 @@
               label_size = 10) %>% 
     as_figure(label = 'ml_modeling_bcr_collagen_mrna', 
               ref_name = 'ml_mrna', 
-              caption = paste('Machine learnig modeling of BCR-free survival', 
+              caption = paste('Machine learning modeling of BCR-free survival', 
                               'with expression levels of the collagen-related', 
                               'transcripts as explanatory factors.'), 
               w = 180, 

### 2025-02-21 paper scripts/render.R

@@ -1,0 +1,27 @@
+# Renders the manuscript and the supplementary material
+
+  insert_head()
+  
+# reading the bibliography -------
+  
+  insert_msg('Raeding the bibliography')
+  
+  coll_bib <- read_bib('./paper/markdown/coll_biblio.bib')
+
+# analysis report -----
+  
+  insert_msg('Rendering manuscript parts')
+  
+  render('./paper/markdown/manuscript_figures_tables.Rmd', 
+         output_format = word_document2(number_sections = FALSE, 
+                                        reference_docx = 'ms_template.docx'), 
+         output_dir = './paper')
+  
+  render('./paper/markdown/manuscript_supplement.Rmd', 
+         output_format = word_document2(number_sections = FALSE, 
+                                        reference_docx = 'ms_template.docx'), 
+         output_dir = './paper')
+  
+# END -----
+  
+  insert_tail()

### 2025-02-21 paper scripts/supplementary_figures.R

@@ -84,21 +84,14 @@
   insert_msg('GBM, the cohort confounder')
   
   suppl_fig$gbm_cohort <- 
-    plot_grid(surv_cohort$forest_plot, 
-              surv_cohort$stat_plot + 
-                 theme(legend.position = 'right') + 
-                 labs(subtitle = 'model performance'), 
-              nrow = 2, 
-              rel_heights = c(1.3, 1), 
-              labels = LETTERS, 
-              label_size = 10) %>% 
+    surv_cohort$forest_plot %>% 
     as_figure(label = 'bcr_survival_gbm_model_cohort_confounder_pooled_geo', 
               ref_name = 'gbm_cohort', 
               caption = paste('Investigation of confounding effect of the', 
                               'cohort on prediction of BCR-free survival', 
                               'by the GBM model in the pooled GEO cohort.'), 
               w = 140, 
-              h = 180)
+              h = 110)
   
 # GBM models: clinical and combined ones --------
   
@@ -182,7 +175,7 @@
               ref_name = 'cl_plot', 
               caption = paste('GBM modeling of BCR-free survival in PCA with', 
                               'clinical prediction and expression of', 
-                              'collagen-related tramscripts:', 
+                              'collagen-related transcripts:', 
                               'survival in tertiles of predictor scores.'), 
               w = 180, 
               h = 220)

### 2025-02-21 tools/globals.R

@@ -196,7 +196,7 @@
       training = 'indianred3')
   
   globals$algo_labels <- 
-    c(ridge = 'Ridge Cox', 
+    c(ridge = 'RIDGE Cox', 
       elnet = 'Elastic Net Cox', 
       lasso = 'LASSO Cox', 
       svm = 'SVM', 
@@ -239,7 +239,7 @@
   globals$cohort_model_labels <- 
     c(cohort = 'cohort', 
       gbm_score = 'GBM predictor', 
-      full = 'GBM predictor + cohort')
+      full = 'GBM predictor + data set')
 
 # END -----
   

### 2025-02-21 univariable survival scripts/plots.R

@@ -214,6 +214,7 @@
                   x = 'BCR-free survival, months') + 
              globals$common_theme + 
              theme(plot.title = element_markdown(), 
+                   plot.title.position = 'plot', 
                    plot.tag = element_blank()) + 
              annotate('text',
                       label = v, 

### 2025-02-22 paper/markdown/manuscript_supplement.Rmd
Supplementary Material update, changes in modeling
Characteristics of the cohorts and modeling subsets included in Supplementary Material. Clearing issues on pre-operative PSA levels in the TCGA cohort. Inclusion of KLK3 - PSA-coding gene - as a surrogate of blood PSA in the clinical predictor set for  the GBM models.

@@
 
-set_flextable_defaults(font.family = 'Cambria', font.size = 10)
+set_flextable_defaults(font.family = 'Cambria', font.size = 9)
+
+```
+
+\newpage
+
+# Supplementary Table
+
+```{r tab-cohorts, tab.cap = "Characteristic of the analyzed cohorts.
+Numeric variables are presented as medians with interquartile ranges (IQR) and ranges.
+Qualitative variables are presented as percentages of categories within the complete observation set."}
+
+flextable::flextable(suppl_tab$cohorts) %>% 
+  set_widths(c(4, rep(3.75, 4))) %>% 
+  merge_v(1) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('PSA: prostate-specific antigen; pT stage: pathological tumor stage; pN stage: pathological lymph node stage; pM stage: pathological metastasis stage; ISUP grade: grading according to the International Society of Urological Pathology.'),
+           part = 'header', 
+           ref_symbols = 'a') %>%
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-pool, tab.cap = "Characteristic of GEO data sets which constitute the pooled GEO cohort.
+Numeric variables are presented as medians with interquartile ranges (IQR) and ranges.
+Qualitative variables are presented as percentages of categories within the complete observation set."}
+
+flextable::flextable(suppl_tab$pool) %>% 
+  set_widths(c(3.5, rep(3, 4), 2.3, 2.3)) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('PSA: prostate-specific antigen; pT stage: pathological tumor stage; pN stage: pathological lymph node stage; pM stage: pathological metastasis stage; ISUP grade: grading according to the International Society of Urological Pathology.'),
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 6:7, 
+            value = as_paragraph("Categorical variables: \u03C7\u00B2 test with Cramer's V effect size statistic.
+Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic.
+Biochemical relapse-free survival: Peto-Peto test.
+P values corrected for multiple testing with the false discovery rate method."),
+            part = 'header', 
+            ref_symbol = 'b') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-gse16560, tab.cap = "Characteristic of the training and test subsets of the GSE16560 cohort.
+For modeling of overall survival, the GSE16560 cohort was randomly split into a training and a test subset in a  2:1 ratio.
+Numeric characteristics of the subsets are presented as medians with interquartile ranges (IQR) and ranges.
+Qualitative variables are presented as percentages of categories within the complete observation set."}
+
+flextable::flextable(suppl_tab$gse16560) %>% 
+  set_widths(c(3.5, 3.5, 3.5, 2.5, 2.5)) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('ISUP grade: grading according to the International Society of Urological Pathology.'), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 4:5, 
+            value = as_paragraph("Categorical variables: \u03C7\u00B2 test with Cramer's V effect size statistic.
+Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic.
+Overall survival: Peto-Peto test.
+P values corrected for multiple testing with the false discovery rate method."),
+            part = 'header', 
+            ref_symbol = 'b')
 
@@
 
-_Co-expression networks of the collagen-related transcripts in PCA were constructed in the pooled GEO (cancer samples: n = `r nrow(net$data$geo_pool)`), TCGA (n = `r row(net$data$tcga)`), and DKFZ cohort (n = `r row(net$data$dkfz)`).
+_Co-expression networks of the collagen-related transcripts in PCA were constructed in the pooled GEO (cancer samples: n = `r nrow(net$data$geo_pool)`), TCGA (n = `r nrow(net$data$tcga)`), and DKFZ cohort (n = `r nrow(net$data$dkfz)`).
 The network edges were defined by pairwise correlations between ComBat-adjusted transcript levels in the cancer tissue with Spearman's $\rho \geq$ 0.5.
@@
 
-_Co-expression networks of the collagen-related transcripts were constructed in the pooled the pooled GEO (cancer samples: n = `r nrow(net$data$geo_pool)`), TCGA (n = `r row(net$data$tcga)`), and DKFZ cohort (n = `r row(net$data$dkfz)`) as presented in Supplementary Figure S\@ref(fig:fig-nets)._
+_Co-expression networks of the collagen-related transcripts were constructed in the pooled the pooled GEO (cancer samples: n = `r nrow(net$data$geo_pool)`), TCGA (n = `r nrow(net$data$tcga)`), and DKFZ cohort (n = `r nrow(net$data$dkfz)`) as presented in Supplementary Figure S\@ref(fig:fig-nets)._
 _The following numeric statistics of vertex importance were computed for each collagen-related transcript: degree as a measure of connectivity (number of correlation partners with Spearman's $\rho \geq$ 0.5), hub score as a measure of overall correlation strength (eigenvector of the affinity matrix), and betweenness as a measure of connectivity and centrality (minimum/maximum-scaled number of the shortest paths between all pairs of vertices that pass through the vertex of interest)._
@@
 
-_Prostate cancer (PCA) patients in the pooled GEO (total: n = `r uni_globals$n_numbers$n_total["geo_pool"]`, biochemical relapse [BCR]: n = `r uni_globals$n_numbers$n_events["geo_pool"]`), TCGA (total: n = `r uni_globals$n_numbers$n_total["tcga"]`, BCR: n = `r uni_globals$n_numbers$n_events["tcga"]`), and DKFZ cohort (total: n = `r uni_globals$n_numbers$n_total["dkfz"]`, BCR: n = `r uni_globals$n_numbers$n_events["dkfz"]`) were classified as high and low expressors for each of the `r `length(globals$genes)` collagen-related transcripts with the expression cutoffs corresponding to the largest differences in survival between the high and low expressors._
+_Prostate cancer (PCA) patients in the pooled GEO (total: n = `r uni_globals$n_numbers$n_total["geo_pool"]`, biochemical relapse [BCR]: n = `r uni_globals$n_numbers$n_events["geo_pool"]`), TCGA (total: n = `r uni_globals$n_numbers$n_total["tcga"]`, BCR: n = `r uni_globals$n_numbers$n_events["tcga"]`), and DKFZ cohort (total: n = `r uni_globals$n_numbers$n_total["dkfz"]`, BCR: n = `r uni_globals$n_numbers$n_events["dkfz"]`) were classified as high and low expressors for each of the `r length(globals$genes)` collagen-related transcripts with the expression cutoffs corresponding to the largest differences in survival between the high and low expressors._
 _Collagen-related transcripts associated with unfavorable and favorable BCR prognosis were identified by univariable Cox proportional hazard regression as presented in Figure 1._ 
@@
 
-_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and blood PSA concentration at diagnosis), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors.
+_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and tumor tissue expression of KLK3 transcript coding for PSA protein), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors.
 Performance of the GBM models was evaluated in the pooled GEO training collective (total: n = `r nrow(surv_clin$score_tbl$geo_pool)`, BCR: n = `r sum(surv_clin$score_tbl$geo_pool$relapse)`), and TCGA (total: n = `r nrow(surv_clin$score_tbl$tcga)`, BCR: n = `r sum(surv_clin$score_tbl$tcga$relapse)`) and DKFZ (total: n = `r nrow(surv_clin$score_tbl$dkfz)`, BCR: n = `r sum(surv_clin$score_tbl$dkfz$relapse)`) validation cohorts._
@@
 
-_(C) Importance of the explanatory variables for BCR-risk prediction by the GBM model with the clinical and transcriptional predictors.
+_(C) Importance of the explanatory variables for BCR risk prediction by the GBM model with the clinical and transcriptional predictors.
 The variable importance was computed as difference in sum of squared error ($\Delta SSE$) attributed to inclusion of a particular explanatory factor in the GBM learner ensemble.
@@
 
-_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and blood PSA concentration at diagnosis), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors, as presented in Supplementary Figure S\@ref(fig:fig-cl-plot)._
+_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and expressin of KLK3 transcript coding for PSA protein), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors, as presented in Supplementary Figure S\@ref(fig:fig-cl-plot)._
 _BCR-free survival in tertiles of the predictor scores of the clinical-only and the combined clinical - expression GBM models was visualized with Kaplan-Meier plots.

### 2025-02-22 BCR scripts/clinic.R

@@ -1,7 +1,7 @@
 # Survival gradient boosted machines with clinical predictors of 
 # BCR risk. 
-# Explanatory variables: Gleason (ISUP), log(PSA) at diagnosis and 
-# pT stage (3+ and 2 versus 1).
+# Explanatory variables: Gleason (ISUP), pT stage (3+ and 2 versus 1), and 
+# expression  of the PSA-expressing gene KLK3 in the tumor tissue. 
 # Response: BCR-free survival, with the time variable subjected to min/max 
 # scaling. 
 # Tuning andf training in the pooled GEO cohort. 

### 2025-02-22 BCR scripts/globals.R

@@ -2,8 +2,14 @@
 # in th pooled GEO, TCGA, and DKFZ cohorts. 
 #
 # BCR-free survival (+1 months to avoid zero survival times) is min/max scaled.
-# Clinical information (Gleadson/ISUP, pT stage, and log PCA) is extracted 
-# as well
+# Clinical information (Gleason/ISUP, pT stage) is extracted 
+# as well and will be used for construction of a clinical model. 
+# Unfortunately, information on pre-operative PSA in the TCGA cohort is 
+# provided only for a small minority of patients (n = 185 with 19 BCR cases), 
+# which excludes it as a predictor of BCR-free survival. 
+# Instead we include ComBat-adjusted log2 expression of PSA (KLK3 gene) in the 
+# clinical predictor set (references: DOI 10.3325/cmj.2020.61.450 and 
+# DO 10.1038/s41391-020-00283-3)
 
   insert_head()
   
@@ -117,19 +123,32 @@
   
   surv_globals$clinic <- 
     surv_globals$clinic[surv_globals$cohorts] %>%
-    map(select, sample_id, gleason_simple, pt_stage, psa_diagnosis) %>% 
+    map(select, sample_id, gleason_simple, pt_stage) %>% 
     map(~filter(.x, complete.cases(.x)))
   
   ## some minimal wrangling: 
-  ## pooling of pT stages,  
-  ## common units and log conversion of PSA
+  ## pooling of pT stages
 
   surv_globals$clinic <- surv_globals$clinic %>% 
     map(mutate, 
-        psa_diagnosis = log(psa_diagnosis + 1), 
+        #psa_diagnosis = log(psa_diagnosis + 1), 
+        #psa_diagnosis = zScores(psa_diagnosis), 
         pt_stage = car::recode(pt_stage, 
                               "'T3' = 'T3+'; 'T4' = 'T4+'"), 
         pt_stage = factor(pt_stage, c('T1', 'T2', 'T3+')))
+  
+  ## merging with the KLK3 expression levels
+
+  surv_globals$psa <- list(geo_pool = geo_pool, 
+                           tcga = tcga, 
+                           dkfz = dkfz) %>% 
+    map(~.x$combat[c('sample_id', 'KLK3')]) %>% 
+    map(mutate, KLK3 = zScores(KLK3))
+  
+  surv_globals$clinic <- 
+    map2(surv_globals$clinic, 
+         surv_globals$psa, 
[…]

### 2025-02-22 BCR scripts/plots.R

@@ -229,16 +229,19 @@
                          5), 
                      rep(list(c(positive = 'steelblue', 
                                 negative = 'steelblue', 
-                                ns = 'gray70')), 
+                                ns = 'gray70', 
+                                clinical = 'aquamarine3', 
+                                `collagen-related\ntranscript` = 'steelblue')), 
                          3))) %>% 
     pmap(plot_surv_importance, 
          form = 'bar', 
-         flip = TRUE) %>% 
+         flip = TRUE, 
+         split_regulation = FALSE) %>% 
     map(~.x + 
           theme(plot.subtitle = element_blank()) + 
           geom_vline(xintercept = 0,
                      linetype = 'dashed') + 
-          theme(legend.position = 'none'))
+          theme(legend.position = 'bottom'))
   
 # Error structure: plots of square errors of prediction -------
   

### 2025-02-22 OS scripts/globals.R

@@ -53,13 +53,7 @@
 
   ## data partition: training and test subset
   
-  set.seed(12345678)
-  
-  os_globals$train_idx <- 
-    createDataPartition(y = factor(os_globals$expression$death), 
-                        times = 1, 
-                        p = 2/3) %>% 
-    unlist
+  os_globals$train_idx <- gse16560$train_idx
   
   os_globals$data <- 
     list(training = os_globals$train_idx, 

### 2025-02-22 exploration scripts/cohorts.R

@@ -50,12 +50,32 @@
                   what = 'table', 
                   pub_styled = TRUE)) %>% 
     format_desc() %>% 
-    set_names(c('variable', 
+    set_names(c('Variable', 
                 globals$study_labels[names(cohorts$data)])) %>% 
     format_summ_tbl(rm_n = FALSE) %>% 
-    mutate(variable = factor(variable, cohorts$lexicon$variable)) %>% 
-    arrange(variable) %>% 
-    mutate(variable = exchange(variable, cohorts$lexicon))
+    mutate(Variable = factor(Variable, cohorts$lexicon$variable)) %>% 
+    arrange(Variable) %>% 
+    mutate(Variable = exchange(as.character(Variable), 
+                               cohorts$lexicon))
+  
+# appending the stat table with the total N numbers ogf cancer samples ------
+  
+  insert_msg('N numbers')
+  
+  cohorts$n_nunbers <- cohorts$data %>% 
+    map_dbl(nrow) %>% 
+    compress(names_to = 'cohort', 
+             values_to = 'n') %>% 
+    mutate(cohort = globals$study_labels[cohort]) %>% 
+    column_to_rownames('cohort') %>% 
+    t %>% 
+    as_tibble %>% 
+    mutate(Variable = 'Cancer samples, N') %>% 
+    relocate(Variable)
+  
+  cohorts$stats <- 
+    full_rbind(cohorts$n_nunbers, 
+               cohorts$stats)
   
 # END -----
   

### 2025-02-22 exploration scripts/gse16560.R

@@ -1,0 +1,139 @@
+# Comparison of the training and test portion of the GS16560 cohort: clinical 
+# factors and overall survival. Only the tabular form: this table will be shown 
+# in the revised manuscript. 
+#
+# Statistical significance: factors - chi-square test  with Cramer's V effect 
+# size statistic, numerics: Mann-Whitney test with r effect size statistic, 
+# survival: Peto-Peto test.
+
+  insert_head()
+  
+# container -------
+  
+  expl_os <- list()
+  
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  ## variable lexicon
+  
+  expl_os$lexicon <- globals$clinical_lexicon %>% 
+    filter(variable %in% c('age', 
+                           'gleason_sum', 
+                           'gleason_simple', 
+                           'death', 
+                           'os_months')) %>% 
+    mutate(variable = ifelse(variable == 'death', 
+                             'death_factor', variable), 
+           format = ifelse(variable == 'os_months', 'numeric', format), 
+           test_type = ifelse(format == 'numeric', 
+                              'kruskal_etasq', 'wilcoxon_r'), 
+           plot_type = ifelse(format == 'numeric', 
+                              'box', 'stack'), 
+           axis_lab = ifelse(format == 'factor', 
+                             '% of cohort', 
+                             ifelse(variable == 'os_months', 
+                                    'months', 'ng/mL')))
+  
+  ## analysis data
+  
+  expl_os$data <- gse16560$clinic %>% 
+    filter(tissue_type == 'tumor') %>% 
+    mutate(death_factor = car::recode(death, "0 = 'no'; 1 = 'yes'"), 
+           death_factor = factor(death_factor, c('no', 'yes')), 
+           observation = 1:nrow(.), 
+           subset = ifelse(observation %in% gse16560$train_idx, 
+                           'training', 'test'), 
+           subset = factor(subset, c('training', 'test'))) %>% 
+    select(sample_id, subset, death, all_of(expl_os$lexicon$variable))
+  
+# Descrptive stats -------
+  
+  insert_msg('Descriptive stats')
+  
+  expl_os$stats <- expl_os$data %>% 
+    explore(split_factor = 'subset', 
+            variables = expl_os$lexicon$variable, 
+            what = 'table', 
+            pub_styled = TRUE) %>% 
+    format_desc
+  
[…]

### 2025-02-22 exploration scripts/pooled_geo.R

@@ -32,15 +32,17 @@
   ## variable lexicon
   
   expl_pool$lexicon <- globals$clinical_lexicon %>% 
-    filter(variable %in% c('psa_diagnosis', 'pt_stage', 'pn_stage', 
-                           'gleason_simple', 'relapse', 'rfs_months')) %>% 
+    filter(variable %in% c('psa_diagnosis', 
+                           'ct_stage', 
+                           'pt_stage', 
+                           'pn_stage', 
+                           'gleason_sum', 
+                           'gleason_simple', 
+                           'relapse', 
+                           'rfs_months')) %>% 
     mutate(variable = ifelse(variable == 'relapse', 
                              'relapse_factor', variable), 
            format = ifelse(variable == 'rfs_months', 'numeric', format), 
-           label = ifelse(label == 'gleason_simple', 
-                          'ISUP', 
-                          ifelse(variable == 'rfs_months', 
-                                 'Follow-up, months', label)), 
            test_type = ifelse(format == 'numeric', 
                               'kruskal_etasq', 'cramer_v'), 
            plot_type = ifelse(format == 'numeric', 
@@ -62,11 +64,7 @@
     map(select, sample_id, relapse, any_of(expl_pool$lexicon$variable)) %>% 
     map(mutate, 
         relapse_factor = car::recode(relapse, "0 = 'no'; 1 = 'yes'"), 
-        relapse_factor = factor(relapse_factor, c('no', 'yes')), 
-        gleason_simple = car::recode(gleason_simple, 
-                                     "'5 - 6' = 'ISUP1'; 
-                                     '7' = 'ISUP2'; 
-                                     '8+' = 'ISUP3+'")) %>% 
+        relapse_factor = factor(relapse_factor, c('no', 'yes'))) %>% 
     compress(names_to = 'cohort') %>% 
     mutate(cohort = factor(cohort, 
                            c('gse54460', 'gse70768', 'gse70769', 'gse220095')))
@@ -92,16 +90,22 @@
   
   insert_msg('Comparison of the clinical features')
   
-  expl_pool$test <- expl_pool$data %>% 
-    compare_variables(variables = expl_pool$lexicon$variable, 
-                      split_factor = 'cohort', 
-                      what = 'eff_size', 
-                      types = expl_pool$lexicon$test_type, 
-                      exact = FALSE, 
[…]

### 2025-02-22 exploration scripts/psa.R

@@ -1,0 +1,170 @@
+# correlation analysis for the pre-operative PSA antigen in blood and 
+# expression levels of the PSA-coding gene KLK3 in the cancer tissue 
+# (ComBat-adjusted log2 expression). The analysis is performed for the 
+# pooled GEO cohort, TCGA, and the DKFZ collective, i.e. the data sets used 
+# later for BCR-free survival modeling.
+#
+# Statistical significance is assessed by Spearman's permutation test.
+
+  insert_head()
+  
+# container -------
+  
+  psa <- list()
+  
+# parallel backend -------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# analysis globals --------
+  
+  insert_msg('analysis globals')
+  
+  ## ComBat-adjusted KLK3 expression and PSA concentrations
+  
+  psa$data <- list(geo_pool = geo_pool, 
+                   tcga = tcga, 
+                   dkfz = dkfz) %>% 
+    map(~.x[c('clinic', 'combat')])
+  
+  for(i in names(psa$data)) {
+    
+    psa$data[[i]] <- 
+      inner_join(psa$data[[i]]$clinic[c('sample_id', 
+                                        'rfs_months', 
+                                        'relapse', 
+                                        'psa_diagnosis')], 
+                 psa$data[[i]]$combat[c('sample_id', 
+                                        'KLK3')], 
+                 by = 'sample_id')
+    
+  }
+  
+  ## data sets for the correlation and BCR-free survival 
+  ## analyses
+  
+  psa[c('cor_data', 'surv_data')] <- 
+    list(c('KLK3', 'psa_diagnosis'), 
+         c('rfs_months', 'relapse', 'KLK3')) %>% 
+    map(~map(psa$data, select, all_of(.x))) %>% 
+    map(map, ~filter(.x, complete.cases(.x)))
+  
+  ## minimal strata size for the survival analyis
+  
+  psa$min_n <-  map_dbl(psa$surv_data, nrow) * 0.25
+  
+# Correlation analysis -----
+  
+  insert_msg('Correlation analysis')
+  
+  psa$cor_test <- psa$cor_data %>% 
+    map(~f_cor_test(.x$KLK3, 
+                    .x$psa_diagnosis, 
+                    type = 'permutation', 
+                    method = 'spearman', 
+                    as_data_frame = TRUE, 
+                    n_iter = 1000)) %>% 
+    compress(names_to = 'cohort') %>% 
+    as_tibble
+  
[…]

### 2025-02-22 exploration.R

@@ -1,5 +1,5 @@
 # Exploratory analysis of expression of collagen-related genes shared by all 
-# transcriptomic datasets
+# transcriptomic datas ets
 #
 # 1) Characteristic of the cohorts and samples
 #
@@ -31,6 +31,7 @@
 
   library(survival)
   library(survminer)
+  library(kmOptimizer)
   
   library(ggrepel)
   library(ggtext)
@@ -55,7 +56,7 @@
   insert_msg('Exploration scripts')
   
   ## general distribution stats
-  ## coexpression analysis and co-expression networks
+  ## co-expression analysis and co-expression networks
   
   c('./exploration scripts/cohorts.R', 
     './exploration scripts/distribution.R', 
@@ -63,11 +64,16 @@
     './exploration scripts/networks.R') %>% 
     source_all(message = TRUE, crash = TRUE)
   
-  ## characteristic of the pooled GEO cohort
+  ## characteristic of the pooled GEO cohort; 
+  ## characteristic of the training and test portions of the GSE16560 cohort; 
+  ## correlation of KLK3 with blood PSA levels, 
+  ## and association of KLK3 with BCR-free survival
   
-  c('./exploration scripts/pooled_geo.R') %>% 
+  c('./exploration scripts/pooled_geo.R', 
+    './exploration scripts/gse16560.R', 
+    './exploration scripts/psa.R') %>% 
     source_all(message = TRUE, crash = TRUE)
-  
+
 # END -------
   
   insert_tail()

### 2025-02-22 import scripts/GSE16560.R

@@ -1,4 +1,8 @@
-# Import and wrangling of the GSE16560 data set
+# Import and wrangling of the GSE16560 data set. 
+# This cohort will be used for modeling of overall survival with the collagen-
+# related gene expression. 
+# To this end, we're defining the training/test partition of the cohort in a 2:1 
+# ratio
 
   insert_head()
   
@@ -68,11 +72,23 @@
     filter(!duplicated(gene_symbol), 
            !duplicated(entrez_id))
   
+# Indexes of the training subset of the GSE16560 cohort -------
+  
+  insert_msg('Indexes for the training subset of the cohort')
+  
+  set.seed(12345678)
+  
+  gse16560$train_idx <- 
+    createDataPartition(y = factor(gse16560$clinic$death), 
+                        times = 1, 
+                        p = 2/3) %>% 
+    unlist
+  
 # Caching the results ------
   
   insert_msg('Caching the results')
   
-  gse16560 <- gse16560[c("clinic", "annotation", "expression")]
+  gse16560 <- gse16560[c("clinic", "annotation", "expression", "train_idx")]
   
   save(gse16560, file = './data/gse16560.RData')
   

### 2025-02-22 import scripts/TCGA.R

@@ -1,4 +1,8 @@
-# Import and wrangling of the TCGA data set (version Cell 2018, cBioportal)
+# Import and wrangling of the TCGA data set (version Cell 2018, cBioportal). 
+# Some clinical information is extracted from the GDC repository 
+# (Gleason scores and the most recent PSA). 
+# PSA concentration at diagnosis is extracted from the seminal paper of the 
+# TCGA consortium (DOI: 10.1016/j.cell.2015.10.025) published in cell 2015.
 
   insert_head()
   
@@ -56,14 +60,22 @@
   
   tcga$clinic$gdc <- tcga$clinic$gdc %>% 
     transmute(patient_id = bcr_patient_barcode, 
-              ## PSA * 10 for the common unit with other data sets 
-              psa_diagnosis = as.numeric(psa) * 10, 
+              psa_recent = as.numeric(psa), 
               gleason_sum = as.numeric(gleason_score), 
               gleason_major = factor(as.numeric(primary_gleason)), 
               gleason_minor = factor(as.numeric(secondary_gleason)), 
               gleason_simple = cut(gleason_sum, 
                                    c(-Inf, 6, 7, Inf), 
                                    c('ISUP1', 'ISUP2', 'ISUP3+')))
+  
+  ## patient's PSA at diagnosis extracted from the seminal paper 
+  ## of the TCGA consortium
+  
+  tcga$clinic$psa <- read_xls('./data/TCGA/mmc2_seminal_tcga.xls')
+  
+  tcga$clinic$psa <- tcga$clinic$psa %>% 
+    transmute(patient_id = PATIENT_ID, 
+              psa_diagnosis = as.numeric(PSA_preop))
   
   ## sample
   
@@ -87,7 +99,7 @@
   
   ## merging
   
-  tcga$clinic <- tcga$clinic[c("patient", "gdc", "sample")] %>% 
+  tcga$clinic <- tcga$clinic[c("patient", "gdc", "sample", "psa")] %>% 
     reduce(left_join, by = 'patient_id') %>% 
     full_rbind(., tcga$clinic$normals)
   
@@ -108,7 +120,7 @@
     filter(!is.na(entrez_id)) %>% 
     annotate_raw_symbol
   
-  ## expression: log2 transformation and aggregatin of duplicated 
+  ## expression: log2 transformation and aggregating of duplicated 
   ## genes by arithmetic mean
   
   tcga$expression <- tcga$expression %>% 

### 2025-02-22 paper scripts/legacy_tables.R

@@ -1,0 +1,626 @@
+# Supplementary tables for the manuscript. 
+# As discussed with the study team, we're not showing the GSE16560 cohort, as 
+# it does not fulfill the strict selection criteria for the studies (biochemical 
+# relapse and relapse-free survival)
+
+  insert_head()
+  
+# containers ------
+  
+  suppl_paper_tbl <- list()
+  
+# globals -------
+  
+  insert_msg('Globals')
+  
+  ## metabolic reaction lexicon
+
+  suppl_paper_tbl$reactions <- ana_meta$models$tcga %>% 
+    components('regulation') %>% 
+    select(react_id) %>% 
+    mutate(react_name = annotate_bigg(react_id, annotation_db = Recon2D), 
+           react_name = unlist(react_name))
+  
+# Table S1: cohort characteristic --------
+  
+  insert_msg('Table S1: cohort characteristic')
+  
+  suppl_paper_tbl$cohorts <- cohorts$stats %>% 
+    select(-GSE16560) %>% 
+    set_names(c('Variable', 
+                globals$study_labels[names(globals$study_labels) != 'gse16560'])) %>% 
+    mdtable(label = 'table_S1_study_cohorts', 
+            ref_name = 'cohorts', 
+            caption = paste('Characteristic of the analyzed cohorts.', 
+                            'Numeric variables are presented as medians with', 
+                            'interquartile ranges (IQR) and ranges.', 
+                            'Qualitative variables are presented as', 
+                            'percentages of categories within the', 
+                            'complete observation set.'))
+  
+# Table S2: collagen genes of interest -----
+  
+  insert_msg('Table S2: collagen genes of interest')
+  
+  suppl_paper_tbl$genes <- globals$genes_interest %>% 
+    mutate(entrez_id = mapIds(org.Hs.eg.db, 
+                              keys = gene_symbol, 
+                              column = 'ENTREZID', 
+                              keytype = 'SYMBOL'),
+           gene_group = car::recode(gene_group, "'Pro' = 'proline turnover'"), 
+           gene_group = factor(gene_group, 
+                               c('proline turnover', 
+                                 'collagen modification', 
+                                 'ECM component', 
+                                 'ECM processing', 
+                                 'adhesion'))) %>% 
+    arrange(gene_group) %>% 
+    select(gene_group, gene_symbol, entrez_id) %>% 
+    set_names(c('Gene group', 'Gene symbol', 'Entrez ID')) %>% 
[…]

### 2025-02-22 paper scripts/links.R

@@ -1,4 +1,4 @@
-# Handling of common links in the Rmarkdown project
+# Handling of common links and biobiography in the Rmarkdown project
 
   insert_head()
 
@@ -15,12 +15,11 @@
          'trafo' = 'https://github.com/PiotrTymoszuk/trafo',
          'figur' = 'https://github.com/PiotrTymoszuk/figur',
          'clustTools' = 'https://github.com/PiotrTymoszuk/clustTools',
-         'caretExtra' = 'https://github.com/PiotrTymoszuk/caretExtra',
          'microViz' = 'https://github.com/PiotrTymoszuk/microViz', 
-         'biggrExtra' = 'https://github.com/PiotrTymoszuk/biggrExtra', 
          'coxExtensions' = 'https://github.com/PiotrTymoszuk/coxExtensions', 
          'kmOptimizer' = 'https://github.com/PiotrTymoszuk/kmOptimizer', 
-         'gseaTools' = 'https://github.com/PiotrTymoszuk/gseaTools') %>%
+         'fastTest' = 'https://github.com/PiotrTymoszuk/fastTest', 
+         'graphExtra' = 'https://github.com/PiotrTymoszuk/graphExtra') %>%
     compress(names_to = 'obj_name',
              values_to = 'x') %>%
     mutate(ref_name = paste0('_', obj_name, '_'))
@@ -28,6 +27,12 @@
   proj_links <- proj_links[c('ref_name', 'x')] %>%
     pmap(mdlink) %>%
     set_names(proj_links$obj_name)
+  
+# reading the bibliography -------
+  
+  insert_msg('Rending the bibliography')
+  
+  coll_bib <- read_bib('./paper/markdown/coll_biblio.bib')
 
 # END ------
 

### 2025-02-22 paper scripts/render.R

@@ -2,12 +2,6 @@
 
   insert_head()
   
-# reading the bibliography -------
-  
-  insert_msg('Raeding the bibliography')
-  
-  coll_bib <- read_bib('./paper/markdown/coll_biblio.bib')
-
 # analysis report -----
   
   insert_msg('Rendering manuscript parts')

### 2025-02-22 paper scripts/supplementary_figures.R

@@ -126,7 +126,8 @@
     guides(x = guide_axis(angle = 45)) + 
     theme(strip.background.y = element_blank(), 
           strip.text.y = element_blank(),
-          axis.text.x = element_markdown(size = 6.5)) 
+          axis.text.x = element_markdown(size = 6.5))  + 
+    labs(fill = 'explanatory\nfactor')
   
   ## the entire figure
   
@@ -228,7 +229,8 @@
           labs(title = paste0(.x$labels$title, 
                               globals$algo_labels[["ridge"]], 
                               sep = ', ')) + 
-          theme(legend.position = 'bottom', 
+          theme(plot.subtitle = element_blank(), 
+                legend.position = 'bottom', 
                 legend.title = element_blank(), 
                 legend.text = element_text(size = 7))) %>% 
     plot_grid(plotlist = .,

### 2025-02-22 paper scripts/supplementary_tables.R

@@ -1,0 +1,72 @@
+# Supplementary tables for the revised manuscript 
+
+  insert_head()
+  
+# container ------
+  
+  suppl_tab <- list()
+  
+# Characteristic of the cohorts: pooled GEO, TCGA, DKFZ, and GSE1650 ----
+  
+  insert_msg('Characteristic of the cohorts')
+  
+  suppl_tab$cohorts <- cohorts$stats %>% 
+    select(Variable, `pooled GEO`, TCGA, DKFZ, GSE16560) %>% 
+    as_mdtable(label = 'cohorts', 
+               ref_name = 'cohorts', 
+               caption = paste('Characteristic of the analyzed cohorts.', 
+                               'Numeric variables are presented as medians with', 
+                               'interquartile ranges (IQR) and ranges.', 
+                               'Qualitative variables are presented as', 
+                               'percentages of categories within the', 
+                               'complete observation set.'))
+  
+# Characteristic of the pooled GEO cohort -------
+  
+  insert_msg('Characteristic of the pooled GEO cohort')
+  
+  suppl_tab$pool <- expl_pool$result_tbl %>% 
+    as_mdtable(label = 'pool', 
+               ref_name = 'pool', 
+               caption = paste('Characteristic of GEO data sets which constitute', 
+                               'the pooled GEO cohort.', 
+                               'Numeric variables are presented as medians with', 
+                               'interquartile ranges (IQR) and ranges.', 
+                               'Qualitative variables are presented as', 
+                               'percentages of categories within the', 
+                               'complete observation set.'))
+  
+# Characteristic of the training and test portions of the GSE1650 cohort ------
+  
+  insert_msg('Characteristic of the training and test portion of the cohort')
+  
+  suppl_tab$gse16560 <- expl_os$result_tbl %>% 
+    as_mdtable(label = 'gse16560', 
+               ref_name = 'gse16560', 
+               caption = paste('Characteristic of the training and test subsets', 
+                               'of the GSE16560 cohort.', 
+                               'For modeling of overall survival, the GSE16560', 
+                               'cohort was randomly split into a training and', 
+                               'a test subset in a  2:1 ratio.', 
+                               'Numeric characteristics of the subsets are', 
+                               'presented as medians with', 
+                               'interquartile ranges (IQR) and ranges.', 
+                               'Qualitative variables are presented as', 
+                               'percentages of categories within the', 
[…]

### 2025-02-22 paper.R

@@ -31,12 +31,21 @@
     './tools/functions.R') %>% 
     source_all(message = TRUE, crash = TRUE)
   
+# Links and bibliography --------
+  
+  insert_msg('Links and bibliography')
+  
+  c('./paper scripts/links.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+
 # Figures, Tables, and Supplementary Material -------
   
   insert_msg('Figures, Tables, Supplements')
   
   c('./paper scripts/figures.R', 
-    './paper scripts/supplementary_figures.R') %>% 
+    './paper scripts/supplementary_figures.R', 
+    './paper scripts/supplementary_tables.R', 
+    './paper scripts/render.R') %>% 
     source_all(message = TRUE, crash = TRUE)
   
 # END ------

### 2025-02-22 tools/functions.R

@@ -203,7 +203,8 @@
                                                negative = 'steelblue', 
                                                ns = 'gray70'), 
                                    gene_variables = globals$genes, 
-                                   flip = FALSE) {
+                                   flip = FALSE, 
+                                   split_regulation = TRUE) {
     
     ## a bar or bubble plot of variable importance stats
     
@@ -219,11 +220,25 @@
              regulation = factor(regulation, c('positive', 'negative')), 
              variable = ifelse(stri_detect(variable, regex = gene_regex), 
                                html_italic(variable), 
-                               exchange(variable, 
-                                        globals$clinical_lexicon)), 
+                               ifelse(variable == 'KLK3', 
+                                      html_italic(variable), 
+                                      exchange(variable, 
+                                               globals$clinical_lexicon))), 
              variable = stri_replace(variable, 
                                      fixed = '_sq', 
                                      replacement = '<sup>2</sup>'))
+    
+    ## handling the special case: clinical and expression values
+    
+    if(any(stri_detect(data$variable, fixed = 'stage'))) {
+      
+      data <- data %>% 
+        mutate(regulation = ifelse(stri_detect(variable, regex = gene_regex), 
+                                   'collagen-related\ntranscript', 'clinical'), 
+               regulation = factor(regulation, 
+                                   c('collagen-related\ntranscript', 'clinical'))) 
+      
+    }
     
     if(!is.null(n_top)) {
       
@@ -234,7 +249,8 @@
     }
     
     data <- data %>% 
-      filter(!is.na(regulation))
+      filter(!is.na(regulation), 
+             .data[[imp_stat]] != 0)
     
     if(form == 'bar') {
       
@@ -285,11 +301,17 @@
       
     }
     
+    if(split_regulation) {
+      
+      imp_plot <- imp_plot + 
+        facet_grid(regulation ~ ., 
+                   scales = 'free', 
+                   space = 'free', 
+                   labeller = as_labeller(labeller))
+      
+    }
+    
     imp_plot <- imp_plot + 
-      facet_grid(regulation ~ ., 
-                 scales = 'free', 
-                 space = 'free', 
-                 labeller = as_labeller(labeller)) + 
       scale_fill_manual(values = palette) + 
       globals$common_theme + 
       labs(title = plot_title, 

### 2025-02-22 tools/globals.R

@@ -87,7 +87,7 @@
       'pn_stage' = 'pN stage', 
       'pm_stage' = 'pM stage', 
       'gleason_sum' = 'Gleason score', 
-      'gleason_simple' = 'ISUP', 
+      'gleason_simple' = 'ISUP grade', 
       'surgical_margins' = 'Surgical margins', 
       'ece' = 'Extracapsular extension', 
       'death' = 'Death', 

### 2025-02-22 paper/markdown/manuscript_supplement.Rmd
Genes of interest, network metrics
Update of 'Supplementary Tables

@@
             part = 'header', 
-            ref_symbol = 'b')
+            ref_symbol = 'b') %>% 
+  theme_vanilla
 
 ```
+
+\newpage
+
+```{r tab-genes, tab.cap = "Collagen-related genes and their classification."}
+
+flextable::flextable(suppl_tab$genes) %>% 
+  merge_v(1) %>% 
+  italic(j = 2, part = 'body') %>% 
+  set_widths(c(3.5, 3, 3)) %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-nets, tab.cap = "Vertex importance statistics for co-expression networks of the collagen-related transcripts.
+The co-expression networks were built for in the pooled GEO, TCGA, and DKFZ cohorts for transcripts with pairwise correlation of expression levels with Spearman's rho >= 0.5.
+Metrics of importance of the vertices of the co-expression networks, degree, hub score, betweenness, and transitivity, were computed.
+Top five vertices with the largest hub score per cohort are presented.
+The full table is available as a supplementary Excel file."}
+
+suppl_tab$nets %>% 
+  group_by(Cohort) %>% 
+  slice_max(`Hub score`, n = 5) %>% 
+  ungroup %>% 
+  flextable %>% 
+  merge_v(1) %>% 
+  italic(j = 2) %>% 
+  set_widths(c(2.3, 3, 3.5, rep(2.3, 4))) %>% 
+  footnote(1, 4, 
+           value = as_paragraph("Degree: number of network neighbors, i.e. correlations with Spearman's \u03C1 >= 0.5."), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 5, 
+           value = as_paragraph("Hub score: eigenvector of the affinity matrix; a metric of the overall correlation strength."), 
+           part = 'header', 
+           ref_symbols = 'b') %>% 
+  footnote(1, 6, 
+           value = as_paragraph('Betweenness: minimum/maximum scaled number of the shortest paths between pairs of network vertices which pass through the vertex of interest; measure of connectivity and centrality.'),
+           part = 'header', 
+           ref_symbols = 'c') %>% 
+  footnote(1, 7, 
+           value = as_paragraph('Transitivity: probability that the adjacent vertices of a vertex are connected; measure of the local densitiy of the co-expression network.'), 
+           part = 'header', 
+           ref_symbols = 'd') %>% 
+  theme_vanilla
+
+```
+
+\newpage
 

### 2025-02-22 paper scripts/supplementary_tables.R

@@ -59,6 +59,81 @@
   
   insert_msg('The collagen-related genes')
   
+  suppl_tab$genes <- globals$genes_lexicon %>% 
+    mutate(entrez_id = mapIds(org.Hs.eg.db, 
+                              keys = gene_symbol, 
+                              column = 'ENTREZID', 
+                              keytype = 'SYMBOL'),
+           gene_group = factor(gene_group, 
+                               c('proline metabolism', 
+                                 'collagen modification', 
+                                 'ECM component', 
+                                 'ECM processing', 
+                                 'adhesion'))) %>% 
+    arrange(gene_group) %>% 
+    select(gene_group, gene_symbol, entrez_id) %>% 
+    set_names(c('Functional classification', 'Gene symbol', 'Entrez ID')) %>% 
+    mdtable(label = 'genes', 
+            ref_name = 'genes', 
+            caption = 'Collagen-related genes and their classification.')
+  
+# Co-expression networks, vertex importance stats --------
+  
+  insert_msg('Co-expression networks, gene importance')
+  
+  ## top 5 genes with the largest degree per cohort are shown, 
+  ## the complete table as a supplementary file; 
+  ## betweenness is min/max scaled
+  
+  suppl_tab$nets <- net$graph_obj[c("geo_pool", "tcga", "dkfz")] %>% 
+    map(get_vertex_attributes) %>% 
+    map(select, 
+        name, gene_group, degree, hub_score, betweenness, transitivity) %>% 
+    map(mutate, 
+        betweenness = minMax(betweenness)) %>% 
+    compress(names_to = 'cohort') %>% 
+    relocate(cohort) %>% 
+    mutate(cohort = factor(cohort, c("geo_pool", "tcga", "dkfz"))) %>% 
+    arrange(cohort) %>% 
+    mutate(cohort = globals$study_labels[as.character(cohort)], 
+           betweenness = signif(betweenness, 2), 
+           hub_score = signif(hub_score, 2), 
+           transitivity = signif(transitivity, 2)) %>% 
+    set_names(c('Cohort', 
+                'Gene symbol', 'Functional classification', 
+                'Degree', 'Hub score', 'Betweenness', 'Transitivity'))
+  
+  suppl_tab$nets <- suppl_tab$nets %>% 
+    as_mdtable(label = 'nets', 
+               ref_name = 'nets', 
+               caption = paste('Vertex importance statistics for', 
+                               'co-expression networks of the collagen-related', 
+                               'transcripts.', 
[…]

### 2025-02-23 paper/markdown/manuscript_supplement.Rmd
Supplementary tables for survival modeling

@@
 flextable::flextable(suppl_tab$pool) %>% 
-  set_widths(c(3.5, rep(3, 4), 2.3, 2.3)) %>% 
+  set_widths(c(3.5, rep(3.3, 4), 2.2, 2.2)) %>% 
   footnote(1, 1, 
@@
   italic(j = 2) %>% 
-  set_widths(c(2.3, 3, 3.5, rep(2.3, 4))) %>% 
+  set_widths(c(2, 2.7, 3.5, rep(2.5, 4))) %>% 
   footnote(1, 4, 
@@
 \newpage
+
+```{r tab-uni-bcr, tab.cap = "Univariable analysis of biochemical relapse (BCR) free survival for the collagen-realted transcripts in the pooled GEO, TCGA, and DKFZ cohort.
+PCA patients were classified as high and low expressors of the collagen-related transcripts by expression cutoffs corresponding to the largest differences in BCR-free survival between the expression strata.
+BCR risk of high as compared with low expressors was modeled by univariable Cox proportional hazard regression.
+P values were corrected for multiple testing with the false discovery rate method.
+The modeling results are shown for the collagen-related genes associated with significant differences in BCR-free survival in all three investigated cohorts.
+The full analysis is available as a supplementary Excel file."}
+
+suppl_tab$uni_bcr %>% 
+  filter(`Gene symbol` %in% reduce(uni_cut$common_significant, union)) %>% 
+  arrange(`Association with BCR risk`, `Gene symbol`, Cohort) %>% 
+  relocate(`Association with BCR risk`, `Gene symbol`, Cohort) %>% 
+  flextable %>% 
+  merge_v(1:2) %>% 
+  italic(j = 2) %>% 
+  set_widths(c(2, 2, 2, 1.8, rep(2.4, 2), 2.2, 2.3, rep(1.7, 2))) %>% 
+  footnote(1, 7, 
+           value = as_paragraph('Harazd ratio (HR) with 95% confidence intervals (CI).'), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 9, 
+           value = as_paragraph("C-index: Harrel's concordance index, a measure of model accuracy.
+High values are expected for a highly accurate model; concordance index = 0.5 is expected for random predictions."),
+           part = 'header', 
+           ref_symbols = 'b') %>% 
+  footnote(1, 10, 
+           value = as_paragraph('IBS: Integrated Brier Score, a metric of overall model calibration.
+Low values are expected for a well calibrate model.
+Intergrated Brier score = 0.25 is expected for random predictions.'),
+           part = 'header', 
+           ref_symbols = 'c') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-tune, tab.cap = "Selection of the optimal parameters of machine learning models of biochemical relapse (BCR) free survival in the pooled GEO training cohort."}
+
+flextable::flextable(suppl_tab$tune) %>% 
+  set_widths(c(2.7, 6, 6)) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('SVM: Support Vector Machines; GBM: Gradient Boosted Machines.'), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1:6, 1, 
+           value = as_paragraph('Explanatory factors: expression levels of the collagen-related transcripts.'), 
+           part = 'body', 
+           ref_symbols = 'b') %>% 
+  footnote(7, 1, 
+           value = as_paragraph('Explanatory factors: pathological tumor stage, ISUP grade, and log2 expression of the PSA coding gene KLK3 in the tumor tissue.'), 
+           part = 'body', 
+           ref_symbols = 'c') %>% 
+  footnote(8, 1, 
+           value = as_paragraph('Explanatory factors: pathological tumor stage, ISUP grade, and log2 expression of the PSA coding gene KLK3 in the tumor tissue, log2 expression of the collagen-related transcripts.'),
+           part = 'body', 
+           ref_symbols = 'd') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-bcr-eval, tab.cap = "Performance of machine learning models at prediction of biochemical relapse (BCR) free survival."}
+
+flextable::flextable(suppl_tab$bcr_eval) %>% 
+  merge_v(1:2) %>% 
+  set_widths(c(3.5, 2.7, 3.5, rep(2.5, 2))) %>% 
+   footnote(1, 1, 
+           value = as_paragraph('SVM: Support Vector Machines; GBM: Gradient Boosted Machines.'), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1:18, 1, 
+           value = as_paragraph('Explanatory factors: expression levels of the collagen-related transcripts.'), 
+           part = 'body', 
+           ref_symbols = 'b') %>% 
+  footnote(19:21, 1, 
+           value = as_paragraph('Explanatory factors: pathological tumor stage, ISUP grade, and log2 expression of the PSA coding gene KLK3 in the tumor tissue.'), 
+           part = 'body', 
+           ref_symbols = 'c') %>% 
+   footnote(22:24, 1, 
+           value = as_paragraph('Explanatory factors: pathological tumor stage, ISUP grade, and log2 expression of the PSA coding gene KLK3 in the tumor tissue, log2 expression of the collagen-related transcripts.'),
+           part = 'body', 
+           ref_symbols = 'd') %>% 
+  footnote(1, 4, 
+           value = as_paragraph("C-index: Harrel's concordance index, a measure of model accuracy.
+High values are expected for a highly accurate model; concordance index = 0.5 is expected for random predictions."),
+           part = 'header', 
+           ref_symbols = 'b') %>% 
+  footnote(1, 5, 
+           value = as_paragraph('IBS: Integrated Brier Score, a metric of overall model calibration.
+Low values are expected for a well calibrate model.
+Intergrated Brier score = 0.25 is expected for random predictions.'),
+           part = 'header', 
+           ref_symbols = 'c') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-os-tune, tab.cap = "Selection of the optimal parameters of machine learning models of overall survival (OS) in the training subset of the GSE16560 cohort.
+The models used log2 expression of the collagen-related genes as explanatory variables."}
+
+flextable::flextable(suppl_tab$os_tune) %>% 
+  set_widths(c(2.7, 6, 6)) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('SVM: Support Vector Machines; GBM: Gradient Boosted Machines.'), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-os-eval, tab.cap = "Performance of machine learning models at prediction of overall survival in the training and test subsets of the GSE16560 cohort.
+The models used log2 expression of the collagen-related genes as explanatory factors."}
+
+flextable::flextable(suppl_tab$os_eval) %>% 
+  set_widths(c(3.5, 2.7, rep(2.5, 2))) %>% 
+  merge_v(1) %>% 
+  footnote(1, 3, 
+           value = as_paragraph("C-index: Harrel's concordance index, a measure of model accuracy.
+High values are expected for a highly accurate model; concordance index = 0.5 is expected for random predictions."),
+           part = 'header', 
+           ref_symbols = 'b') %>% 
+  footnote(1, 4, 
+           value = as_paragraph('IBS: Integrated Brier Score, a metric of overall model calibration.
+Low values are expected for a well calibrate model.
+Intergrated Brier score = 0.25 is expected for random predictions.'),
+           part = 'header', 
+           ref_symbols = 'c') %>% 
+  theme_vanilla
+
+```
 

### 2025-02-23 paper scripts/legacy_tables.R

@@ -492,7 +492,7 @@
                                    "'lambda' = '\u03BB'; 
                                    'type' = 'SVM model type'; 
                                    'gamma.mu' = '\u03B3'; 
-                                   'mtry' = 'number of variables per try, mtry'; 
+                                   'mtry' = 'number of variables per tree, mtry'; 
                                    'splitrule' = 'splitting rule'; 
                                    'nsplit' = 'number of splits'; 
                                    'nodesize' = 'minimal node size'; 

### 2025-02-23 paper scripts/supplementary_tables.R

@@ -10,7 +10,11 @@
   
   insert_msg('Characteristic of the cohorts')
   
+  ## Gleason scores are not shown any more, because they overlap 
+  ## with ISUP
+  
   suppl_tab$cohorts <- cohorts$stats %>% 
+    filter(!stri_detect(Variable, fixed = 'Gleason')) %>% 
     select(Variable, `pooled GEO`, TCGA, DKFZ, GSE16560) %>% 
     as_mdtable(label = 'cohorts', 
                ref_name = 'cohorts', 
@@ -25,7 +29,10 @@
   
   insert_msg('Characteristic of the pooled GEO cohort')
   
+  ## Gleason scores are not shown; redundant with ISUP
+  
   suppl_tab$pool <- expl_pool$result_tbl %>% 
+    filter(!stri_detect(Variable, fixed = 'Gleason')) %>% 
     as_mdtable(label = 'pool', 
                ref_name = 'pool', 
                caption = paste('Characteristic of GEO data sets which constitute', 
@@ -40,7 +47,10 @@
   
   insert_msg('Characteristic of the training and test portion of the cohort')
   
+  ## Gleason scores treated as above
+  
   suppl_tab$gse16560 <- expl_os$result_tbl %>% 
+    filter(!stri_detect(Variable, fixed = 'Gleason')) %>% 
     as_mdtable(label = 'gse16560', 
                ref_name = 'gse16560', 
                caption = paste('Characteristic of the training and test subsets', 
@@ -129,10 +139,228 @@
   ## the common markers are shown, the rest is available 
   ## as a supplementary Excel file
   
+  suppl_tab$uni_bcr <- uni_cut$modeling %>% 
+    compress(names_to = 'cohort') %>% 
+    transmute(Cohort = globals$study_labels[cohort], 
+              Cohort = factor(Cohort, 
+                              globals$study_labels[names(uni_cut$modeling)]), 
+              `Gene symbol` = gene_symbol, 
+              `Association with BCR risk` = marker, 
+              `log2 expression cutoff` = signif(cutoff, 2), 
+              `Low expressors, N` = paste0('total: n = ', n_total_low, 
+                             '\nBCR: n = ', n_events_low), 
+              `High expressors, N` = paste0('total: n = ', n_total_high, 
+                              '\nBCR: n = ', n_events_low), 
+              `HR, 95%CI` = paste0(signif(hr, 2), '\n[', 
[…]

### 2025-02-24 paper/markdown/manuscript_figures_tables.Rmd
Supplementary Methods
Supplementary Methods and a backbone for the main methods. Changes in labels and legends of Supplementary Figures

@@
 
+# Methods
+
+Analyses of collagen-related transcriptome and survival in transcriptomic cohorts were done with R version 4.2.3 (R foundation). 
+Their details are provided in __Supplementary Methods__.
+
+## Transcriptome data sets
+
+Seven publicly available transcriptome data sets of primary prostate cancer samples, 
+the TCGA prostate cancer cohort [@Liu2018; @Abeshouse2015] (n = `r tcga$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples, RNA sequencing), 
+GSE16560 [@Sboner2010] (n = `r gse16560$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray)
+GSE54460 [@Long2014] (n = `r gse54460$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), 
+GSE70768  [@Ross-Adams2015] (n = `r gse70768$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray), 
+GSE70769  [@Ross-Adams2015] (n = `r gse70769$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray), 
+GSE220095 [@Schimmelpfennig2023] (n = `r gse220095$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), and 
+DKFZ [@Gerhauser2018] (n = `r dkfz$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing) 
+were re-analyzed. 
+Selection criteria for the transcriptomic cohorts were: information of biochemical relapse and relapse-free survival or death and overall survival, and availability of expression data for `r nrow(globals$genes_interest)` collagen-related transcript investigated in our previous collagen project [@Heidegger2024].
+
+Normalized gene expression data and clinical information were fetched from the cBioportal repository (DKFZ, TCGA) or Gene Expression Omnibus (GEO). 
+The microarray signals of RNA sequencing counts were analyzed as $log_2$-transformed counts adjusted for differences in distribution between the cohorts with ComBat algorithm [@Leek2012]. 
+ComBat-adjusted expression and clinical information for the `r globals$study_labels[geo_pool$studies] %>% collapse_and` studies was merged into a 'pooled GEO cohort'. 
+In machine learning models of BCR-free survival, the pooled GEO cohort was used for training, and TCGA and DKFZ served as validation cohorts. 
+The GSE16560 cohort split into a training and test subset in a 2:1 ratio was used for modeling of overall survival. 
+Characteristic of the cohorts and a list of the collagen-related genes of  interest are are provided in __Supplementary Tables S1 - S4__.
+
+## Co-expression network analyses
+
+## Univariable modeling of BCR-free survival
+
+## Multi-parameter modeling of BCR-free and overall survival
+
+## Data and code availability
+
+Publicly available data sets were analyzed. 
+Formatted data sets used for analyses will be made available upon request to the corresponding author. 
+The R analysis pipeline for the transcriptome data is available from GitHub (https://github.com/PiotrTymoszuk/BCR_collagen).
+
+\newpage
+
 # Figures

### 2025-02-24 paper/markdown/manuscript_supplement.Rmd
Supplementary Methods
Supplementary Methods and a backbone for the main methods. Changes in labels and legends of Supplementary Figures

@@
 
-# Supplementary Table
+# Supplementary Methods for transcriptome analyses
+
+## Software
+
+The analysis was done with R version 4.2.3 (R Foundation).
+
+Tabular data were handled with the packages _tidyverse_ [@Wickham2019], _rlang_ [@Henry2022] and [_trafo_](https://github.com/PiotrTymoszuk/trafo). 
+Text data were handled with _stringi_ [@Gagolewski2021]. 
+For transcormation of numeric variables and variable distribution analyses, package [_microViz_](https://github.com/PiotrTymoszuk/microViz) was used.
+
+Import of the TCGA and DKFZ data sets from the cBioportal repository was accomplished with in-house-developed R scripts. 
+Transcriptome data sets from the Gene Expression Omnibus were fetched with the _GEOquery_ package [@Sean2007]. 
+Gene and probe annotation was accomplished with the _AnnotationDbi_ [@Pages2022] and _org.Hs.eg.db_ packages [@Carlson2022]. 
+ComBat adjustment of transcript expression levels for batch effects [@Leek2012] was accomplished with [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET) package. 
+Training/test cohort split was accomplished with _caret_ [@Kuhn2008].
+
+For descriptive statistics, statistical hypothesis testing, effect size calculation, and correlation analyses, packages [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) and [_fastTest_](https://github.com/PiotrTymoszuk/fastTest) were used.
+Co-expression network analyses were performed with packages _igraph_ [@Csardi2006] and [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra).
+
+Multi-parameter modeling of biochemical relapse (BCR) free survival and overall survival was done with the packages 
+_glmnet_ [@Friedman2010] and [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET) (Ridge, Elastic Net and LASSO cox regression), 
+_survivalsvm_ [@Fouodo2018] (Support Vector Machines [SVM]), 
+_rfsrc_ [@Ishwaran2008; @Ishwaran2022a] (Random Forest), 
+_gbm_ [@Greenwell2022] (Gradient Boosted Machines [GBM]), 
+_survival_ [@Therneau2000], 
+_surviminer_ [@Kassambara2016],  
+[_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions), and 
+_caret_ [@Kuhn2008]. 
+Univariable analysis of relapse-free survival was accomplished with packages [_kmOptimizer_](https://github.com/PiotrTymoszuk/kmOptimizer'), _survival_, _survminer_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions).
+
+For visualization of the results, the packages 
+_ggplot2_ [@Wickham2019] (scatter, dot, bar and Forest plots), 
+_survminer_, [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions), 
+[_kmOptimizer_](https://github.com/PiotrTymoszuk/kmOptimizer') (Kaplan-Meier plots, plots of Brier scores), and 
+[_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra) (network plots) were used. 
+Figures and tables were created with the packages _cowplot_ [@Wilke2019] and _flextable_ [@Gohel2022], respectively.
+
+Parts of the manuscript and Supplementary Material were written in _rmarkdown_ environment [@Allaire2022] with _bookdown_ package [@Xie2016], and rendered as Word files with _knitr_ [@Xie2022]. 
+Figure and table objects in the _rmarkdown_ documents were managed with [_figur_](https://github.com/PiotrTymoszuk/figur).
+
+## Data sources and data import
+
+Transcriptome cohorts were selected from studies deposited at cBioportal and Gene Expression Omnibus (GEO) with the following criteria: biochemical relapse and biochemical relapse-free survival or death and overall survival information, and expression data for `r nrow(globals$genes_interest)` collagen-related genes investigated in our previous collagen project [@Heidegger2024].
+
+The TCGA prostate cancer data set [@Liu2018; @Abeshouse2015] and DKFZ data set [@Gerhauser2018] consisted of normalized RNA sequencing data for `r tcga$expression %>% filter(tissue_type == 'tumor') %>% nrow` and `r dkfz$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples, respectively, with accompanying clinical information and were obtained from the cBioportal repository with in-house-developed R scripts.
+
+The GSE16560 [@Sboner2010] (microarray, Human 6k Transcriptionally Informative Gene Panel for DASL, n = `r gse16560$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples), 
+GSE54460 [@Long2014] (RNA sequencing, n = `r gse54460$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples), 
+GSE70768 [@Ross-Adams2015] (Illumina HumanHT-12 V4.0 expression beadchip microarray, n = `r gse70768$expression %>% filter(tissue_type == 'tumor') %>% nrow`), 
+GSE70769 [@Ross-Adams2015] (Illumina HumanHT-12 V4.0 expression beadchip microarray, n = `r gse70769$expression %>% filter(tissue_type == 'tumor') %>% nrow`), and 
+GSE220095 [@Schimmelpfennig2023] (RNA sequencing, n = `r gse220095$expression %>% filter(tissue_type == 'tumor') %>% nrow`) data sets with normalized whole-transcriptome and basic clinical information were fetched from GEO with the `getGEO()` function from the _GEOquery_ package.
+
+Whole transcriptome expression levels were transformed with the $log_2(expression)$ and $log_2(count + 1)$ function for, respectively, microarray and RNA sequencing data sets, and subjected to adjustment for batch cohort effects with the ComBat algorithm [@Leek2012] (function `multi_process()`, package [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET)).
+Subsequently, the ComBat-adjusted log2-transformed expression data and clinical information from `r globals$study_labels[geo_pool$studies] %>% collapse_and` data sets was merged into a 'pooled GEO cohort'.
+The pooled GEO cohort, TCGA, and DKFZ cohorts were used in co-expression network analyses, as well as uni- and multi-parameter modeling of BCR-free survival. 
+In the multi-parameter modeling of BCR-free survival, the pooled GEO cohort was used as the training collective, while the TCGA and DKFZ cohorts were used for model validation. 
+The GSE16560 cohort with all-case mortality information and survival times was used for multi-parameter modeling of overall survival. 
+The models of overall survival were developed in a training portion of the GSE16560 cohort with 2/3 of cancer samples, and validated in the remaining tumor specimens. 
+The training/test cohort split was done with function `createDataPartition()` from _caret_ package.
+
+Characteristics of the GSE16560, pooled GEO, TCGA, and DKFZ cohort are presented in __Supplementary Table S\@ref(tab:tab-cohorts)__. 
+Characteristics of the GEO data sets that constituted the pooled GEO cohort are listed in __Table \@ref(tab:tab-pool)__. 
+Characteristics of the training and test subsets of the GSE16560 cohort are summarized in __Table \@ref(tab:tab-gse16560)__. 
+The investigated collagen-related genes with their functional classifications are listed in __Table \@ref(tab:tab-genes)__. 
+## Descriptive statistics, statistical hypothesis testing, effect size and statistical significance
+
+Numeric values were referenced as medians with interquartile ranges and ranges. 
+Qualitative variables were presented as percentages and counts of their categories within the complete observation set. 
+Descriptive statistics were computed with function `explore()` from [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) package.
+
+Differences in numeric variables between two groups were investigated by Mann-Whitney test with r effect size statistic. 
+Differences in numeric variables between three or more groups were assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic. 
+Differences in distribution of categorical variables were assessed by $\chi^2$ test with Cramer's V effect size statistic. 
+Statistical hypothesis testing for differences between analysis groups were investigated with function `compare_variables()` from [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) package and the tool set of [_fastTest_](https://github.com/PiotrTymoszuk/fastTest) package.
+
+Effect sizes were interpreted as follows [@Cohen2013]:
+
+* r, weak: 0.1 - 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5
+
+* $\eta^2$: weak: 0.02 - 0.13, moderate: 0.13 - 0.26, large: $\geq$ 0.26
+
+* Cramer's V, weak: 0.1 - 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5
+
+P values were corrected for multiple testing with the false discovery rate method (FDR) [@Benjamini1995] within each analysis step and cohort. 
+Effects were considered statistically significant for FDR-corrected p values < 0.05.
+
+## Co-expression networks of the collagen-related genes
+
+Co-expression networks of the collagen-related transcripts were constructed as weighted undirected graphs [@Csardi2006] in the pooled GEO, TCGA, and DKFZ cohorts with `as_iGraph()` function from [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra) package.
+The network edges were defined by pairwise correlations between ComBat-adjusted $log_2$-transformed levels of the transcripts with Spearman's $\rho \geq$ 0.5. 
+Isolated vertices of the networks, i.e.
+transcripts lacking correlations with other transcripts with $\rho \geq$ 0.5, were removed (function `prune_degree()`, package [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra)).
+The following vertex importance statistics were computed for the co-expression networks with `summary()` method from [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra) package called for the graph objects:
+(1) degree as a measure of connectivity (number of network neighbors, i.e. correlations of the expression levels with $\rho \geq$ 0.5), 
+(2) hub score as a metric of overall correlation strength (eigenvector of the network's affinity matrix), 
+(3) betweenness as a measure of connectivity and centrality (minimum/maximum-scaled number of the shortest paths between the vertex pairs that pass through the vertex of interest), and 
+(4) transitivity as a measure of local network density (probability that the neighbor vertices of a vertex are connected with each other). 
+The vertex importance statistics are summarized in __Supplementary Table S\@ref(tab:tab-nets)__. 
+The co-expression networks were visualized with the Kamada-Kawai algorithm by calling `plot()` method for the co-expression graph objects (package [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra), modified implementation of _ggnetwork_) [@Briatte2021].
+
+## Univariable analysis of BCR-free survival
+
+The univariable analysis of BCR-free survival for the collagen-related genes was done in the pooled GEO, TCGA, and DKFZ cohorts. 
+For each of the collagen-related transcripts, cancer patients were classified as high and low expressors by cutoffs of ComBat-processed $log_2$-transformed expression levels which corresponded to the largest difference in biochemical relapse-free survival between the expression strata.
+The minimal size of the expression strata size was set to 25% of the observations; the cutoff search criterion was defined by minimal p value of Mentel-Henszel test. 
+The classification was done with `find_cutoff()` function from [_kmOptimizer_](https://github.com/PiotrTymoszuk/kmOptimizer)) package.
+
+BCR risk in the high expressors as compared with the low expressors for each of the collagen-related transcripts was modeled by univariable Cox proportional hazard regression (function `coxph()` and `as_coxex()`, packages _survival_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions)) [@Therneau2000].
+Hazard ratios (HR) for BCR risk in the high as compared with low expressors with 95% confidence intervals served as the inference statistics of the models. 
+The inference p values ($p(HR = 1)$), were corrected for multiple testing with the FDR method. 
+Accuracy of the model was assessed by 
+Harrell's concordance index [@Harrell1996] (high values are characteristic for high concordance between the predicted and observed BCR risk),  
+overall model calibration and confidence of predictions was gauged by Integrated Brier Score (low values are expected for well calibrated and highly confident models) [@Graf1999]. 
+The inference and model fit metrics were computed by calling `summary()` method from [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) for the model objects. 
+Genes found to be associated with unfavorable (pFDR < 0.05, HR > 1) or favorable BCR prognosis (pFDR < 0.05, HR < 1) in all three cohorts were considered as standalone markers of BCR prognosis listed in __Supplementary Table S\@ref(tab:tab-uni-bcr)__.
+
+## Multi-parameter modeling of BCR-free survival
+
+Multi-parameter modeling of BCR-free survival was done in the pooled GEO, TCGA, and DKFZ cohorts with six machine learning algorithms: 
+RIDGE regularized Cox proportional hazard regression [@Simon2011; @Friedman2010], 
+Elastic Net regularized Cox proportional hazard regression [@Simon2011; @Friedman2010; @Zou2005], 
+LASSO regularized Cox proportional hazard regression [@Simon2011; @Tibshirani1996; @Friedman2010], 
+Support Vector Machines (SVM) [@Fouodo2018; @Weston1998], 
+Random Forest [@Breiman2001; @Ishwaran2008; @Ishwaran2022a], and 
+Gradient Boosted Machines (GBM) [@Friedman2001; @Greenwell2022; @Natekin2013; @Friedman2002]. 
+Tuning, training, and predictions of the models were handled by in-house developed wrappers around the respective R implementations of the algorithms in packages _glmnet_, _survivalsvm_, _rfsrc_, and _gbm_.
+
+BCR-free survival was modeled with the following explanatory variables:
+
+1. ComBat-adjusted $log_2$-transformed expression levels of the collagen-related transcripts in the tumor tissue provided as first and second order terms (all algorithms)
+
+2.
+clinical predictors of BCR risk, pathological tumor stage, ISUP grade, and ComBat-adjusted $log_2$ expression levels of the PSA-coding transcript, _KLK3_ as a surrogate of blood PSA concentrations [@Lunger2021; @Boyukozer2020] (GBM algorithm)
+
+3.
+ComBat-adjusted $log_2$-transformed expression levels of the collagen-related transcripts (first and second order), pathological tumor stage, ISUP grade, and ComBat-adjusted $log_2$ expression levels of _KLK3_ (GBM algorithm)
+
+Numeric explanatory variables, i.e.
+first- and second-order ComBat-adjusted $log_2$ transcript levels were normalized by conversion to Z-scores prior to modeling (function `zScores()`, package [_microViz_](https://github.com/PiotrTymoszuk/microViz)).
+Because of differences in follow-up times between the training and validation cohorts, the BCR-free survival times were minimum/maximum scaled in each of the cohort (function `minMax()`, package [_microViz_](https://github.com/PiotrTymoszuk/microViz)).
+
+Selection (tuning) of the optimal set of parameters controlling the model behavior such as regularization, number of learners, or kernel function, was accomplished by minimizing model errors or maximizing concordance indexes for out-of-bag predictions or out-of-fold predictions in cross-validation in the pooled GEO cohort.
+The tuning criteria and the optimal values of the tunable parameters are summarized in __Supplementary Table S\@ref(tab:tab-tune)__. 
+The survival models were trained in the pooled GEO cohort for the optimal combinations of the tunable parameters.
+
+For the pooled GEO training cohort, and the TCGA and DKFZ validation collectives, predictor scores were returned by calling `predict()` method for the modeling objects. 
+Next, univariable Cox proportional hazard regression models were constructed with the BCR-free survival as the response and each of the predictor scores as the sole explanatory variable (functions `coxph()` and `as_coxex()`, packages _survival_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions)) [@Therneau2000].
+The proportional hazard assumption of the Cox proportional hazard models was checked with the `summary(type = 'assumptions')` method, which implements the genuine `cox.zph()` algorithm [@Grambsch1994].
+The Cox proportional hazard models were used for evaluation of machine learning accuracy with Harrell's concordance index [@Harrell1996] and assessment of global calibration with Integrated Brier Score [@Graf1999].
+Those performance statistics were computed by calling `summary(type = 'fit')` method from [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) package for the Cox proportional hazard modeling objects (__Supplementary Table S\@ref(tab:tab-bcr-eval)__).
+To assess calibration and confidence of BCR risk prediction of the machine learning models, we plotted Brier Scores defined as a mean square difference between the predicted BCR probability and the 0/1-coded observed BCR event [@Brier1950; @Graf1999] for unique survival time points.
+With this graphical method we evaluated predictions made by the best-performing GBM models (__Supplementary Figure S\@ref(fig:fig-clinic)B__). 
+In addition, ability of the machine learning models to assign PCA patients to low-, intermediate-, and high-risk groups was investigated by comparing BCR-free survival between patients stratified by tertiles of the predictor scores with Kaplan-Meier plots and Peto-Peto test (functions `calibrate.coxex()`, `plot.calibrator()`, and `surv_pvalue()`, packages [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) and _survminer_) as presented for the GBM models in __Figure 2B__ and __Supplementary Figure \@ref(fig:fig-cl-plot)__.
+
+Importance of the explanatory variables for the predictive performance of the GBM models in the training pooled GEO cohort were measured with the relative influence method [@Friedman2001] and expressed as gradient of the sum of squared errors attributed to particular variables in the learner ensemble ($\Delta SSE$, __Figure 2C__, __Supplementary Figure S\@ref(fig:fig-clinic)__).
+
+In order to assess the confounding effect of single studies that constituted the pooled GEO data set on results of modeling of BCR-free survival with the best performing GBM algorithm, three nested Cox proportional hazard models were constructed with the pooled GEO data:
+(1) a model with study as the sole explanatory variable, 
+(2) a model with normalized (Z-score) GBM linear predictor score as the sole explanatory variable, and 
+(3) an additive full mode with study and normalized GBM linear predictor score as the explanatory variables. 
+Hazard ratio estimates for the study and GBM predictor terms with 95% confidence intervals were extracted from the models with `summary()` method from the [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) package.
+Contribution of the data set and GBM predictor terms to the additive model were explored by likelihood ratio test (LRT, method `anova()`, package _stats_). 
+As presented in __Supplementary Figure S\@ref(fig:fig-gbm-cohort)__, adjustment for the study effect did not change the hazard ratio estimate of the GBM term in comparison of inference statistics of the GBM predictor-only and the additive Cox model.
+Additionally, the highly significant result of the LRT for the GBM predictor term and the non-significant LRT result for the study term indicate that the GBM predictor but not the study effect contributed substantially to the additive model.
+Collectively, these results let us conclude that prediction of BCR-free survival in the pooled GEO cohort is independent from differences in BCR-free survival between studies that constitute the pooled collective.
+
+## Multi-parameter modeling of overall survival
+
+Insufficient number of death cases in the TCGA cohort (n = `r tcga$clinic$death %>% sum(na.rm = TRUE)`) precluded any serious modeling of overall survival. 
+To overcome this limitation, we resorted to the GSE16560 watchful waiting study cohort with long observation times and reliable overall survival data for n = `r nrow(gse16560$clinic)` PCA patients with `r gse16560$clinic$death %>% sum` deceased individuals [@Sboner2010].
+This collective was randomly spit into a training and test subset in a 2:1 ratio, and the training portion was used for tuning and training of machine learning models of overall survival employing the algorithms described for the BCR-free survival modeling.
+The explanatory factors were first- and second-order terms of ComBat-adjusted $log_2$-transformed levels of the collagen-related transcript in the PCA tissue. 
+Those expression variables were converted to Z-scores prior to modeling. 
+The tuning and training process was analogical to the modeling of BCR-free survival (for the best combinations of tunable parameters, see __Supplementary Table S\@ref(tab:tab-os-tune)__). 
+The models were evaluated in the training and test subset of the GSE16560 cohort by computing concordance indexes and Integrated Brier Scores, visual assessment of Brier scores for unique time points, and comparison of overall survival between tertiles of predictor scores as described for the models of BCR-free survival.
+The model performance metrics are summarized in __Supplementary Table S\@ref(tab:tab-os-eval)__.
+
+## Data and code availability
+
+Publicly available data sets were analyzed. 
+Formatted data sets used for analyses will be made available upon request to the corresponding author. 
+The R analysis pipeline for the transcriptome data is available from GitHub (https://github.com/PiotrTymoszuk/BCR_collagen).
+
+\newpage
+
+# Supplementary Tables
 
@@
            value = as_paragraph('IBS: Integrated Brier Score, a metric of overall model calibration.
-Low values are expected for a well calibrate model.
+Low values are expected for a well calibrated model.
 Intergrated Brier score = 0.25 is expected for random predictions.'),
@@
 
-```{r tab-os-eval, tab.cap = "Performance of machine learning models at prediction of overall survival in the training and test subsets of the GSE16560 cohort.
+```{r tab-os-eval, tab.cap = "Performance of machine learning models at prediction of overall survival (OS) in the training and test subsets of the GSE16560 cohort.
 The models used log2 expression of the collagen-related genes as explanatory factors."}
@@
   footnote(1, 3, 
-           value = as_paragraph("C-index: Harrel's concordance index, a measure of model accuracy.
+           value = as_paragraph("C-index: Harrell's concordance index, a measure of model accuracy.
 High values are expected for a highly accurate model; concordance index = 0.5 is expected for random predictions."),
@@
 transcripts without correlation partners, were removed._
-_The co-expression networks were visualized with Fruchterman-Reingold algorithm.
+_The co-expression networks were visualized with Kamada-Kawai algorithm.
 Each point represents a single transcript in the network; point color codes for functional classification of the transcript.
@@
 
-_Co-expression networks of the collagen-related transcripts were constructed in the pooled the pooled GEO (cancer samples: n = `r nrow(net$data$geo_pool)`), TCGA (n = `r nrow(net$data$tcga)`), and DKFZ cohort (n = `r nrow(net$data$dkfz)`) as presented in Supplementary Figure S\@ref(fig:fig-nets)._
+_Co-expression networks of the collagen-related transcripts were constructed in the pooled GEO (cancer samples: n = `r nrow(net$data$geo_pool)`), TCGA (n = `r nrow(net$data$tcga)`), and DKFZ cohort (n = `r nrow(net$data$dkfz)`) as presented in Supplementary Figure S\@ref(fig:fig-nets)._
 _The following numeric statistics of vertex importance were computed for each collagen-related transcript: degree as a measure of connectivity (number of correlation partners with Spearman's $\rho \geq$ 0.5), hub score as a measure of overall correlation strength (eigenvector of the affinity matrix), and betweenness as a measure of connectivity and centrality (minimum/maximum-scaled number of the shortest paths between all pairs of vertices that pass through the vertex of interest)._
 _The vertex importance metrics for the collagen-related transcripts in the co-expression networks are shown in dot plots.
-Each point represents a single collage-related transcript, point color codes for functional classification of the transcript._
+Each point represents a single collagen-related transcript, point color codes for functional classification of the transcript._
 
@@
 _Collagen-related transcripts associated with unfavorable and favorable BCR prognosis were identified by univariable Cox proportional hazard regression as presented in Figure 1._ 
-_COL1A1, COL11A1, and COL11A2 were identified as the strongest unfavorable BCR risk markers, as measured by mean Harrell's concordance index (measure of model accuracy) in the investigated cohorts.
+_COL1A1, COL11A1, and COL11A2 were identified as the strongest unfavorable BCR risk markers, as metered by mean Harrell's concordance index (measure of model accuracy) in the investigated cohorts.
 Fractions of BCR-free patients in the high and low expression strata are visualized in Kaplan-Meier plots.
 The log2 expression cutoffs used for the high/low expressor classification, hazard-ratios (HR) of BCR risk in the high as compared with the low expression strata with 95% confidence intervals (95% CI) are displayed in the plot captions.
-P values (p(HR = 0)) corrected for multiple testing with the false discovery rate method are shown in the plots.
+P values corrected for multiple testing with the false discovery rate method are shown in the plots.
 Numbers of patients and BCR cases in the expression strata are indicated in the plot legends._
@@
 The log2 expression cutoffs used for the high/low expressor classification, hazard-ratios (HR) of BCR risk in the high as compared with the low expression strata with 95% confidence intervals (95% CI) are displayed in the plot captions.
-P values (p(HR = 0)) corrected for multiple testing with the false discovery rate method are shown in the plots.
+P values corrected for multiple testing with the false discovery rate method are shown in the plots.
 Numbers of patients and BCR cases in the expression strata are indicated in the plot legends._
@@
 
-```{r fig-gbm-cohort, fig.width = figur::convert(suppl_fig$gbm_cohort, to = 'in')$w, fig.height = figur::convert(suppl_fig$gbm_cohort, to = 'in')$h, fig.cap = 'Investigation of confounding effect of the cohort on prediction of BCR-free survival by the GBM model in the pooled GEO cohort.'}
+```{r fig-gbm-cohort, fig.width = figur::convert(suppl_fig$gbm_cohort, to = 'in')$w, fig.height = figur::convert(suppl_fig$gbm_cohort, to = 'in')$h, fig.cap = 'Investigation of the confounding study effect of the cohort on prediction of BCR-free survival by the GBM model in the pooled GEO cohort.'}
 
@@
 
-__Figure \@ref(fig:fig-gbm-cohort). Investigation of confounding effect of the cohort on prediction of BCR-free survival by the GBM model in the pooled GEO cohort.__
+__Figure \@ref(fig:fig-gbm-cohort). Investigation of confounding study effect on prediction of BCR-free survival by the GBM model in the pooled GEO cohort.__
 
@@
 The Gradient Boosted Machines (GBM) model was trained in the pooled GEO cohort (total: n = `r surv_globals$n_numbers$n_total[1]`, BCR: n = `r surv_globals$n_numbers$n_events[1]`)._
-_To investigate if and to which extend the BCR risk prediction depends on the confounding effect of single data sets constituting the pooled GEO cohort (`r globals$study_labels[geo_pool$studies] %>% collapse_and`), we constructed a canonical Cox proportional hazard regression model of BCR-free survival with the GBM predictor score and assignment to the data sets as independent variables.
-Performance and inference of this data set-adjusted GBM model was compared with performance and inference of the initial GBM model._
+_To investigate if and to which extend the BCR risk prediction depends on the confounding effect of single studies constituting the pooled GEO cohort (`r globals$study_labels[geo_pool$studies] %>% collapse_and`), we constructed a canonical Cox proportional hazard regression model of BCR-free survival with the GBM predictor score and assignment to the study as independent variables.
+Inference of this data set-adjusted GBM model was compared with inference of the initial GBM model._
 
-_Coefficient estimates expressed as hazard ratios (HR) with 95% confidence intervals of the data set-adjusted, the initial and the data set-only survival models are presented in a Forest plot.
-Significance of contribution of the cohort and GBM predictor score to the BCR-free survival prediction were assessed by likelihood ratio test (LRT), whose results are presented in the plot facets.
-Note the significant contribution of the GBM predictor score to the prediction of BCR-free survival and the non-significant effect of the data set._
+_Coefficient estimates expressed as hazard ratios (HR) with 95% confidence intervals of the study-adjusted, the initial and the study-only survival models are presented in a Forest plot.
+Significance of contribution of the study and GBM predictor score to the BCR-free survival prediction were assessed by likelihood ratio test (LRT), whose results are presented in the plot facets.
+Note the significant contribution of the GBM predictor score to the prediction of BCR-free survival and the non-significant effect of the study._
 
@@
 
-_(A) Numeric metrics of survival model performance, Harrell's concordance index (C-index, measured of model accuracy) and Integrated Brier Score (IBS, measure of overall calibration, low values are characteristic for well calibrated models), are presented in a dot plot.
+_(A) Numeric metrics of survival model performance, Harrell's concordance index (C-index, measure of model accuracy) and Integrated Brier Score (IBS, measure of overall calibration, low values are characteristic for well calibrated models), are presented in a dot plot.
 Each point represents a single GBM model, point color codes for the model type.
-Models fit in the same cohort are connected with lines.
+Models in the same cohort are connected with lines.
 Note the substantial improvement of the GBM model accuracy and calibration upon inclusion of the collagen-related transcript expression levels in the explanatory variable set in the pooled GEO and DKFZ cohorts._
@@
 
-_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and expressin of KLK3 transcript coding for PSA protein), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors, as presented in Supplementary Figure S\@ref(fig:fig-cl-plot)._
+_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and expression of KLK3 transcript coding for PSA protein), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors, as presented in Supplementary Figure S\@ref(fig:fig-cl-plot)._
 _BCR-free survival in tertiles of the predictor scores of the clinical-only and the combined clinical - expression GBM models was visualized with Kaplan-Meier plots.

### 2025-02-24 BCR scripts/cohort.R

@@ -156,7 +156,7 @@
                     color_variable = 'model_type', 
                     palette = unname(globals$cohort_model_colors), 
                     labels = globals$cohort_model_labels, 
-                    plot_title = 'Data set confounder, pooled GEO cohort', 
+                    plot_title = 'Study confounder, pooled GEO cohort', 
                     plot_subtitle = 'GBM survival modeling')
   
 # Forest plot of the HRs -------
@@ -208,7 +208,7 @@
     theme(axis.title.y = element_blank(), 
           strip.text.y = element_text(angle = 0, 
                                       hjust = 0)) + 
-    labs(title = 'Data set confounder, pooled GEO cohort', 
+    labs(title = 'Study confounder, pooled GEO cohort', 
          subtitle = 'Inference and likelihood ratio test', 
          x = 'HR, 95% CI')
   

### 2025-02-24 BCR scripts/globals.R

@@ -9,7 +9,7 @@
 # which excludes it as a predictor of BCR-free survival. 
 # Instead we include ComBat-adjusted log2 expression of PSA (KLK3 gene) in the 
 # clinical predictor set (references: DOI 10.3325/cmj.2020.61.450 and 
-# DO 10.1038/s41391-020-00283-3)
+# DOI 10.1038/s41391-020-00283-3)
 
   insert_head()
   

### 2025-02-24 exploration scripts/networks.R

@@ -145,7 +145,7 @@
            plot_title = paste(globals$study_labels[names(net$graph_obj)], 
                               i[[2]], sep = ', ')) %>% 
       pmap(plot, 
-           layout = layout.fruchterman.reingold, 
+           layout = layout.kamada.kawai, 
            vertex_fill_variable = i[[1]], 
            vertex_color_variable = NULL, 
            vertex_color = 'black', 

### 2025-02-24 paper scripts/links.R

@@ -19,7 +19,8 @@
          'coxExtensions' = 'https://github.com/PiotrTymoszuk/coxExtensions', 
          'kmOptimizer' = 'https://github.com/PiotrTymoszuk/kmOptimizer', 
          'fastTest' = 'https://github.com/PiotrTymoszuk/fastTest', 
-         'graphExtra' = 'https://github.com/PiotrTymoszuk/graphExtra') %>%
+         'graphExtra' = 'https://github.com/PiotrTymoszuk/graphExtra', 
+         'htGLMNET' = 'https://github.com/PiotrTymoszuk/htGLMNET') %>%
     compress(names_to = 'obj_name',
              values_to = 'x') %>%
     mutate(ref_name = paste0('_', obj_name, '_'))

### 2025-02-24 paper scripts/supplementary_figures.R

@@ -11,8 +11,11 @@
   insert_msg('Co-expression of the collagen-related genes')
   
   suppl_fig$nets <- 
-    net$graph_plots$gene_group[c("geo_pool", "tcga", "dkfz")] %>% 
-    map(~.x + theme(legend.position = 'none')) %>% 
+    map2(net$graph_plots$gene_group[c("geo_pool", "tcga", "dkfz")], 
+         c("geo_pool", "tcga", "dkfz"), 
+         ~.x + 
+           labs(title = globals$study_labels[.y]) + 
+           theme(legend.position = 'none')) %>% 
     c(list(get_legend(net$graph_plots$gene_group[[1]]))) %>% 
     plot_grid(plotlist = ., 
               ncol = 2, 
@@ -31,6 +34,7 @@
   
   suppl_fig$hubs <- net$stat_plots[c("geo_pool", "tcga", "dkfz")] %>% 
     map(~.x + theme(legend.position = 'none')) %>% 
+    c(list(get_legend(net$stat_plots$geo_pool))) %>% 
     plot_grid(plotlist = ., 
               ncol = 2, 
               align = 'hv', 
@@ -87,8 +91,8 @@
     surv_cohort$forest_plot %>% 
     as_figure(label = 'bcr_survival_gbm_model_cohort_confounder_pooled_geo', 
               ref_name = 'gbm_cohort', 
-              caption = paste('Investigation of confounding effect of the', 
-                              'cohort on prediction of BCR-free survival', 
+              caption = paste('Investigation of the confounding study', 
+                              'effect on prediction of BCR-free survival', 
                               'by the GBM model in the pooled GEO cohort.'), 
               w = 140, 
               h = 110)

### 2025-02-24 paper scripts/supplementary_tables.R

@@ -356,7 +356,7 @@
     as_mdtable(label = 'os_eval', 
                ref_name = 'os_eval', 
                caption = paste('Performance of machine learning models at', 
-                               'prediction of overall survival in the', 
+                               'prediction of overall survival (OS) in the', 
                                'training and test subsets of the GSE16560', 
                                'cohort.', 
                                'The models used log2 expression of the', 

### 2025-02-24 tools/globals.R

@@ -237,9 +237,9 @@
       full = 'firebrick4')
   
   globals$cohort_model_labels <- 
-    c(cohort = 'cohort', 
+    c(cohort = 'study', 
       gbm_score = 'GBM predictor', 
-      full = 'GBM predictor + data set')
+      full = 'GBM predictor + study')
 
 # END -----
   

### 2025-02-25 paper/markdown/manuscript_figures_tables.Rmd
Methods for the main manuscript
ROC analysis for BCR is still pernding

@@
 Seven publicly available transcriptome data sets of primary prostate cancer samples, 
-the TCGA prostate cancer cohort [@Liu2018; @Abeshouse2015] (n = `r tcga$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples, RNA sequencing), 
-GSE16560 [@Sboner2010] (n = `r gse16560$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray)
-GSE54460 [@Long2014] (n = `r gse54460$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), 
-GSE70768  [@Ross-Adams2015] (n = `r gse70768$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray), 
-GSE70769  [@Ross-Adams2015] (n = `r gse70769$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray), 
-GSE220095 [@Schimmelpfennig2023] (n = `r gse220095$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), and 
-DKFZ [@Gerhauser2018] (n = `r dkfz$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing) 
+the [TCGA](https://www.cbioportal.org/study/summary?id=prad_tcga_pan_can_atlas_2018) prostate cancer cohort [@Liu2018; @Abeshouse2015] (n = `r tcga$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples, RNA sequencing),
+[GSE16560](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE16560) [@Sboner2010] (n = `r gse16560$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray)
+[GSE54460](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE54460) [@Long2014] (n = `r gse54460$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), 
+[GSE70768](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE70768) [@Ross-Adams2015] (n = `r gse70768$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray), 
+[GSE70769](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE70769) [@Ross-Adams2015] (n = `r gse70769$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray), 
+[GSE220095](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE220095) [@Schimmelpfennig2023] (n = `r gse220095$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), and 
+[DKFZ](https://www.cbioportal.org/study/summary?id=prostate_dkfz_2018) [@Gerhauser2018] (n = `r dkfz$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing) 
 were re-analyzed. 
@@
 
+Co-expression networks of the collagen-related transcripts were constructed in the poled GEO, DKFZ, and TCGA cohorts. 
+The networks' edges were defined by pairwise correlations of the ComBat-adjusted $log_2$ expression levels with Spearman's $\rho \geq$ 0.5; transcripts without any network neighbors were removed. 
+Importance of the network's vertices, i.e. the collagen-related transcripts, was assessed by 
+degree as a measure of connectivity (number of neighbors), 
+hub score as a metric of overall correlation strength (eigenvector of the affinity matrix), 
+betweenness as a measure of connectivity and centrality (minimum/maximum-scaled number of the shortest paths between all pairs of vertices that pass through the vertex of interest), and 
+transitivity (fraction of neighbor vertices that are connected with each other) [@Csardi2006] (__Supplementary Table S5__).
+
 ## Univariable modeling of BCR-free survival
 
-## Multi-parameter modeling of BCR-free and overall survival
+For univariable analysis of BCR-free survival, PCA patients in the pooled GEO, TCGA, and DKFZ cohort were classified as high and low expressors of each collagen-related gene with a cutoffs corresponding to the largest difference in survival between the expression strata.
+BCR risk of the high as compared with low expressors was assessed by unavailable Cox proportional hazard regression. 
+Inference of the Cox proportional hazard models was assessed by hazard ratio (HR) with 95% confidence intervals, and p values corrected for multiple testing with the false discovery rate method (FDR).
+Harrell's concordance index [@Harrell1996] and Integrated Brier Score [@Graf1999] were used as statistics of, respectively, model accuracy and calibration (__Supplementary Table S6__).
+
+## Multi-parameter modeling of BCR-free survival
+
+Machine learning models of BCR-free survival were constructed for the following sets of explanatory variables: 
+(1) first- and second-order terms of ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, 
+(2) pathological tumor stage, ISUP grade, and ComBat-adjusted $log_2$ mRNA levels of the PSA-coding gene _KLK3_ [@Lunger2021; @Boyukozer2020] as clinical predictors of BCR, and 
+(3) pathological tumor stage, ISUP grade, ComBat-adjusted $log_2$ _KLK3_ mRNA levels, as well as the first- and second-order terms of ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts.
+The models were trained in the pooled GEO cohort with six algorithms (RIDGE regularized Cox proportional hazard regression, Elastic Net regularized Cox proportional hazard regression, LASSO regularized Cox proportional hazard regression, Support Vector Machines [SVM], Random Forest, and Gradient Boosted Machines [GBM]) [@Simon2011; @Tibshirani1996; @Zou2005; @Friedman2010; @Fouodo2018; @Weston1998; @Breiman2001; @Ishwaran2008; @Friedman2001; @Greenwell2022; @Friedman2002].
+Performance of the models was evaluated by comparison of the predicted and actual BCR-free survival in the pooled GEO training cohort, and TCGA and DKFZ validation collectives with Harrell's concordance index and Integrated Brier Scores (__Supplementary Tables S7__ - __S8__).
+Additionally, BCR-free survival in patients stratified by tertiles of predictor scores returned by the machine learning models was explored with Kaplan-Meier plots and Peto-Peto tests. 
+Finally, confidence and calibration of the models was assessed by plots of Brier Scores for unique follow-up times.
+
+## Multi-parameter modeling of overall survival
+
+Overall survival of PCA patients was modeled in the GSE16560 watchful waiting cohort [@Sboner2010]. 
+First- and second-order terms of ComBat-adjusted $log_2$-transformed expression levels of the collagen-related genes served as explanatory factors. 
+Machine learning models using the algorithms outlined above for the BCR-free survival were trained in a training subset of the GSE16560 cohort, and validated in the test subset. 
+Performance of the models of overall survival was evaluated in the training and test subsets of the GSE16560 cohort as described above for BCR-free survival (__Supplementary Tables S9__ - __S10__).
 
@@
 
-Publicly available data sets were analyzed. 
+Publicly available data sets were analyzed: the links are provided in __Methods/Transcriptome data sets__. 
 Formatted data sets used for analyses will be made available upon request to the corresponding author. 
@@
 _HR with 95% confidence intervals for BCR risk in the high versus low expression strata are presented in a Forest plot (left).
-The model performance statistics, Harrell's concordance index (C-index, measure of model accuracy), and Integrated Brier Scores (IBS, measure of overall model calibration, low values indicative of good calibration) are presented in a dot plot (right)._
+The model performance statistics, Harrell's concordance index (C-index, measure of model accuracy), and Integrated Brier Scores (IBS, measure of model calibration, low values indicative of good calibration) are presented in a dot plot (right)._
 
@@
 
-_(A) Metrics of performance of the survival models, Harrell's concordance index (C-index, measure of model accuracy) and Integrated Brier Score (IBS, measure of overall model calibration, low values are characteristic for good calibration), are presented in dot plots.
+_(A) Metrics of performance of the survival models, Harrell's concordance index (C-index, measure of model accuracy) and Integrated Brier Score (IBS, measure of model calibration, low values are characteristic for good calibration), are presented in dot plots.
 Dashed lines represent C-index and IBS values expected for random BCR risk prediction.

### 2025-02-25 paper/markdown/manuscript_supplement.Rmd
Methods for the main manuscript
ROC analysis for BCR is still pernding

@@
 
-The TCGA prostate cancer data set [@Liu2018; @Abeshouse2015] and DKFZ data set [@Gerhauser2018] consisted of normalized RNA sequencing data for `r tcga$expression %>% filter(tissue_type == 'tumor') %>% nrow` and `r dkfz$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples, respectively, with accompanying clinical information and were obtained from the cBioportal repository with in-house-developed R scripts.
+The [TCGA](https://www.cbioportal.org/study/summary?id=prad_tcga_pan_can_atlas_2018) prostate cancer data set [@Liu2018; @Abeshouse2015] and [DKFZ](https://www.cbioportal.org/study/summary?id=prostate_dkfz_2018) data set [@Gerhauser2018] consisted of normalized RNA sequencing data for `r tcga$expression %>% filter(tissue_type == 'tumor') %>% nrow` and `r dkfz$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples, respectively, with accompanying clinical information and were obtained from the cBioportal repository with in-house-developed R scripts.
 
-The GSE16560 [@Sboner2010] (microarray, Human 6k Transcriptionally Informative Gene Panel for DASL, n = `r gse16560$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples), 
-GSE54460 [@Long2014] (RNA sequencing, n = `r gse54460$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples), 
-GSE70768 [@Ross-Adams2015] (Illumina HumanHT-12 V4.0 expression beadchip microarray, n = `r gse70768$expression %>% filter(tissue_type == 'tumor') %>% nrow`), 
-GSE70769 [@Ross-Adams2015] (Illumina HumanHT-12 V4.0 expression beadchip microarray, n = `r gse70769$expression %>% filter(tissue_type == 'tumor') %>% nrow`), and 
-GSE220095 [@Schimmelpfennig2023] (RNA sequencing, n = `r gse220095$expression %>% filter(tissue_type == 'tumor') %>% nrow`) data sets with normalized whole-transcriptome and basic clinical information were fetched from GEO with the `getGEO()` function from the _GEOquery_ package.
+The [GSE16560](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE16560) [@Sboner2010] (microarray, Human 6k Transcriptionally Informative Gene Panel for DASL, n = `r gse16560$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples),
+GSE54460(https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE54460) [@Long2014] (RNA sequencing, n = `r gse54460$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples), 
+[GSE70768](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE70768) [@Ross-Adams2015] (Illumina HumanHT-12 V4.0 expression beadchip microarray, n = `r gse70768$expression %>% filter(tissue_type == 'tumor') %>% nrow`),
+[GSE70769](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE70769) [@Ross-Adams2015] (Illumina HumanHT-12 V4.0 expression beadchip microarray, n = `r gse70769$expression %>% filter(tissue_type == 'tumor') %>% nrow`), and
+[GSE220095](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE220095) [@Schimmelpfennig2023] (RNA sequencing, n = `r gse220095$expression %>% filter(tissue_type == 'tumor') %>% nrow`) data sets with normalized whole-transcriptome and basic clinical information were fetched from GEO with the `getGEO()` function from the _GEOquery_ package.
 
@@
 Harrell's concordance index [@Harrell1996] (high values are characteristic for high concordance between the predicted and observed BCR risk),  
-overall model calibration and confidence of predictions was gauged by Integrated Brier Score (low values are expected for well calibrated and highly confident models) [@Graf1999]. 
+model calibration and confidence of predictions was gauged by Integrated Brier Score (low values are expected for well calibrated and highly confident models) [@Graf1999]. 
 The inference and model fit metrics were computed by calling `summary()` method from [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) for the model objects. 
@@
   footnote(1, 10, 
-           value = as_paragraph('IBS: Integrated Brier Score, a metric of overall model calibration.
+           value = as_paragraph('IBS: Integrated Brier Score, a metric of model calibration.
 Low values are expected for a well calibrate model.
@@
   footnote(1, 5, 
-           value = as_paragraph('IBS: Integrated Brier Score, a metric of overall model calibration.
+           value = as_paragraph('IBS: Integrated Brier Score, a metric of model calibration.
 Low values are expected for a well calibrated model.
@@
   footnote(1, 4, 
-           value = as_paragraph('IBS: Integrated Brier Score, a metric of overall model calibration.
+           value = as_paragraph('IBS: Integrated Brier Score, a metric of model calibration.
 Low values are expected for a well calibrate model.
@@
 
-_(A) Numeric metrics of survival model performance, Harrell's concordance index (C-index, measure of model accuracy) and Integrated Brier Score (IBS, measure of overall calibration, low values are characteristic for well calibrated models), are presented in a dot plot.
+_(A) Numeric metrics of survival model performance, Harrell's concordance index (C-index, measure of model accuracy) and Integrated Brier Score (IBS, measure of calibration, low values are characteristic for well calibrated models), are presented in a dot plot.
 Each point represents a single GBM model, point color codes for the model type.
@@
 
-_(A) Metrics of performance of the survival models, Harrell's concordance index (C-index, measure of model accuracy) and Integrated Brier Score (IBS, measure of overall model calibration, low values are characteristic for good calibration), are presented in dot plots.
+_(A) Metrics of performance of the survival models, Harrell's concordance index (C-index, measure of model accuracy) and Integrated Brier Score (IBS, measure of model calibration, low values are characteristic for good calibration), are presented in dot plots.
 Dashed lines represent C-index and IBS values expected for random death risk prediction.

### 2025-02-25 paper scripts/links.R

@@ -29,6 +29,23 @@
     pmap(mdlink) %>%
     set_names(proj_links$obj_name)
   
+# links to the data sets --------
+  
+  data_links <- 
+    c(TCGA = 'https://www.cbioportal.org/study/summary?id=prad_tcga_pan_can_atlas_2018', 
+      DKFZ = 'https://www.cbioportal.org/study/summary?id=prostate_dkfz_2018', 
+      GSE16560 = 'https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE16560', 
+      GSE54460 = 'https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE54460', 
+      GSE70768 = 'https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE70768', 
+      GSE70769 = 'https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE70769', 
+      GSE220095 = 'https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE220095') %>% 
+    tibble(ref_name = names(.), 
+           x = .)
+  
+  data_links <- data_links %>%
+    pmap(mdlink) %>% 
+    set_names(data_links$ref_name)
+  
 # reading the bibliography -------
   
   insert_msg('Rending the bibliography')

### 2025-02-25 paper/markdown/manuscript_figures_tables.Rmd
ROC analyses for BCR
Analysis of prediction of VCR at one, two. three, and five year follow-up by ROC. Inference for the GBM models

@@
 Performance of the models was evaluated by comparison of the predicted and actual BCR-free survival in the pooled GEO training cohort, and TCGA and DKFZ validation collectives with Harrell's concordance index and Integrated Brier Scores (__Supplementary Tables S7__ - __S8__).
-Additionally, BCR-free survival in patients stratified by tertiles of predictor scores returned by the machine learning models was explored with Kaplan-Meier plots and Peto-Peto tests. 
-Finally, confidence and calibration of the models was assessed by plots of Brier Scores for unique follow-up times.
+BCR-free survival in patients stratified by tertiles of predictor scores returned by the machine learning models was explored with Kaplan-Meier plots and Peto-Peto tests. 
+Additionally, confidence and calibration of the models was assessed by plots of Brier Scores for unique follow-up times. 
+Finally, ability to predict BCR at one, two, three, and five years after diagnosis was assessed by receiver-operating characteristic (ROC) [@Heagerty2000].
 
@@
 Machine learning models using the algorithms outlined above for the BCR-free survival were trained in a training subset of the GSE16560 cohort, and validated in the test subset. 
-Performance of the models of overall survival was evaluated in the training and test subsets of the GSE16560 cohort as described above for BCR-free survival (__Supplementary Tables S9__ - __S10__).
+Performance of the models of overall survival was evaluated in the training and test subsets of the GSE16560 cohort as described above for BCR-free survival (__Supplementary Tables S10__ - __S11__).
 

### 2025-02-25 paper/markdown/manuscript_supplement.Rmd
ROC analyses for BCR
Analysis of prediction of VCR at one, two. three, and five year follow-up by ROC. Inference for the GBM models

@@
 _surviminer_ [@Kassambara2016],  
-[_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions), and 
-_caret_ [@Kuhn2008]. 
+[_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions), 
+_caret_ [@Kuhn2008], and 
+_survivalROC_ [@Heagerty2000; @Heagerty2022].
+
 Univariable analysis of relapse-free survival was accomplished with packages [_kmOptimizer_](https://github.com/PiotrTymoszuk/kmOptimizer'), _survival_, _survminer_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions).
@@
 With this graphical method we evaluated predictions made by the best-performing GBM models (__Supplementary Figure S\@ref(fig:fig-clinic)B__). 
-In addition, ability of the machine learning models to assign PCA patients to low-, intermediate-, and high-risk groups was investigated by comparing BCR-free survival between patients stratified by tertiles of the predictor scores with Kaplan-Meier plots and Peto-Peto test (functions `calibrate.coxex()`, `plot.calibrator()`, and `surv_pvalue()`, packages [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) and _survminer_) as presented for the GBM models in __Figure 2B__ and __Supplementary Figure \@ref(fig:fig-cl-plot)__.
+Ability of the machine learning models to assign PCA patients to low-, intermediate-, and high-risk groups was investigated by comparing BCR-free survival between patients stratified by tertiles of the predictor scores with Kaplan-Meier plots and Peto-Peto test (functions `calibrate.coxex()`, `plot.calibrator()`, and `surv_pvalue()`, packages [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) and _survminer_) as presented for the GBM models in __Figure 2B__ and __Supplementary Figure \@ref(fig:fig-cl-plot)__.
+Prediction of BCR at one, two, three, and five years after diagnosis by the GBM models was assessed by receiver-operating characteristic (ROC) with area under the curve (AUC) as a metric of prediction accuracy, sensitivity, and specificity.
+The ROC analysis was performed for predictor scores of the GBM models as marker variables with the nearest-neighbor algorithm proposed by Heagerty et al.
+[@Heagerty2000; @Heagerty2022] (function `survivalROC()`, package _survivalROC_).
+The ROC AUC values are outlined in __Supplementary Table S\@ref(tab:tab-roc)__.
 
@@
 
+```{r tab-roc, tab.cap = "Prediction of biochemical relapse (BCR) at one, two, three, and five year after diagnosis by gradient boosted models employing clinical predictors and expression of the collagen-related transcripts as explanatory factors."}
+
+flextable::flextable(suppl_tab$roc) %>% 
+  merge_v(1:2) %>% 
+  set_widths(c(2.5, 2.3, 4, rep(2.5, 3))) %>% 
+  footnote(1, 3, 
+            value = as_paragraph('GBM, expression, explanatory factors: expression levels of the collagen-related transcripts; GBM, clinic, explanatory factors: pathological tumor stage, ISUP grade, and log2 expression of the PSA coding gene KLK3 in the tumor tissue; GBM, clinic + expression: pathological tumor stage, ISUP grade, and log2 expression of the PSA coding gene KLK3 in the tumor tissue, log2 expression of the collagen-related transcripts.'),
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 6, 
+           value = as_paragraph('ROC AUC: area under the curve of receiver-operating characteristic.'), 
+           part = 'header', 
+           ref_symbols = 'b') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
 ```{r tab-os-tune, tab.cap = "Selection of the optimal parameters of machine learning models of overall survival (OS) in the training subset of the GSE16560 cohort.
@@
 
-__Figure \@ref(fig:fig-gbm-cohort). Investigation of confounding study effect on prediction of BCR-free survival by the GBM model in the pooled GEO cohort.__
+__Supplementary Figure S\@ref(fig:fig-gbm-cohort). Investigation of confounding study effect on prediction of BCR-free survival by the GBM model in the pooled GEO cohort.__
 
@@
 
-__Figure \@ref(fig:fig-clinic). Modeling of BCR-free survival by GBM algorithm with clinical predictors and mRNA expression of the collagen-related genes.__
+__Supplementary Figure S\@ref(fig:fig-clinic). Modeling of BCR-free survival by GBM algorithm with clinical predictors and mRNA expression of the collagen-related genes.__
 
@@
 
-```{r fig-cl-plot, fig.width = figur::convert(suppl_fig$cl_plot, to = 'in')$w, fig.height = figur::convert(suppl_fig$cl_plot, to = 'in')$h, fig.cap = 'GBM modeling of BCR-free survival in PCA with clinical prediction and expression of collagen-related transcripts: survival in tertiles of predictor scores.'}
+```{r fig-cl-plot, fig.width = figur::convert(suppl_fig$cl_plot, to = 'in')$w, fig.height = figur::convert(suppl_fig$cl_plot, to = 'in')$h, fig.cap = 'GBM modeling of BCR-free survival in PCA with clinical predictors and expression of collagen-related transcripts: survival in tertiles of predictor scores.'}
 
@@
 
-__Figure \@ref(fig:fig-cl-plot). GBM modeling of BCR-free survival in PCA with clinical prediction and expression of collagen-related transcripts: survival in tertiles of predictor scores.__
+__Supplementary Figure S\@ref(fig:fig-cl-plot).
+GBM modeling of BCR-free survival in PCA with clinical predictors and expression of collagen-related transcripts: survival in tertiles of predictor scores.__
 
-_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and expression of KLK3 transcript coding for PSA protein), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors, as presented in Supplementary Figure S\@ref(fig:fig-cl-plot)._
+_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and expression of KLK3 transcript coding for PSA protein), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors, as presented in Supplementary Figure S\@ref(fig:fig-clinic)._
+
 _BCR-free survival in tertiles of the predictor scores of the clinical-only and the combined clinical - expression GBM models was visualized with Kaplan-Meier plots.
@@
 Note improved resolution of the survival curves for the combined clinical/expression model as compared with the clinical-only model in the pooled GEO and DKFZ cohort._
+
+\newpage
+
+```{r fig-cl-iference, fig.width = figur::convert(suppl_fig$cl_inference, to = 'in')$w, fig.height = figur::convert(suppl_fig$cl_inference, to = 'in')$h, fig.cap = 'GBM modeling of BCR-free survival in PCA with clinical predictors and expression of collagen-related transcripts: inference.'}
+
+suppl_fig$cl_inference$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-cl-iference). GBM modeling of BCR-free survival in PCA with clinical predictors and expression of collagen-related transcripts: inference.__
+
+_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and expression of KLK3 transcript coding for PSA protein), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors, as presented in Supplementary Figure S\@ref(fig:fig-clinic)._
+
+_Predictor scores of the GBM models were computed for the training collective, and the TCGA and DKFZ validation cohorts.
+Subsequently, univariable Cox proportional hazard models of BCR-free survival were constructed with normalized values of the predictor scores (Z-scores) as explanatory factors.
+Inference statistics of those Cox proportional hazard models, hazard ratios (HR) with 95% confidence intervals (95%CI),m are presented in a Forest plot._
+
+\newpage
+
+```{r fig-roc, fig.width = figur::convert(suppl_fig$roc, to = 'in')$w, fig.height = figur::convert(suppl_fig$roc, to = 'in')$h, fig.cap = 'GBM modeling of BCR-free survival in PCA with clinical predictors and expression of collagen-related transcripts: prediction of BCR.'}
+
+suppl_fig$roc$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-roc). GBM modeling of BCR-free survival in PCA with clinical predictors and expression of collagen-related transcripts: prediction of BCR.__
+
+_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and expression of KLK3 transcript coding for PSA protein), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors, as presented in Supplementary Figure S\@ref(fig:fig-clinic)._
+
+_Ability of the GBM models to predict BCR at one, two, three, and five years after diagnosis in the pooled GEO training cohort, and the TCGA and DKFZ validation collectives was investigated by receiver-operating characteristic (ROC) with are under the curve (AUC) as a metric of accuracy, sensitivity, and specificity._
+
+_(A) Values of AUC or prediction of BCR at one, two, three, and five years after diagnosis presented in a dot plot.
+Each point represents a single time point.
+Values obtained with the same survival model are connected with lines._
+
+_(B) ROC curves for prediction of BCR at one, two, three, and five years after diagnosis (TR: true positive, FP: false positive).
+color of the curve represents the survival model.
+AUC values are indicated in the plots.
+Total numbers of patients in the modeling data and numbers of BCR cases at the time point are displayed in the plot captions._
 
@@
 
-__Figure \@ref(fig:fig-os). Modeling of overall survival in PCA with expression levels of the collagen-related transcripts as explanatory factors.__
+__Supplementary Figure S\@ref(fig:fig-os). Modeling of overall survival in PCA with expression levels of the collagen-related transcripts as explanatory factors.__
 

### 2025-02-25 BCR scripts/clinical_plots.R

@@ -12,6 +12,10 @@
   
   insert_msg('Analysis globals')
   
+  ## models of interest
+  
+  surv_clplots$models <- c("gbm_clinic", "gbm", "gbm_combi")
+  
   ## plotting labels and colors 
   
   surv_clplots$algo_labels <- 
@@ -21,23 +25,27 @@
   
   surv_clplots$algo_colors <- 
     c(reference = 'gray60', 
-      globals$algo_colors[c("gbm", "gbm_clinic", "gbm_combi")])
+      globals$algo_colors[surv_clplots$models])
+  
+  surv_clplots$algo_labeller <- surv_clplots$algo_labels %>% 
+    stri_replace(fixed = ', ', replacement = '\n') %>% 
+    set_names(names(surv_clplots$algo_labels))
   
   ## global performance stats 
   
   surv_clplots$stats <- 
-    surv_summary$stats[c("gbm_clinic", "gbm", "gbm_combi")] %>% 
+    surv_summary$stats[surv_clplots$models] %>% 
     compress(names_to = 'algorithm') %>% 
     mutate(cohort_lab = globals$study_labels[cohort], 
            cohort_lab = ifelse(algorithm == 'gbm_clinic', 
                                cohort_lab, NA), 
            algorithm = factor(algorithm, 
-                              c('gbm_clinic', 'gbm', 'gbm_combi')))
+                              surv_clplots$models))
   
   ## Brier scores for the unique survival time points
 
   surv_clplots$brier_scores <- 
-    surv_summary$brier_scores[c("gbm_clinic", "gbm", "gbm_combi")] %>% 
+    surv_summary$brier_scores[surv_clplots$models] %>% 
     map2(., 
          list(c('time', 'reference', 'training'), 
               c('time', 'training'), 
@@ -46,17 +54,52 @@
            map(select, all_of(y))) %>% 
     transpose %>% 
     map(reduce, full_join, by = c('time')) %>% 
-    map(set_names, c('time', 'reference', 'gbm_clinic', 'gbm', 'gbm_combi')) %>% 
+    map(set_names, c('time', 'reference', surv_clplots$models)) %>% 
     map(pivot_longer, 
-        cols = c(reference, gbm_clinic, gbm, gbm_combi), 
+        cols = all_of(c('reference', surv_clplots$models)), 
         names_to = 'algorithm', 
         values_to = 'brier_score') %>% 
     map(mutate, 
         algorithm = factor(algorithm, 
-                           c('reference', 
[…]

### 2025-02-25 BCR scripts/globals.R

@@ -35,8 +35,7 @@
     map(select, sample_id, relapse, rfs_months) %>% 
     map(mutate, 
         rfs_months = rfs_months + 1, 
-        scaled_rfs_months = minMax(rfs_months) + 0.01) %>% 
-    map(select, -rfs_months)
+        scaled_rfs_months = minMax(rfs_months) + 0.01)
   
   ## genes of interest: first and second order terms
   

### 2025-02-25 BCR scripts/roc.R

@@ -1,0 +1,87 @@
+# Prediction of BCR at 1, 2, 3, and 5 years after diagnosis by the machine 
+# learning models of RFS-free survival. 
+# The analysis based on predictor scores is possible only for the Cox-like 
+# models, i.e. not possible for the Random Forest model. 
+# ROC for survival time points is calculated with the NNE (nearest-neighbor 
+# estimate) method implemented in `survivalROC` package.
+
+  insert_head()
+  
+# container -------
+  
+  surv_roc <- list()
+  
+# parallel backend --------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# analysis data ------
+  
+  insert_msg('Analysis data')
+  
+  ## predictor scores of the GBM models, 
+  ## RFS-free survival at the identity scale, 
+  ## and 0/1-coded BCR event
+  
+  surv_roc$data <- surv_summary$tertile_data %>% 
+    map(map, 
+        select, 
+        sample_id, predictor_score)
+  
+  surv_roc$rfs_survival <- surv_globals$data %>% 
+    map(select, sample_id, rfs_months, relapse)
+  
+  for(i in names(surv_roc$data)) {
+    
+    surv_roc$data[[i]] <- 
+      map2(surv_roc$data[[i]], 
+           surv_roc$rfs_survival, 
+           inner_join, by = 'sample_id')
+    
+  }
+  
+  ## transposing: 
+  ## performance of different models in one cohort is interesting
+  
+  surv_roc$data <- transpose(surv_roc$data)
+  
+  ## time points of interest, months
+  
+  surv_roc$times <- c(12, 24, 36, 60)
+  
+  surv_roc$times <- 
+    set_names(surv_roc$times, 
+              paste0('time_', surv_roc$times))
+  
+# ROC analysis for survival time points ---------
+  
+  insert_msg('ROC analysis for survival time points')
+  
+  for(i in names(surv_roc$times)) {
+    
+    surv_roc$roc_obj[[i]] <- 
+      list(data_lst = surv_roc$data, 
+           span = surv_roc$data %>% 
+             map(~.x[[1]]) %>% 
+             map_dbl(~0.25 * nrow(.x)^(-0.2))) %>% 
+      pmap(surv_roc_times, 
+           time_variable = 'rfs_months', 
+           status_variable = 'relapse', 
+           marker_variable = 'predictor_score', 
[…]

### 2025-02-25 BCR scripts/summary.R

@@ -73,9 +73,9 @@
         training = test, 
         test = NA)
 
-# Score tertiles and tests for differences between the score tertiles -------
+# Normalized predictor scores, score tertiles and tests for differences between the score tertiles -------
   
-  insert_msg('Tertiles and tertile tests')
+  insert_msg('Normalized predictor scores, tertiles and tertile tests')
   
   ## this data is available only for the Cox-like models, SVM, and GBM
   ## but not for the Random Forests
@@ -107,8 +107,22 @@
         any_of(c('collagen_score', 
                  'svm_score', 
                  'clinic_score', 
-                 'gbm_score')))
+                 'gbm_score'))) %>% 
+    map(map, 
+        set_names, 
+        c('sample_id', 
+          'scaled_rfs_months', 
+          'relapse', 
+          'score_cuts', 
+          'predictor_score'))
   
+  ## normalized predictor scores (Z-scores)
+
+  surv_summary$tertile_data <- surv_summary$tertile_data %>% 
+    map(map, 
+        mutate, 
+        z_score = zScores(predictor_score))
+    
   ## tertile N numbers: total and events
   
   surv_summary$tertile_n <- surv_summary$tertile_data %>% 
@@ -145,18 +159,22 @@
   
   insert_msg('Comparison of C-indexes, GBM models')
   
-  ## with the one-show non-parameteric test proposed by Kang 2015
+  ## with the one-show non-parametric test proposed by Kang 2015
   ## and implemented by compareC package
   
   ## predictor scores 
   
   surv_summary$c_data <- 
     surv_summary$tertile_data[c("gbm_clinic", "gbm_combi")] %>% 
-    map(map, select, -score_cuts) %>% 
+    map(map, select, -score_cuts, -z_score) %>% 
     transpose %>% 
     map(reduce, 
         left_join, 
-        by = c('sample_id', 'scaled_rfs_months', 'relapse'))
+        by = c('sample_id', 'scaled_rfs_months', 'relapse')) %>% 
+    map(set_names, 
+        c('sample_id', 
+          'scaled_rfs_months', 'relapse', 
+          'clinic_score', 'gbm_score'))
   
   ## testing 
   
@@ -169,11 +187,47 @@
     map(mutate, 
         c_index = 1 - est.c)
   
+# Inference statistics for the normalized predictor scores --------
+  
[…]

### 2025-02-25 BCR.R

@@ -18,6 +18,7 @@
   library(survminer)
   library(coxExtensions)
   library(compareC)
+  library(survivalROC)
 
   library(furrr)
   library(soucer)
@@ -79,9 +80,12 @@
                          'with clinical predictors'))) %>% 
     pwalk(access_cache)
   
-  ## plots of the model evaluation 
+  ## summary statistics, 
+  ## ROC for selected, clinically relevant survival time points
+  ## plots of the model evaluation and comparison of model types
   
   c('./BCR scripts/summary.R', 
+    './/BCR scripts/roc.R', 
     './BCR scripts/plots.R', 
     './BCR scripts/clinical_plots.R', 
     './BCR scripts/cohort.R') %>% 

### 2025-02-25 paper scripts/supplementary_figures.R

@@ -179,11 +179,62 @@
     as_figure(label = 'bcr_survival_gbm_clinic_expression_km_plots', 
               ref_name = 'cl_plot', 
               caption = paste('GBM modeling of BCR-free survival in PCA with', 
-                              'clinical prediction and expression of', 
+                              'clinical predictors and expression of', 
                               'collagen-related transcripts:', 
                               'survival in tertiles of predictor scores.'), 
               w = 180, 
               h = 220)
+  
+# Inference, normalized HR, for the GBM models of BCR -------
+  
+  insert_msg('Inference of the GBM models')
+  
+  suppl_fig$cl_inference <- 
+    as_figure(surv_clplots$inference_plot + 
+                theme(legend.position = 'none'), 
+              label = 'bcr_survival_gbm_clinic_expression_inference', 
+              ref_name = 'cl_iference', 
+              caption =  paste('GBM modeling of BCR-free survival in PCA with', 
+                               'clinical predictors and expression of', 
+                               'collagen-related transcripts: inference.'), 
+              w = 100, 
+              h = 100)
+  
+# ROC, prediction of BCR at particular time points -------
+  
+  insert_msg('ROC for BCR prediction')
+  
+  ## upper panel: summary plot
+  
+  suppl_fig$roc$upper <- surv_clplots$auc_summary
+  
+  ## bottom panel: ROC curves for selected time points
+  
+  suppl_fig$roc$bottom <- 
+    surv_clplots$roc_plots[c("time_12", "time_36", "time_60")] %>% 
+    unlist(recursive = FALSE) %>% 
+    map(~.x + theme(legend.position = 'none')) %>% 
+    plot_grid(plotlist = ., 
+              ncol = 3, 
+              align = 'hv', 
+              axis = 'tblr')
+  
+  ## the entire figure
+  
+  suppl_fig$roc <- 
+    plot_grid(suppl_fig$roc$upper, 
+              suppl_fig$roc$bottom, 
+              nrow = 2, 
+              rel_heights = c(1, 3), 
+              labels = LETTERS, 
+              label_size = 10) %>% 
+    as_figure(label = 'bcr_survival_gbm_clinic_expression_roc', 
+              ref_name = 'roc', 
+              caption =  paste('GBM modeling of BCR-free survival in PCA with', 
+                               'clinical predictors and expression of', 
+                               'collagen-related transcripts: prediction of BCR.'), 
+              w = 190, 
+              h = 230)
   
 # Modeling of overall survival in the GSE16560 cohort -------
   

### 2025-02-25 paper scripts/supplementary_tables.R

@@ -274,6 +274,32 @@
                                'prediction of biochemical relapse (BCR) free', 
                                'survival.'))
   
+# AUC for prediction of BCR by the GBM models -----
+  
+  insert_msg('AUC for predicion of BCR at selected time points')
+  
+  suppl_tab$roc <- surv_roc$roc_obj %>% 
+    map(map, ~.x$stats) %>% 
+    map(compress, names_to = 'cohort') %>% 
+    reduce(rbind) %>% 
+    filter(stri_detect(algorithm, regex = '^gbm')) %>% 
+    transmute(`Time after diagnosis, months` = predict_time, 
+              Cohort = globals$study_labels[cohort], 
+              `Model` = surv_clplots$algo_labels[algorithm], 
+              `Total patients, N` = n_total, 
+              `Fraction of BCR` = signif(1 - survival_rate, 2), 
+              `ROC AUC` = signif(auc, 2))
+  
+  suppl_tab$roc <- suppl_tab$roc %>% 
+    as_mdtable(label = 'roc', 
+                ref_name = 'roc', 
+               caption = paste('Prediction of biochemical relapse (BCR) at', 
+                               'one, two, three, and five year after diagnosis', 
+                               'by gradient boosted models employing clinical', 
+                               'predictors and expression of the', 
+                               'collagen-related transcrips as', 
+                               'explanatory factors.'))
+  
 # tuning parameters for the models of overall survival --------
   
   insert_msg('Tuning parameters for the models of overall survival')

### 2025-02-25 tools/functions.R

@@ -15,6 +15,7 @@
   library(ggrepel)
   library(caret)
   library(figur)
+  library(survivalROC)
 
 # data import ------
 
@@ -414,6 +415,154 @@
       relocate(gene_symbol, cutoff)
 
   } 
+  
+# ROC analysis for survival time points ----------
+  
+  surv_roc_times <- function(data_lst,
+                             time_variable = 'rfs_months', 
+                             status_variable = 'relapse', 
+                             marker_variable = 'predictor_score', 
+                             predict_time = 12, ...) {
+    
+    ## ROC for survival at a given time point for 
+    ## a list of data frames with predictor scores
+    
+    data_lst <- data_lst %>% 
+      map(~filter(.x, complete.cases(.x)))
+    
+    roc_lst <- data_lst %>% 
+      future_map(function(x) survivalROC(Stime = x[[time_variable]], 
+                                         status = x[[status_variable]], 
+                                         marker = x[[marker_variable]], 
+                                         predict.time = predict_time, ...), 
+                 .options = furrr_options(seed = TRUE))
+    
+    ## extraction of AUC statistics, n_numbers and survival rates
+    
+    stats <- data_lst %>% 
+      map_dbl(nrow) %>% 
+      compress(names_to = 'algorithm', 
+               values_to = 'n_total')
+    
+    stats <- stats %>% 
+      mutate(predict_time = predict_time, 
+             survival_rate = map_dbl(roc_lst, ~.x$Survival), 
+             auc = map_dbl(roc_lst, ~.x$AUC))
+    
+    ## extraction of the cut points, true positives (TP), 
+    ## and false positives (FP)
+    
+    cut_data <- roc_lst %>% 
+      map(~.x[c('cut.values', 'TP', 'FP')]) %>% 
+      map(as_tibble) %>% 
+      compress(names_to = 'algorithm')
+    
+    ## output of the stats and ROC cut points, the later will be used for 
+    ## ROC plots
+    
+    list(stats = stats, 
+         cuts = cut_data)
+    
+  }
+  
+  plot_surv_roc_times <- function(roc_obj, 
+                                  plot_title_prefix = 'BCR', 
+                                  plot_title_suffix = NULL, 
+                                  palette = globals$algo_colors, 
+                                  labels = globals$algo_labels, 
[…]

### 2025-02-25 tools/globals.R

@@ -212,7 +212,7 @@
       svm = 'gray60', 
       rf = 'darkolivegreen', 
       gbm = 'orangered2', 
-      gbm_clinic = 'aquamarine3', 
+      gbm_clinic = 'aquamarine4', 
       gbm_combi = 'firebrick4')
   
   ## comparison of the collagen models with survival models

### 2025-02-27 exploration scripts/networks.R
Delete exploration scripts/networks.R
change of the sourcing sequence

@@ -1,193 +1,0 @@
-# co-expression networks of the collagen-related genes. 
-# The networks are built for correlations of ComBat-adjusted log2 expression 
-# levels with spearman's rho > 0.5. 
-#
-# Communities are identified by Leiden's algorithm.
-
-  insert_head()
-  
-# container -------
-  
-  net <- list()
-  
-# analysis globals: variables, their attributes and ComBat expression -------
-  
-  insert_msg('Analysis globals')
-  
-  ## variables and their attribute (functional group)
-  
-  net$variables <- globals$genes
-  
-  net$attr_tbl <- globals$genes_lexicon %>% 
-    set_names(c('variable', 'gene_group'))
-  
-  ## ComBat-adjusted log2 expression levels
-  
-  net$data <- globals$study_exprs %>% 
-    eval %>% 
-    map(~.x$combat) %>% 
-    map(column_to_rownames, 'sample_id') %>% 
-    map(select, all_of(net$variables))
-  
-  ## attribute lexicon as a list of tuples used later for plotting
-  
-  net$attr_lexicon <- 
-    list(community_id = c('community_id', 'gene communities'), 
-         gene_group = c('gene_group', 'gene function'), 
-         degree = c('degree', 'degree'), 
-         hub_score = c('hub_score', 'hub score'), 
-         betweennes = c('betweenness', 'betweenness')) 
-  
-# Graph objects -------
-  
-  insert_msg('Graph objects')
-  
-  ## isolated nodes are removed
-  ## setting the attributes
-  
-  net$graph_obj <- net$data %>% 
-    map(as_iGraph, 
-        feature_type = 'columns', 
-        fun = cor, 
-        cutoff = 0.5, 
-        weighted = TRUE, 
-        diag = FALSE) %>% 
-    map(prune_degree, cutoff = 0) %>% 
-    map(set_vertex_attributes, 
-        net$attr_tbl)
-  
-# Communities ---------
-  
-  insert_msg('Communities')
-  
-  set.seed(12345)
-  
-  ## communities with single genes (if present are lumped)
-  
-  net$communities <- net$graph_obj %>% 
-    map(cluster_leiden, 
-        objective_function = 'modularity', 
-        resolution_parameter = 1,  
-        n_iterations = 100) %>% 
-    map(comm_lump, 
-        cutoff = 1, 
-        other_name = 'other')
-  
-  ## appending the graph nodes with the community 
-  ## assignment information 
-  
[…]

### 2025-02-27 paper/markdown/manuscript_supplement.Rmd
BCR-risk association in network plots
 BCR risk marker status (univariable, shared by three cohorts) indicated in co-expression network plots)

@@
 _The co-expression networks were visualized with Kamada-Kawai algorithm.
-Each point represents a single transcript in the network; point color codes for functional classification of the transcript.
+Each point represents a single transcript in the network.
+Point color codes for functional classification of the transcript and BCR risk association status shared by all three cohorts in unavailable survival analysis.
 Edges, i.e.
-correlations of expression levels with $\rho \geq$ 0.5, are depicted as gray lines.
-Numbers of transcripts (vertices) and correlations with $\rho \geq$ 0.5 (edges) are displayed in the plot captions._
+correlations of expression levels with $\rho \geq$ 0.5, are depicted as gray lines._
+_Network dimensions, pooled GEO: `r dimensions(net$graph_obj$geo_pool)['vertices']` vertices and `r dimensions(net$graph_obj$geo_pool)['edges']` edges; TCGA:  `r dimensions(net$graph_obj$tcga)['vertices']` vertices and `r dimensions(net$graph_obj$tcga)['edges']` edges; DKFZ: `r dimensions(net$graph_obj$dkfz)['vertices']` vertices and `r dimensions(net$graph_obj$dkfz)['edges']` edges._
 

### 2025-02-27 exec.R

@@ -15,6 +15,10 @@
       'BCR.R', 
       'OS.R', 
       
+      ## co-expression network analysis
+      
+      'networks.R', 
+      
       ## manuscript parts
       'paper.R'
       

### 2025-02-27 exploration.R

@@ -56,12 +56,11 @@
   insert_msg('Exploration scripts')
   
   ## general distribution stats
-  ## co-expression analysis and co-expression networks
+  ## co-expression analysis
   
   c('./exploration scripts/cohorts.R', 
     './exploration scripts/distribution.R', 
-    './exploration scripts/coexpression.R', 
-    './exploration scripts/networks.R') %>% 
+    './exploration scripts/coexpression.R') %>% 
     source_all(message = TRUE, crash = TRUE)
   
   ## characteristic of the pooled GEO cohort; 

### 2025-02-27 network scripts/networks.R

@@ -1,0 +1,232 @@
+# Co-expression networks of the collagen-related genes. 
+# The networks are built for correlations of ComBat-adjusted log2 expression 
+# levels with spearman's rho > 0.5. 
+#
+# Communities are identified by Leiden's algorithm.
+
+  insert_head()
+  
+# container -------
+  
+  net <- list()
+  
+# analysis globals: variables, their attributes and ComBat expression -------
+  
+  insert_msg('Analysis globals')
+  
+  ## variables and their attributes: 
+  ## functional group and association with BCR risk
+  
+  net$variables <- globals$genes
+  
+  net$attr_tbl$gene_group <- globals$genes_lexicon %>% 
+    set_names(c('variable', 'gene_group'))
+  
+  net$attr_tbl$bcr <- uni_cut$common_significant %>% 
+    map(~tibble(variable = .x)) %>% 
+    compress(names_to = 'bcr_risk')
+  
+  net$attr_tbl <- net$attr_tbl %>% 
+    reduce(left_join, by = 'variable') %>% 
+    mutate(bcr_risk = ifelse(is.na(bcr_risk), 'ns', bcr_risk), 
+           bcr_risk = factor(bcr_risk, c('unfavorable', 'favorable', 'ns')))
+  
+  ## ComBat-adjusted log2 expression levels
+  
+  net$data <- globals$study_exprs %>% 
+    eval %>% 
+    map(~.x$combat) %>% 
+    map(column_to_rownames, 'sample_id') %>% 
+    map(select, all_of(net$variables))
+  
+  ## attribute lexicon as a list of tuples used later for plotting
+  
+  net$attr_lexicon <- 
+    list(community_id = c('community_id', 'gene communities'), 
+         gene_group = c('gene_group', 'gene function'), 
+         degree = c('degree', 'degree'), 
+         hub_score = c('hub_score', 'hub score'), 
+         betweennes = c('betweenness', 'betweenness'), 
+         bcr_risk = c('bcr_risk', 'BCR marker')) 
+  
+# Graph objects -------
+  
+  insert_msg('Graph objects')
+  
+  ## isolated nodes are removed
+  ## setting the attributes
+  
+  net$graph_obj <- net$data %>% 
+    map(as_iGraph, 
+        feature_type = 'columns', 
+        fun = cor, 
+        cutoff = 0.5, 
+        weighted = TRUE, 
+        diag = FALSE) %>% 
[…]

### 2025-02-27 networks.R

@@ -1,0 +1,40 @@
+# Co-expression networks analyses for the collagen-related genes
+
+# tools --------
+
+  library(tidyverse)
+  library(rlang)
+  library(trafo)
+  library(stringi)
+
+  library(igraph)
+  library(graphExtra)
+  
+  library(ggrepel)
+  library(ggtext)
+  library(RColorBrewer)
+  
+  library(furrr)
+  library(soucer)
+  
+  insert_head()
+  
+  explore <- exda::explore
+  select <- dplyr::select
+  reduce <- purrr::reduce
+  set_rownames <- trafo::set_rownames
+  
+  c('./tools/globals.R', 
+    './tools/functions.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+  
+# analysis scripts --------
+  
+  insert_msg('Analysis scripts')
+  
+  c('./network scripts/networks.R') %>% 
+    source_all(message = TRUE)
+
+# END ------
+  
+  insert_tail()

### 2025-02-27 paper scripts/supplementary_figures.R

@@ -10,23 +10,38 @@
   
   insert_msg('Co-expression of the collagen-related genes')
   
+  ## functional classification and association with 
+  ## BCR risk in univariable survival analysis
+  
   suppl_fig$nets <- 
-    map2(net$graph_plots$gene_group[c("geo_pool", "tcga", "dkfz")], 
-         c("geo_pool", "tcga", "dkfz"), 
-         ~.x + 
-           labs(title = globals$study_labels[.y]) + 
-           theme(legend.position = 'none')) %>% 
-    c(list(get_legend(net$graph_plots$gene_group[[1]]))) %>% 
-    plot_grid(plotlist = ., 
+    list(gene_group = net$graph_plots$gene_group[c("geo_pool", "tcga", "dkfz")], 
+         bcr_risk = net$graph_plots$bcr_risk[c("geo_pool", "tcga", "dkfz")]) %>% 
+    transpose %>% 
+    unlist(recursive = FALSE) %>% 
+    map(~.x + 
+          theme(legend.position = 'none', 
+                plot.subtitle = element_blank())) %>%
+    plot_grid(plotlist = .,
               ncol = 2, 
               align = 'hv', 
-              axis = 'tblr') %>% 
+              axis = 'tblr',
+              labels = c('A', '', 
+                         'B', '', 
+                         'C', ''), 
+              label_size = 10) %>% 
+    plot_grid(plot_grid(get_legend(net$graph_plots$gene_group$geo_pool + 
+                                     theme(legend.position = 'bottom')), 
+                        get_legend(net$graph_plots$bcr_risk$geo_pool + 
+                                     theme(legend.position = 'bottom')), 
+                        nrow = 2), 
+              nrow = 2, 
+              rel_heights = c(0.9, 0.1)) %>% 
     as_figure(label = 'co_expression_networks_collagen_related_genes', 
               ref_name = 'nets', 
               caption = paste('Co-expression networks of the collagen-related', 
                               'transcripts in PCA tissue.'), 
               w = 180, 
-              h = 180)
+              h = 230)
   
 # Co-expression networks, vertex importance ------
   
